---
title: "Recovering a Latent Outcome Scale"
subtitle: "Focused binary and ordinal comparisons of double machine learning and SUDO"
author: "Brandon M. Greenwell"
format:
  pdf:
    documentclass: article
    fontsize: 10pt
    geometry:
      - margin=0.68in
    colorlinks: true
    keep-tex: false
execute:
  echo: false
  warning: false
  message: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{microtype}
  - \usepackage{enumitem}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{0.35em}
  - \setlist{nosep,leftmargin=1.3em}
  - \AtBeginDocument{\pagenumbering{gobble}}
---

## The binary problem

With a binary outcome, we observe only 0 or 1 even when the scientific target
lives on an underlying continuous scale. Suppose

$$
P(Y=1\mid D,X)=F\{\theta_0D+g_0(X)\},
$$

Here $D$ is the treatment, $X$ contains background variables that may affect
both treatment and outcome, and $F$ turns the underlying index into a
probability. The target $\theta_0$ is the treatment effect on that underlying
index. Ordinary double machine learning applied to the $0/1$ label answers a
different question: how the treatment changes the probability-scale outcome.
A logistic version of double machine learning can recover $\theta_0$, but only
when the logistic link is correct. Flexible nuisance models do not repair a
wrong link.

Surrogate-Assisted Double Machine Learning (SUDO) first fits the full outcome
distribution. It then replaces each binary label with a continuous surrogate
that is consistent with both the fitted model and the observed label. When
that outcome model is correct,

$$
E(S\mid D,X)=\theta_0D+g_0(X).
$$

The usual treatment-effect calculation can then be applied to $S$. The same
downstream calculation works with a logit, complementary log-log, or another
supported link. SUDO is flexible about the link, but it is not automatically
robust: it can still fail if the fitted outcome distribution is wrong.

## A deliberately simple binary experiment

The simulation changes only the response link. For each replication,

$$
X\sim N(0,1),\qquad
D=0.5X+\nu,\quad \nu\sim N(0,0.7^2),
$$

$$
\eta=a_F+0.8D+0.5X,\qquad Y\sim\operatorname{Bernoulli}\{F(\eta)\}.
$$

The true latent-scale treatment effect is $\theta_0=0.8$. The first setting
uses a logit link. The second uses a complementary log-log link, with its
intercept chosen so the baseline outcome probability is one half. The
treatment is continuous, the confounding pattern is linear, and each sample
has 2,000 observations. The design is intentionally simple so the response
link is the only important difference.

Five estimators are fit on the same 300 simulated samples under each link:

1. **Raw-label DML with partially linear regression (PLR):** treats 0 and 1 as
   ordinary numbers. It estimates a treatment effect on that numeric scale,
   not the latent scale.
2. **Partially logistic DML:** uses a logistic model designed to estimate the
   latent coefficient when the true link is logit.
3. **Logit SUDO:** builds the continuous surrogate using a logit outcome model.
4. **Cloglog SUDO:** builds the surrogate using a complementary log-log
   outcome model.
5. **Correct-link generalized linear model (GLM):** fits the correct simple
   response model. It serves as a benchmark.

All DML procedures use five folds. SUDO averages 25 valid surrogate draws for
each data set. The supporting prediction models are intentionally smooth and
well matched to the simple design.

\newpage

## Terms used in the binary comparison

- **SUDO** means *Surrogate-Assisted Double Machine Learning*. It first
  restores a continuous outcome scale and then estimates the treatment effect.
- **DML** means *double machine learning*. It uses machine learning to predict
  the parts of treatment and outcome explained by $X$, then relates what is
  left of the outcome to what is left of the treatment.
- **PLR** means *partially linear regression*. The treatment enters with one
  coefficient, while the effect of $X$ may be learned flexibly.
- **GLM** means *generalized linear model*. Here it is the familiar binary
  regression model fit with the link that generated the data.
- **Logit** and **cloglog** are two different rules for translating a latent
  index into an outcome probability. Cloglog abbreviates complementary
  log-log.

## Binary results

Bias is the average estimate minus the true latent coefficient, 0.8. Coverage
is the fraction of 95% confidence intervals that contain 0.8.

```{r}
bias <- read.csv("../../python/results/binary_link_bias.csv",
                 check.names = FALSE)
method_order <- c(
  "Raw-label DML PLR", "Partially logistic DML", "SUDO (logit)",
  "SUDO (cloglog)", "Correct-link GLM"
)
make_display <- function(data, value) {
  logit <- data[data$truth == "logit", ]
  cloglog <- data[data$truth == "cloglog", ]
  data.frame(
    Estimator = method_order,
    `Logit truth` = logit[[value]][match(method_order, logit$estimator)],
    `Cloglog truth` = cloglog[[value]][match(method_order, cloglog$estimator)],
    check.names = FALSE
  )
}
knitr::kable(
  make_display(bias, "bias_vs_latent_theta"),
  caption = "Bias relative to the latent coefficient.",
  digits = 3, booktabs = TRUE, align = c("l", "r", "r")
)
```

```{r}
coverage <- read.csv("../../python/results/binary_link_coverage.csv",
                     check.names = FALSE)
knitr::kable(
  make_display(coverage, "coverage_of_latent_theta"),
  caption = "Coverage of nominal 95% intervals for the latent coefficient.",
  digits = 3, booktabs = TRUE, align = c("l", "r", "r")
)
```

Under logit truth, partially logistic DML and logit SUDO are essentially
unbiased, with coverage 0.953 and 0.980. Under cloglog truth, partially
logistic DML has bias 0.295 and coverage 0.217. Cloglog SUDO has bias 0.004
and coverage 0.953, nearly matching the correctly specified GLM. SUDO with
the wrong link performs poorly in both directions, as expected.

Raw-label PLR has zero coverage of $\theta_0$ under both links because it is
estimating a different quantity. This is not a defective PLR confidence
interval. It is the consequence of treating a binary label as though its
numeric scale were the target scale.

## What the binary experiment shows

The choice of link changes the target that can be recovered. Partially
logistic DML works well when the logistic model is right. SUDO keeps the same
final treatment-effect calculation while allowing the outcome model to use a
different supported link.

This experiment establishes the link-flexibility principle under a simple,
well-fitted full outcome model. It does not establish validity for an arbitrary
black-box distribution learner. Such learners still require direct
theta-level bias checks and full-pipeline coverage validation.

\newpage

## The SUDO algorithm

The binary and ordinal examples use the same core procedure. They differ only
in the chosen outcome model and in how the surrogate response is generated.

Steps 3 through 5 use the latent surrogate response proposed in the surrogate
residual literature and implemented by `sure::surrogate(method = "latent")`.
The `sure` package subtracts the full fitted mean from this response to create
a diagnostic residual. SUDO instead keeps the response as its outcome and,
in Step 6, removes only the part explained by $X$. This leaves the treatment
component available for estimation. The Rao-Blackwell surrogate response is
the conditional average of the same random surrogate that `sure` generates.

Stage 20 verifies exact seeded agreement for binary logit and probit and for
ordinal logit, probit, and cloglog, after accounting for `sure`'s harmless
ordinal threshold-origin shift. One implementation exception is binary
cloglog: local `sure` 0.3.0.9000 uses the ordinal Gumbel-min convention, while
a binomial cloglog model requires Gumbel-max. SUDO uses the required binary
convention; the discrepancy is recorded in
[`sure` issue 45](https://github.com/bgreenwell/sure/issues/45).

1. **Choose and state the latent scale.** Select a response link and latent
   error distribution that give the treatment coefficient a scientific
   meaning. The binary examples use logit or cloglog. The ordinal example uses
   a proportional-odds logit model. This choice is an assumption, not something
   DML can discover from the category labels alone.

2. **Split the observations into folds.** Each observation is held out once.
   All predictions for that observation must come from models trained on other
   folds. This is cross-fitting.

3. **Fit the model that will generate the surrogate.** Using the training
   folds, model the probability of every possible $Y$ from both treatment $D$
   and background variables $X$. For a partially linear treatment effect, use
   a latent index of the form $\alpha D+f(X)$: $f(X)$ may be learned flexibly,
   but treatment enters through one coefficient. Predict the index, category
   probabilities, and any ordinal cutpoints for the held-out fold. No
   surrogate response has been created yet.

4. **Find the latent-variable interval.** The observed category tells us where
   the continuous latent variable $Z_i$ can lie. Call this interval $A_i$. A
   binary label places $Z_i$ above or below one boundary. An ordinal label
   places $Z_i$ between two adjacent cutpoints. This step identifies the
   allowed range but still does not create the surrogate.

5. **Generate the surrogate response $S$.** For every held-out observation,
   combine $A_i$ with the fitted outcome distribution. This is the step where
   the new continuous response is actually created. There are two routes:

   - **Rao-Blackwell surrogate response:** set
     $S_i=\widehat E(Z_i\mid Y_i,D_i,X_i)$. This conditional-mean version is
     used in the ordinal example.
   - **Randomized surrogate response:** draw
     $S_i\sim\widehat{\mathcal L}(Z_i\mid Y_i,D_i,X_i)$. For proper multiple
     imputation, first redraw the fitted outcome-model parameters, including
     ordinal cutpoints. Repeat this step to create several surrogate responses.
     This is the route used in the binary example.

   At the end of Step 5, the collection $(S_1,\ldots,S_n)$ is the surrogate
   response used in all remaining steps.

6. **Apply DML to the surrogate response.** On the training folds, separately
   learn the expected treatment $E(D\mid X)$ and expected surrogate response
   $E(S\mid X)$. Predict both for the held-out fold and form residuals
   $\widetilde D=D-\widehat E(D\mid X)$ and
   $\widetilde S=S-\widehat E(S\mid X)$. For ordinal multiple imputation, the
   surrogate-response regression is refit for every surrogate data set.

7. **Estimate the treatment effect.** After every observation has held-out
   residuals, compute

   $$
   \widehat\theta=
   \frac{\sum_i \widetilde D_i\widetilde S_i}
        {\sum_i \widetilde D_i^2}.
   $$

   In plain language, this compares surrogate responses between people whose
   treatments differ more than their background variables would predict.

8. **Combine repeated surrogate responses when needed.** With multiple imputation,
   repeat Steps 5 through 7 and average the treatment estimates. Rubin's rules
   combine ordinary sampling uncertainty with the variation across surrogate
   data sets. The Rao-Blackwell route produces one surrogate response and does
   not need this pooling step.

9. **Account for fitting uncertainty.** For a flexible learned outcome model,
   resample the original observations and rerun Steps 2 through 8 from the
   beginning. The standard deviation of these full-pipeline bootstrap estimates
   is the standard error. The ordinal example uses 99 such bootstrap samples.

10. **Validate the complete procedure.** Before using a new outcome learner,
    simulate new data with a known $\theta$, tune on separate simulation seeds,
    and check final bias, interval coverage, numerical failures, treatment
    overlap, valid probabilities, and ordered cutpoints. Cross-fitting protects
    against ordinary prediction error; it cannot repair a wrong outcome link or
    a poor full outcome model.

\newpage

## Ordinal extension: category labels are not a scale

For an ordinal outcome, we know the category order but not the distance
between categories. Calling five categories 1 through 5 does not establish
that the distance from 1 to 2 equals the distance from 4 to 5. The simulation
uses the following exact data-generating process:

$$
\begin{aligned}
P(D=1\mid X)&=\operatorname{expit}(0.45X_1-0.35X_4),\\
Z&=D+g_0(X)+\varepsilon,\qquad \varepsilon\sim\operatorname{Logistic}(0,1),\\
g_0(X)&=0.65X_1+0.55\sin(X_2)+0.45(X_3^2-1)-0.40X_4.
\end{aligned}
$$

- **$X$:** 50 background variables. Only the first four affect treatment or
  outcome; the remaining 46 are noise variables.
- **$D$:** a binary treatment whose probability depends on $X_1$ and $X_4$.
- **$Z$:** the continuous latent variable. Treatment adds exactly 1 to
  $Z$, so the target is $\theta_0=1$.
- **$Y$:** the observed five-category response, obtained by cutting $Z$ at
  $-1.5$, $-0.3$, $0.8$, and $1.8$.

Each simulated sample has 3,000 observations. The outcome depends on linear,
sine, and squared terms, so the learner must find a nonlinear pattern among
many irrelevant variables.

### A flexible, structured learner

The full outcome distribution is fit with `mboost`, which builds a flexible
model by repeatedly adding small spline-based improvements. We use its
proportional-odds model: one treatment effect shifts the latent variable in the
same way across all four category boundaries. Its predictor is

$$
\alpha D+\sum_{k=1}^{50}f_k(X_k),
$$

The functions $f_k$ are smooth curves learned from the data. Five-fold
cross-fitting means that every observation is evaluated by a model trained
without that observation's fold. A separate 20-sample simulation selected
2,000 boosting steps from $\{500,1000,2000,5000,10000\}$ by comparing bias in
the final treatment estimate. The later results use new simulation samples.

Suppose a person is observed in category 3. The fitted cutpoints tell us that
this person's latent variable $Z$ lies between the second and third cutpoints.
The model uses the person's treatment and background variables to describe how
$Z$ is distributed inside that interval. SUDO uses the model's average value
of $Z$ inside the interval as the surrogate response $S$. Two people in
category 3 can
therefore receive different surrogate values because their fitted
distributions differ. Using the exact conditional average rather than a
random draw gives the **Rao-Blackwell surrogate response**.

SUDO then uses DML to compare the surrogate responses between treated and
untreated people after adjusting for $X$. Each standard error comes from 99
bootstrap samples; the complete fitting and cross-fitting process is rerun in
every bootstrap sample.

### Ordinal estimators in plain language

- **SUDO, Rao-Blackwell surrogate:** uses the model-implied conditional mean of
  the latent variable given the observed category as $S$, then uses $S$ as the
  outcome in DML.
- **Direct mboost index contrast:** reads the treatment effect directly from
  the boosted ordinal model. It is a diagnostic for shrinkage in that model.
- **Oracle latent-variable PLR:** analyzes the unobserved $Z$ using the true
  treatment model. This impossible-in-practice estimator is a best-case check.
- **Score-coded DML:** assigns numbers to the categories and applies DML to
  those numbers.
- **Oracle score-coded DML:** gives score-coded DML the true nuisance
  functions. If it still misses 1, machine-learning error is not the cause.

\newpage

### Ordinal results

The tables report 50 new simulation samples. Their columns have the following
meanings:

- **MC SE** is the Monte Carlo standard error. It measures simulation noise in
  the reported bias, not uncertainty within one data set.
- **SD** is the standard deviation of the estimates across the 50 samples.
- **Mean bootstrap SE** is the average standard error estimated within a
  sample. **Coverage** is the fraction of 95% intervals containing the truth.

Coverage is shown only for SUDO, for which full-pipeline intervals were built.

```{r}
ordinal <- read.csv("../../R/results/stage19_mboost_ordinal.csv",
                    check.names = FALSE)
ordinal_order <- c("rao_blackwell", "direct_ordinal", "oracle_latent_plr")
ordinal_labels <- c(
  rao_blackwell = "SUDO, Rao-Blackwell surrogate",
  direct_ordinal = "Direct mboost index contrast",
  oracle_latent_plr = "Oracle latent-variable PLR"
)
latent_results <- ordinal[match(ordinal_order, ordinal$estimator), ]
latent_results$Estimator <- unname(ordinal_labels[latent_results$estimator])
latent_results$Estimate <- 1 + latent_results$bias
latent_results <- latent_results[, c(
  "Estimator", "Estimate", "bias", "mc_se_bias", "sd", "mean_se", "coverage"
)]
row.names(latent_results) <- NULL
names(latent_results) <- c(
  "Estimator", "Mean", "Bias", "MC SE", "SD", "Mean boot. SE", "Coverage"
)
old_na <- getOption("knitr.kable.NA")
options(knitr.kable.NA = "")
knitr::kable(
  latent_results,
  caption = "Latent-scale estimates; the true coefficient is 1.",
  digits = 3, booktabs = TRUE,
  align = c("l", "r", "r", "r", "r", "r", "r")
)
options(knitr.kable.NA = old_na)
```

SUDO's average estimate is 0.995, almost exactly the true value of 1. Its bias
is $-0.005$, compared with a Monte Carlo SE of 0.011. Coverage is 0.920. With
only 50 samples, coverage itself has Monte Carlo SE 0.038, so 0.920 is
statistically compatible with 0.950. The direct mboost estimate is low by
0.061 because boosting shrinks the fitted treatment contrast.

Ordinary DML was also applied to two different numeric encodings of the same
ordered categories: equally spaced $(0,2,4,6,8)$ and unevenly spaced
$(0,0.5,4,7.5,8)$. Oracle versions use the true propensity and true
conditional mean of the encoded outcome.

```{r}
score_rows <- c(
  "dml_equal", "dml_middle", "oracle_dml_equal", "oracle_dml_middle"
)
score_labels <- c(
  dml_equal = "Fitted", dml_middle = "Fitted",
  oracle_dml_equal = "Oracle", oracle_dml_middle = "Oracle"
)
coding_labels <- c(
  dml_equal = "Equally spaced", dml_middle = "Unevenly spaced",
  oracle_dml_equal = "Equally spaced", oracle_dml_middle = "Unevenly spaced"
)
score_results <- ordinal[match(score_rows, ordinal$estimator), ]
score_table <- data.frame(
  Nuisances = unname(score_labels[score_results$estimator]),
  `Category scores` = unname(coding_labels[score_results$estimator]),
  Mean = 1 + score_results$bias,
  Bias = score_results$bias,
  `MC SE` = score_results$mc_se_bias,
  check.names = FALSE
)
knitr::kable(
  score_table,
  caption = "DML applied directly to numeric category scores.",
  digits = 3, booktabs = TRUE,
  align = c("l", "l", "r", "r", "r")
)
```

Changing only the category numbers shifts the ordinary DML estimate by 0.174
(Monte Carlo SE 0.004). Even with the true nuisance functions, the shift is
0.170 (Monte Carlo SE 0.005). Better prediction and more cross-fitting cannot
recover the latent coefficient from arbitrary category scores. These DML
estimators are not broken. They accurately answer a question defined by the
chosen numbers, but that question is not the latent treatment effect.

The ordinal experiment extends the binary lesson: SUDO can use a flexible,
high-dimensional outcome model and still estimate a clearly defined latent
effect with useful uncertainty estimates. The proportional-odds assumption
and the common treatment effect remain important. This result does not show
that SUDO survives an arbitrary black box or a wrong ordinal link.

For transparency, the original simulation script also demanded that the
direct boosted contrast have bias below 0.03. Its bias was $-0.061$, so the
script ended with a failed overall assertion. The SUDO bias, coverage,
standard-error, numerical, oracle, and category-coding checks all passed.
