---
title: "Surrogate-Assisted Double Machine Learning: A Uniform Interface for Discrete Outcomes"
author:
  - name: Brandon M. Greenwell
    email: greenwbm@ucmail.uc.edu
    affiliations:
      - name: "University of Cincinnati Carl H. Lindner College of Business"
        department: Department of Operations, Business Analytics, and Information Systems
        address: 2906 Woodside Drive
        city: Cincinnati, Ohio
        country: United States
        postal-code: 45221
    orcid: 0000-0002-8120-0084
  - name: Other Authors
    email: other.authors@email.com
    affiliations:
      - name: "University of XYZ"
authorcols: true
date: today
abstract: |
  Regression on a discrete label generally targets the wrong scale for a
  cumulative-link or log-mean effect. Surrogate-Assisted Double Machine
  Learning (SUDO) first completes each observed category to a continuous
  surrogate using a fitted conditional CDF, then applies ordinary
  partialling-out. If the fitted distribution is correct, randomized
  probability-integral-transform completion restores the scalar treatment
  index in conditional mean. Repeated completions with proper
  multiple-imputation draws provide inference for parametric models; a
  full-pipeline bootstrap serves the same role for flexible learners.
  Simulations show nominal coverage in the validated binary and ordinal
  designs and also expose the method's main limitation: error in the
  imputation distribution is not removed by DML orthogonality. SUDO matches,
  rather than dominates, tailored binary estimators. Its clearest use is for
  ordinal outcomes, where the same construction targets a latent
  proportional-odds coefficient without a new outcome-specific score. A wine
  quality application illustrates the method. An oracle experiment validates
  the count completion, while fitted count learners remain unpromoted.
format:
  arxiv-pdf:
    keep-tex: false
bibliography: references.bib
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}
---

# Introduction

Many outcomes are discrete: relapse is binary, survey responses are ordinal,
and trips are counts. Double Machine Learning [DML, @chernozhukov2018] is
well suited to a continuous partially linear outcome because it removes
flexibly estimated covariate effects and reads the treatment effect from a
residual-on-residual regression. Applying the same calculation to a discrete
label is easy, but it usually targets a residualized contrast in
$E[Y\mid D,X]$. It does not recover a cumulative-link coefficient or a log
conditional-mean effect. The problem is the outcome scale.

This paper takes a simple route around that mismatch: complete the observed
category to a continuous surrogate on the target scale, then run the usual
partialling-out calculation. The completion uses the fitted conditional CDF
to draw within the jump corresponding to the observed category. Under a
correct distribution, that randomized probability-integral transform has a
known continuous reference law. Adding the fitted scalar index produces a
surrogate whose conditional mean is the desired partially linear index.

The construction repurposes the surrogate-variable approach of @liu2018,
originally developed for residual diagnostics in ordinal regression, as an
estimation device. @cheng2021 extend that approach to unordered
discrete-choice models with vector utilities, while @greenwell2018 provide
its binary and ordinal implementation in the `sure` package. SUDO stays in
the scalar-index setting: a distribution learner supplies a pointwise index
and conditional CDF, and SUDO supplies the completion, partialling-out, and
uncertainty calculation. Binary, ordinal, and count laws then share one
estimator interface.

That uniformity does not make SUDO uniformly better. Tailored orthogonal
scores exist for logistic binary models [@liu2021lplr] and Poisson log-mean
effects [@drukker2022]. In our binary comparisons, SUDO matches the correct
tailored estimator and direct coefficient rather than improving on them. The
main practical opening is ordinal regression, where we identify no published
orthogonal-score analogue for the latent proportional-odds coefficient.

The paper makes three points. First, randomized-CDF completion turns a broad
class of discrete outcomes into an ordinary partially linear problem.
Second, inference must propagate completion and imputation-model uncertainty:
proper multiple-imputation draws do so for parametric models, while a
full-pipeline bootstrap does so for validated flexible learners. Third, DML
orthogonality protects the adjustment regressions but not the distribution
used to create the surrogate. Simulation therefore remains essential when a
new imputation learner is introduced. We validate the binary and ordinal
implementations, illustrate the ordinal method with wine quality, and report
an oracle count bridge without promoting the fitted count analysis.

# The surrogate idea

Let $D$ be a treatment, $X \in \mathbb{R}^p$ covariates, and $Y$ a
discrete outcome with conditional CDF $G_0(y\mid D,X)$. The true
distribution has a scalar additive index
$$\eta_0(D,X)=\theta_0D+g_0(X).$$ {#eq-index}
For binary and ordinal outcomes this is a latent-utility index: binary
$Y = \mathbf{1}\{Z > 0\}$, or ordinal,
$Y = j \iff \kappa_{j-1} < Z \le \kappa_j$ for thresholds
$-\infty = \kappa_0 < \kappa_1 < \dots < \kappa_{J-1} < \kappa_J = \infty$.
The latent utility follows a partially linear model

$$Z = \theta_0 D + g_0(X) + \varepsilon, \qquad
\varepsilon \mid D,X \sim F$$ {#eq-latent}
with $g_0$ an arbitrary nuisance function and $F$ a known error distribution
(logistic for proportional odds, extreme-value for complementary log-log,
and so on). The target $\theta_0$ is the treatment effect on this latent
scale, with $g_0$ left nonparametric. It is not a probability-scale average
treatment effect. Under a logistic link, $\exp(\theta_0)$ is a
proportional-odds ratio.
For Poisson or negative-binomial counts we use
$E[Y\mid D,X]=\exp\{\eta_0(D,X)\}$, making $\theta_0$ a log conditional-mean
effect and $\exp(\theta_0)$ a conditional rate ratio.

This notation matches the surrogate literature up to sign convention.
@liu2018 write the ordinal latent variable as
$Z=-f(X,\beta)+\varepsilon$, draw the surrogate from
$S\mid(Y,X)\sim Z\mid(Y,X)$, and define the surrogate residual
$R^S=S-E_0(S\mid X)$. Here $\eta=-f$, treatment is included in the index,
and the reference error is centered, so the corresponding correct-model
residual is $R^S=S-\eta(D,X)$. SUDO uses the surrogate $S$ as an outcome for
estimation rather than using $R^S$ only for diagnostics. The vector surrogate
of @cheng2021 is broader, but is not required for the scalar estimands studied
here.

Identification requires four conditions.

1. *Selection on observables.* No unobserved confounder acts beyond $X$.
   Together with the conditional location-family statement in @eq-latent,
   this assumption requires the latent error to have the same normalized law
   across treatment values given $X$.
2. *A constant index-scale effect.* The partially linear form in @eq-index
   makes $\theta_0$ a single number rather than a function of $X$.
3. *A correctly specified discrete conditional distribution.* Completion
   uses the entire fitted CDF jump, including the link and thresholds for
   cumulative models or the family and dispersion for counts.
4. *Overlap.* Treatment retains support after adjustment, so its residual is
   not annihilated.

The first and fourth are standard causal conditions. The second and third
are modeling choices shared with any structured distributional analysis.

## From a category to a continuous outcome

FWL estimates a treatment coefficient by regressing an outcome residual on a
treatment residual after removing $X$ from both. DML combines that identity
with cross-fitting and flexible nuisance regressions [@chernozhukov2018].
SUDO changes only the outcome supplied to this calculation.

Fit a structured distributional model of $Y$ on $(D,X)$. It must return the
scalar index $\hat\eta_i$, the fitted CDF immediately below and at the
observed outcome, and any global distribution parameters. Draw
$$
V_i\mid Y_i,D_i,X_i\sim
\operatorname{Unif}\{\hat G(Y_i-1\mid D_i,X_i),
                      \hat G(Y_i\mid D_i,X_i)\},
\qquad S_i=\hat\eta_i+H^{-1}(V_i),
$$ {#eq-randomized-pit}
where $H$ is a continuous mean-zero reference law. For binary and ordinal
cumulative links, we set $H$ to the centered link-error law. This gives the
truncated latent-variable surrogate of @liu2018, expressed through its CDF
jump, up to the corresponding intercept centering. For counts we use standard
normal $H$, following the randomized-quantile-residual convention of
@dunn1996. The reference law changes the completion variance but not the
target.

For the normal count reference, the completion has an analytic
Rao-Blackwell form. With $a=\hat G(Y-1\mid D,X)$,
$b=\hat G(Y\mid D,X)$, and standard-normal density $\phi$,
$$
E\{\Phi^{-1}(V)\mid Y,D,X\}
=\frac{\phi\{\Phi^{-1}(a)\}-\phi\{\Phi^{-1}(b)\}}{b-a},
$$
with the continuous limiting value for a zero-width numerical jump. The
count stages report this analytic completion beside the $B=25$ randomized
completion.

**Proposition 1 (randomized-CDF completion).** *If $G=G_0$ and
$\eta=\eta_0$, and if $H$ is continuous with
$E\{H^{-1}(V)\}=0$ for $V\sim\operatorname{Unif}(0,1)$, then the completion
in @eq-randomized-pit satisfies $E[S\mid D,X]=\eta_0(D,X)$. Consequently,*

$$S = \theta_0 D + g_0(X) + \eta, \qquad E[\eta \mid D, X] = 0$$

*and population FWL recovers $\theta_0$.*

*Proof.* Conditional on $(D,X)$, each discrete outcome selects one disjoint
CDF jump with probability equal to that jump's width. Uniform randomization
inside the selected jump therefore makes $V\mid D,X$ uniform on $(0,1)$.
The reference quantile has conditional mean zero, leaving $\eta_0(D,X)$.
$\square$

For cumulative links, this is the conditioning identity behind the
surrogate-variable results of @liu2018 and @cheng2021. The estimation step is
to use $S$, rather than only its diagnostic residual, as the outcome in FWL
or DML.

With an estimated distribution, the equality is asymptotic under consistency
and approximate otherwise. This limitation matters because the imputation
model is not protected by DML orthogonality.

The following base R example compares the fitted logistic coefficient, FWL
on the unobserved utility, and FWL on one surrogate completion.

```{r}
#| label: surrogate-fwl-demo
#| echo: true
set.seed(1)
n <- 40000
theta <- 1.5
X <- rnorm(n)
D <- rbinom(n, 1, plogis(0.8 * X))
Z <- theta * D + X + rlogis(n)
Y <- as.integer(Z > 0)

imputation_fit <- glm(Y ~ D + X, binomial)
index_hat <- predict(imputation_fit)
cdf_at_zero <- plogis(-index_hat)
uniform_draw <- ifelse(
  Y == 1,
  cdf_at_zero + runif(n) * (1 - cdf_at_zero),
  runif(n) * cdf_at_zero
)
surrogate <- index_hat + qlogis(
  pmin(pmax(uniform_draw, 1e-9), 1 - 1e-9)
)

treatment_residual <- resid(lm(D ~ X))
fwl <- function(outcome) {
  sum(treatment_residual * resid(lm(outcome ~ X))) /
    sum(treatment_residual^2)
}
round(c(
  glm = unname(coef(imputation_fit)["D"]),
  oracle = fwl(Z),
  surrogate = fwl(surrogate)
), 3)
```

All three estimates are close to $\theta_0 = 1.5$. Using only the binary
labels, the surrogate completion recovers the coefficient that FWL would
estimate if the latent utility were observed. Replacing the binary index
with an ordinal cumulative-link index and truncating to category-specific
intervals gives the ordinal construction. The remainder of the paper adds
two elements to this basic estimator: machine-learning partialling and
inference that accounts for the uncertainty hidden by a single completion.

# Estimation and inference

**The estimator.** Cross-fit the two adjustment nuisances
$\hat m(X) \approx E[D \mid X]$ and $\hat\ell(X) \approx E[S \mid X]$, and
form the residual-on-residual estimate

$$\hat\theta = \frac{\sum_i (S_i - \hat\ell(X_i))(D_i - \hat m(X_i))}
{\sum_i (D_i - \hat m(X_i))^2}$$ {#eq-theta}

The $\hat\ell$ regression uses $X$ only: the index carried $D$'s signal
into $S$, and partialling on $X$ alone isolates $\theta_0$. The score
$\varphi = (S - \ell(X) - \theta(D - m(X)))(D - m(X))$ is Neyman-orthogonal in
$(\ell, m)$: its derivatives in both nuisance directions vanish by the
tower property, so first-order error in $\hat\ell$ or $\hat m$ does not bias
$\hat\theta$. This protection covers the two adjustment nuisances, not the
imputation model that produced the surrogate.

**Proper multiple-imputation inference.** A single completion understates
uncertainty, so we draw $B$ surrogates and pool with Rubin's rules
[@rubin1987]: the estimate is $\bar\theta = B^{-1}\sum_b \hat\theta^{(b)}$
and the total variance is $T = \bar W + (1 + 1/B) B_{\!*}$, with a
Barnard-Rubin small-sample $t$ reference [@barnard1999]. Reusing one fitted
$\hat\beta$ across all completions leaves its shared estimation error
invisible to the between-draw variance.
A *proper* draw redraws the imputation model's parameters,
$\beta^{(b)} \sim N(\hat\beta, \widehat{\mathrm{Cov}}(\hat\beta))$
(thresholds included for ordinal outcomes), before each completion. This is
the large-sample normal form of a proper posterior-predictive draw. It
accounts for sampling noise, surrogate randomness, and imputation-model
uncertainty. @tbl-ladder shows why all three matter.

For ordinal outcomes, thresholds are parameters too and must be perturbed
jointly with the coefficients. Surrogate-residual mean and variance checks
provide a diagnostic for the assumed link and mean structure [@liu2018], but
they do not replace target-level validation.

## What completion adds to the fitted index

Let $R_D=D-m_0(X)$ and let a fitted distribution learner return the pointwise
index $\hat\eta(D,X)$ without requiring a treatment coefficient. Write
$\mu_Y=E\{H^{-1}(V)\mid Y,D,X,\hat G\}$ for the conditional mean reference
correction. The randomized estimator converges to the same completion mean
as its Rao-Blackwellized form.

**Proposition 2 (index projection plus correction).** *With the population
treatment nuisance, the population completion target is*
$$
\theta_{\mathrm{SUDO}}
=\frac{E\{R_D\hat\eta(D,X)\}}{E(R_D^2)}
 +\frac{E(R_D\mu_Y)}{E(R_D^2)}.
$$ {#eq-coefficient-correction}

*Proof.* Substitute $S=\hat\eta(D,X)+\mu_Y$ into the population FWL
numerator and separate its two terms. If
$\hat\eta(D,X)=\hat\alpha D+\hat f(X)$, then
$E\{R_D\hat f(X)\}=0$ and $E(R_D D)=E(R_D^2)$, so the first term reduces to
$\hat\alpha$. $\square$

SUDO therefore projects the fitted index onto the residualized treatment and
applies an outcome-informed correction. For a structured additive fit the
projection is its treatment coefficient. The correction can help when
treatment regularization shrinks the index projection, hurt through
covariate-direction error, or vanish. It is not a generic improvement over a
direct coefficient.
For binary logit, the correction reduces to an entropy-weighted residual
moment.

## Where orthogonality stops

The orthogonality established above protects $\hat\theta$ from error in the
adjustment nuisances but says nothing about the imputation model, whose
error is baked into the completed data $S$ rather than differenced away.
The imputation model sits outside the orthogonality guarantee because its
error is built into the outcome supplied to DML.

Here, a *black-box imputation model* means a flexible
distributional learner that supplies a pointwise scalar index and conditional
CDF. It need not parameterize or expose a treatment coefficient. The count
validation isolates $D$ from the joint $X$ interaction block. An arbitrary
probability vector without a scalar index on the target scale remains outside
scope. Flexible imputation models are cross-fit so each supplied index is an
out-of-fold prediction.

Either way $\hat\eta$ is fixed with respect to the completion, while
$Y \mid D, X$ is generated by the truth. Since
$S$ is $\hat\eta$ plus truncated noise,

$$E[S \mid D, X] = \psi(\hat\eta, \eta), \quad
\psi(v, e) = v + F(e)\,\mu_+(v) + (1 - F(e))\,\mu_-(v)$$ {#eq-ess}

with $\mu_\pm$ the truncated-noise mean corrections. The correctly centered
draw is exact, $\psi(e, e) = e$, the correct-model surrogate identity in
@liu2018 and @cheng2021,
but index error passes through with a factor strictly below one,

$$c(e) = \left.\frac{\partial \psi}{\partial v}\right|_{v=e}
= 1 - f(-e)\,\bigl(\mu_+(e) - \mu_-(e)\bigr), \qquad f = F'$$ {#eq-ce}

Read $c$ as the fraction of index error that survives the truncation. With
no truncation the surrogate would be index plus noise and its mean would move
one-for-one with $\hat\eta$, giving $c = 1$; the observed category drags it
back, and $c$ is what is left.

Under a logistic link, $c$ rises from $1-\log 2=0.31$ at $e=0$ toward one
in the tails (@fig-passthrough). In fact,
$$c(e) = 1 - H\bigl(F(e)\bigr), \qquad H(p) = -p\log p - (1-p)\log(1-p),$$ {#eq-centropy}
so self-correction equals outcome entropy. An uncertain label sharply
constrains the latent utility and corrects more of the fitted-index error. A
nearly deterministic label adds little information and corrects almost none.

```{r}
#| label: fig-passthrough
#| echo: false
#| fig-width: 5
#| fig-height: 2.6
#| fig-cap: "Fraction of fitted-index error that survives completion under a logistic link. An informative label supplies the strongest correction near the center; the correction fades as the label becomes nearly deterministic."
mu_plus <- function(index) {
  integrate(function(error) error * dlogis(error), -index, Inf)$value /
    plogis(index)
}
mu_minus <- function(index) {
  integrate(function(error) error * dlogis(error), -Inf, -index)$value /
    plogis(-index)
}
pass_through <- function(index) {
  1 - dlogis(index) * (mu_plus(index) - mu_minus(index))
}
index <- seq(0, 6, by = 0.05)
pass_through_value <- vapply(index, pass_through, numeric(1))
par(mar = c(4, 4, 1, 1), cex = 0.9)
plot(
  index, pass_through_value,
  type = "l", lwd = 2, ylim = c(0, 1), las = 1,
  xlab = "index level e",
  ylab = expression("pass-through factor " * c(e))
)
abline(h = 1, lty = 3, col = "grey60")
origin_value <- pass_through(0)
points(0, origin_value, pch = 19)
text(
  0.1, origin_value + 0.1, sprintf("%.2f", origin_value),
  pos = 4, cex = 0.9
)
```

Fitted-index error enters the SUDO target at first order. If the effective
treatment component is displaced by $\delta_D$, the leading bias is
approximately $\bar c_1\delta_D$, where $\bar c_1$ averages $c(e)$ over
the treatment-relevant index distribution. Error in the covariate component
can also survive because completion is nonlinear in the fitted index.
Partialling-out does not erase either channel. Target-level bias, not
predictive fit alone, is therefore the relevant validation criterion.

# Machine-learning imputation models

A parametric distribution model supplies both an index and a covariance
matrix, so proper draws are immediate. A black-box learner supplies neither
a treatment coefficient nor a joint parameter covariance. We therefore ask
only for pointwise index and CDF predictions and handle structure and
inference outside the learner.

For the validated binary experiments, we restrict the imputation index to
$\hat\alpha D+\hat f(X)$. A GAM, MARS bases built from $X$ only, and
neural-network backfitting all implement this structure. The restriction
prevents treatment-by-covariate interactions, but it does not guarantee a
good fit or a small target bias. Two other partially linear learners failed
the bias criterion, including one tuned by predictive risk.

When a learner has no usable covariance, the full-pipeline bootstrap
resamples the observations and reruns the entire procedure, including the
imputation learner, folds, completions, and adjustment nuisances. A
within-fold refit misses important sources of variation. In the validated
binary designs, normal intervals based on the full-pipeline bootstrap
standard deviation attain nominal coverage.

These results support a workflow, not a blanket theorem for arbitrary
learners: impose a scalar-index structure suited to the estimand, tune on
target-level simulation bias, and require coverage validation before
promotion.

# Simulation study {#sec-sims}

The simulations draw independent standard-normal covariates, assign treatment
from those covariates, and coarsen a partially linear latent utility:

$$X_1, \dots, X_p \stackrel{\text{iid}}{\sim} N(0, 1), \qquad
D \mid X \sim \mathrm{Bernoulli}\bigl(\Lambda(\pi(X))\bigr), \qquad
Z = \theta_0 D + g_0(X) + \varepsilon,$$ {#eq-dgp}
$$\varepsilon \sim F, \qquad
Y = j \iff \kappa_{j-1} < Z \le \kappa_j,$$

The correctly specified one-covariate design isolates inference. The
five-covariate machine-learning design uses
$$\pi(X) = X_4 + \cos X_5, \qquad g_0(X) = X_1^2 + \sin X_2 + 0.5 X_3,$$
with binary or three-category ordinal outcomes. @tbl-dgp lists the full
designs.

The coverage target is 0.95. Standard deviation (SD) is run-to-run
variability; mean standard error (mean SE) is the average reported
uncertainty. The tables are generated from committed result files.

```{r}
#| label: results-setup
#| include: false
results_dir <- "../../R/results"
read_result <- function(filename) {
  read.csv(file.path(results_dir, filename), check.names = FALSE)
}
sg <- function(x, d = 3) sprintf(paste0("%+.", d, "f"), x)
nn <- function(x, d = 3) sprintf(paste0("%.", d, "f"), x)
```

**Proper draws are what deliver coverage.** With a simple binary design and
a correct parametric imputation model, only the proper scheme reaches
nominal coverage; the naive interval covers 0.80 and the improper 0.92
(@tbl-ladder).

```{r}
#| label: tbl-ladder
#| echo: false
#| tbl-cap: "Inference schemes, binary outcome, $n = 5000$, $\\theta_0 = 1.5$, 500 replications. $B$ is the number of surrogate draws."
s2 <- read_result("stage2_summary.csv")
estimators <- c("naive", "improper_B25", "proper_B25")
result <- s2[match(estimators, s2$estimator), ]
table_data <- data.frame(
  Scheme     = c("Naive single draw", "Improper, $B=25$", "Proper, $B=25$"),
  Bias       = sg(result$bias),
  SD         = nn(result$sd),
  `Mean SE`  = nn(result$mean_se),
  Coverage   = nn(result$coverage),
  check.names = FALSE)
knitr::kable(table_data, align = c("l", "r", "r", "r", "r"))
```

**Ordinal outcomes.** With three categories and machine-learning adjustment,
SUDO meets the bias criterion with valid coverage (@tbl-ordinal). The
outcome nuisance $E[S\mid X]$ must be refit for each ordinal completion;
reusing one fit inflated the within-draw variance. The corrected procedure
has bias 0.016 and coverage 0.960 in 200 replications.

```{r}
#| label: tbl-ordinal
#| echo: false
#| tbl-cap: "Ordinal outcome, $J = 3$, $\\theta_0 = 1.0$. Parametric arm 500 replications; corrected spline + ML arm 200, with the outcome nuisance refit per completion."
s4 <- read_result("stage4_summary.csv")
s5 <- read_result("stage5r_summary.csv")
parametric_result <- s4[s4$estimator == "sudo_ordinal_clm", ]
refit_result <- s5[s5$estimator == "ordinal_refit_n2000_B25", ]
table_data <- data.frame(
  Configuration = c("Parametric, $n=3000$", "Spline + ML nuisances, $n=2000$"),
  Bias      = sg(c(parametric_result$bias, refit_result$bias)),
  SD        = nn(c(parametric_result$sd, refit_result$sd)),
  `Mean SE` = nn(c(parametric_result$mean_se, refit_result$mean_se)),
  Coverage  = nn(c(parametric_result$coverage, refit_result$coverage)),
  check.names = FALSE)
knitr::kable(table_data, align = c("l", "r", "r", "r", "r"))
```

**Direct and tailored comparisons.** SUDO does not dominate a correct
outcome-specific estimator (@tbl-no-superiority). In the correct ordinal
parametric design, SUDO and the direct cumulative-link coefficient agree. Under a
cloglog truth, SUDO, the tailored orthogonal score, and the direct cloglog
coefficient have indistinguishable bias. In the high-dimensional binary
design, SUDO attenuates treatment-regularization bias but does not remove it.
These comparisons define the method's scope rather than rank efficiency.

```{r}
#| label: tbl-no-superiority
#| echo: false
#| tbl-cap: "Direct and tailored-score comparisons. The table reports the validated common-sample cells most relevant to the scope claim."
s8 <- read_result("stage8_summary.csv")
s11 <- read_result("stage11_debiased_comparison.csv")
s12 <- read_result("stage12_link_flexibility.csv")
o <- s8[s8$design == "D1_correct" &
          s8$estimator %in% c("clm_direct", "sudo"), ]
h <- s11[s11$n == 2000 & s11$arm == "pen" &
           s11$estimator %in% c("PL_OS", "sudo"), ]
l <- s12[s12$estimator %in% c("sudo_cloglog", "plos_cloglog",
                              "glm_cloglog"), ]
table_data <- rbind(
  data.frame(Design = ifelse(o$design == "D1_correct",
                             "Ordinal, correct", "Ordinal, misspecified"),
             Estimator = ifelse(o$estimator == "clm_direct", "Direct", "SUDO"),
             Bias = sg(o$bias), SD = nn(o$sd)),
  data.frame(Design = paste0("Binary high-dimensional, ", h$arm),
             Estimator = ifelse(h$estimator == "PL_OS",
                                "Tailored orthogonal", "SUDO"),
             Bias = sg(h$bias), SD = nn(h$sd)),
  data.frame(Design = "Binary cloglog, correct link",
             Estimator = c("Direct", "Tailored orthogonal", "SUDO")[
               match(l$estimator, c("glm_cloglog", "plos_cloglog",
                                    "sudo_cloglog"))],
             Bias = sg(l$bias), SD = nn(l$sd))
)
knitr::kable(table_data, align = c("l", "l", "r", "r"))
```

**Link misspecification.** Generating Gumbel noise but assuming a logistic
link substantially biases the latent-scale effect (+0.77 for binary, +0.58
for ordinal outcomes, with zero coverage). The correct link removes this
bias (bias $<0.005$). Conditional second-moment drift in the surrogate
residual flags the wrong fit but is not specific to the link, so we use it as
a diagnostic rather than a formal test [@liu2018].

**Partially-linear imputation learners.** GAM, MARS, and neural backfitting
fit the same $\alpha D+f(X)$ structure. All three pass the prespecified
bias criterion, and full-pipeline-bootstrap normal coverage ranges from 0.94
to 0.98 (@tbl-pl-learners). With 50 replications, these differences are too
small to rank the learners.

```{r}
#| label: tbl-pl-learners
#| echo: false
#| tbl-cap: "Partially-linear imputation learners with full-pipeline bootstrap, $n=2000$, 50 replications, 50 outer resamples, and 15 surrogate completions per resample."
s3s <- read_result("stage3s_summary.csv")
learner_name <- c(
  pl_gam = "GAM",
  pl_mars = "MARS",
  pl_backfit = "Neural backfit"
)
table_data <- data.frame(
  Learner = unname(learner_name[s3s$learner]),
  `$\\theta_0$` = nn(s3s$theta, 1),
  Bias = sg(s3s$bias),
  `SD/Mean SE` = nn(s3s$sd_over_mean_se),
  `Normal coverage` = nn(s3s$coverage_normal),
  `Percentile coverage` = nn(s3s$coverage_percentile),
  check.names = FALSE
)
knitr::kable(table_data, align = c("l", "r", "r", "r", "r", "r"))
```

Partial linearity is necessary in these experiments but not sufficient. Two
other partially linear learners fail the bias screen, including one selected
by cross-validated predictive risk. Prediction error is therefore a
diagnostic, not a tuning target for the treatment coefficient. The validated
configurations were screened on target-level Monte Carlo bias and evaluated
on separate seeds; a data-only selection rule remains open.

With continuous treatment, bias remains below 1.5% of the truth in the binary
and ordinal designs. Strong-signal coverage is less stable, reinforcing the
need to rerun the full-pipeline bootstrap rather than transfer an uncertainty
result from a binary-treatment experiment.

## Count oracle bridge

We isolate count completion from fitted-learner error by supplying the true
scalar index and count CDF. The six
prespecified cells cross Poisson and negative-binomial laws, binary and
continuous treatments, and two signal strengths where applicable. At
$n=2000$, with 100 replications, $B=25$ randomized completions, five folds,
and 99 bootstrap resamples, both analytic Rao-Blackwell and randomized
completion pass every bias, coverage, and SD-to-SE gate
(@tbl-count-oracle). Their paired differences are within two completion
Monte Carlo standard errors in every cell.

```{r}
#| label: tbl-count-oracle
#| echo: false
#| tbl-cap: "Count oracle bridge with the true index and CDF supplied. Entries pair Rao-Blackwell and randomized completion. Each cell uses $n=2000$, 100 replications, five folds, and 99 bootstrap resamples; randomized SUDO uses $B=25$."
s15o <- read_result("stage15o_count_oracle.csv")
cell_order <- c(
  "continuous_poisson_tlog1p5", "continuous_negbin_tlog1p5",
  "continuous_poisson_tlog2", "continuous_negbin_tlog2",
  "binary_poisson_tlog2", "binary_negbin_tlog2"
)
estimator_order <- c("rao_blackwell", "randomized")
s15o$cell <- factor(s15o$cell, levels = cell_order)
s15o$estimator <- factor(s15o$estimator, levels = estimator_order)
result <- s15o[order(s15o$cell, s15o$estimator), ]
rao_blackwell <- result[result$estimator == "rao_blackwell", ]
randomized <- result[result$estimator == "randomized", ]
stopifnot(identical(as.character(rao_blackwell$cell),
                    as.character(randomized$cell)))
table_data <- data.frame(
  Design = paste(
    ifelse(rao_blackwell$treatment == "continuous", "Continuous", "Binary"),
    ifelse(rao_blackwell$family == "poisson", "Poisson", "NB"),
    sep = ", "
  ),
  `$\\theta_0$` = ifelse(
    abs(rao_blackwell$truth - log(1.5)) < 1e-8,
    "$\\log(1.5)$", "$\\log(2)$"
  ),
  `Bias (RB / randomized)` = paste(
    sg(rao_blackwell$bias), sg(randomized$bias), sep = " / "
  ),
  `SD/SE (RB / randomized)` = paste(
    nn(rao_blackwell$sd / rao_blackwell$mean_se),
    nn(randomized$sd / randomized$mean_se),
    sep = " / "
  ),
  `Coverage (RB / randomized)` = paste(
    nn(rao_blackwell$coverage), nn(randomized$coverage), sep = " / "
  ),
  check.names = FALSE
)
knitr::kable(table_data, align = c("l", "c", "r", "r", "r"))
```

This bridge validates the count completion and bootstrap when the supplied
distribution is correct. It does not validate a fitted count learner. The
constrained-XGBoost point screen failed its promotion gate, so fitted count
support and the prespecified bikeshare application remain outside the paper's
claims.

# Application: volatile acidity and wine quality

We apply ordinal SUDO to the UCI wine-quality data [@cortez2009]. Quality is
the median ordinal score from blind tasters. We analyze the 1,599 red and
4,898 white wines separately; observed scores range from 3 to 8 for red wine
and 3 to 9 for white. The treatment is standardized volatile acidity, an
acetic-acid fault associated with vinegar-like aroma. The adjustment set is
the other ten standardized physicochemical measurements.

The data are observational. A causal reading requires the measured chemistry
to be a sufficient pre-treatment adjustment set and winemaking practice to
affect quality only through that chemistry and volatile acidity
(@fig-dag). We do not claim that these assumptions are established. The
analysis illustrates the method on a genuinely ordinal outcome.

```{mermaid}
%%| label: fig-dag
%%| fig-width: 5.2
%%| fig-pos: H
%%| fig-cap: "Assumed application DAG. Volatile acidity (the treatment) affects latent quality directly; unobserved winemaking practice confounds both through the measured chemistry, the adjustment set."
flowchart LR
  P((winemaking<br/>practice)):::u --> VA[volatile acidity]
  P --> X[other chemistry]
  X --> VA
  VA --> Q[latent quality]
  X --> Q
  classDef u fill:#eee,stroke-dasharray:4 3;
```

\FloatBarrier

Volatile acidity retains 75% of its marginal standard deviation after
adjustment in red wine and 94% in white wine. Both imputation models are
proportional-odds logits with volatile acidity entering linearly. One uses
the adjustment variables linearly; the other uses a three-degree-of-freedom
natural spline for each variable, with no treatment-by-covariate interactions.
Ordinal SUDO uses 50 proper draws, five-fold cross-fitted GAM nuisances, and
Rubin pooling. The two specifications give similar estimates (@tbl-wine).

Because volatile acidity is standardized, the coefficients are per one
standard deviation, about 0.18 g/dm$^3$ for red wine and 0.10 for white.
A one-SD increase multiplies the odds of scoring above any fixed quality
level by about 0.59 for red wine and 0.64 for white wine. The direct
cumulative-link coefficient is included as a model-based comparator.

```{r}
#| label: tbl-wine
#| echo: false
#| tbl-cap: "Effect of a one-SD increase in volatile acidity on latent wine quality. Model-based coefficient from the linear cumulative-link fit (standard error in parentheses); SUDO estimates with 95% intervals in brackets."
wine_results <- read_result("wine_application.csv")
wine_row <- function(wine_name, label) {
  linear <- wine_results[
    wine_results$wine == wine_name & wine_results$full_model == "linear",
  ]
  spline <- wine_results[
    wine_results$wine == wine_name & wine_results$full_model == "pl_spline",
  ]
  data.frame(
    Wine = label,
    `Overlap $R^2(D \\sim X)$` = nn(linear$overlap_r2, 2),
    `Cumulative-link` = sprintf(
      "$%.2f$ (%.2f)", linear$naive_theta, linear$naive_se
    ),
    `SUDO (linear)` = sprintf(
      "$%.2f$ [$%.2f,%.2f$]",
      linear$sudo_theta, linear$ci_lo, linear$ci_hi
    ),
    `SUDO (spline)` = sprintf(
      "$%.2f$ [$%.2f,%.2f$]",
      spline$sudo_theta, spline$ci_lo, spline$ci_hi
    ),
    check.names = FALSE)
}
table_data <- rbind(wine_row("red", "Red"), wine_row("white", "White"))
knitr::kable(table_data, align = c("l", "r", "r", "r", "r"))
```

The negative sign is stable when any one adjustment covariate is removed.
An omitted-variable calibration suggests that a confounder would need to
explain roughly 15% of the residual treatment and outcome variation for red
wine and 10% for white wine to move the estimate to zero [@cinelli2020].
Because that calibration was derived for linear regression with an observed
outcome, we report it as a heuristic rather than a formal SUDO bound.

The completion itself provides substantial self-correction in this
application: the fitted pass-through averages 0.21 for red wine and 0.17 for
white wine. This does not establish robustness to an arbitrary wrong
imputation model, but it shows why the ordinal categories carry useful
information beyond the fitted index.

# Discussion

SUDO turns a discrete-outcome problem into a familiar continuous
partialling-out problem. The fitted distribution identifies the target
scale; randomized completion restores that scale in conditional mean; DML
then removes flexible covariate effects. This separation is the main value
of the method.

The method does not dominate a correct direct or tailored estimator. Its
advantage is a common interface, especially for ordinal outcomes for which a
tailored orthogonal score is not readily available. Proper draws or a
full-pipeline bootstrap are part of the estimator, not optional refinements.

The same separation also exposes the central risk. DML orthogonality protects
the adjustment regressions, but not the distribution learner that creates
the surrogate. Partial linearity prevents treatment-by-covariate interactions
but does not guarantee small treatment bias. Each new learner and outcome
family therefore needs target-level simulation and coverage checks.

The validated contribution is currently binary and ordinal. The count oracle
bridge shows that the completion calculation extends cleanly to Poisson and
negative-binomial laws, but the fitted count learner did not pass promotion.
The next methodological need is stable distribution learners and data-driven
tuning rules that control target bias.

# Data and software

The wine-quality data come from the University of California, Irvine Machine
Learning Repository [@cortez2009]. The `sudo` Python package provides the
parametric estimator for binary and ordinal outcomes, including proper draws
and the per-completion ordinal outcome-nuisance refit. The partially-linear
machine-learning imputation learners and the full-pipeline bootstrap are
implemented in R, which is also where every method is prototyped and
validated before release. Count and flexible ordinal adapters remain R-only
until their fitted-learner confirmation stages pass. The open-source
repository contains the code and committed result files behind every table
and figure under the MIT License:
<https://github.com/bgreenwell/sudo>.

# Declaration of artificial intelligence use

We used generative artificial intelligence (AI) to assist with code and
summaries of reference materials. The authors reviewed and validated all
outputs, retain full ownership of the final manuscript, and accept sole
responsibility for any errors.

# Appendix: simulation and application details {#sec-appendix}

Each design draws a latent utility
$Z = \theta_0 D + g_0(X) + \varepsilon$ and converts it to an observed
category. Most designs use logistic $\varepsilon$; the exceptions are
specified below. The designs vary the covariate structure, cutpoints, and
fitted models; covariates remain standard normal throughout. The `R/` and
`R/results/` directories contain the full data-generating code and committed
per-cell results.

## Data-generating processes

All designs instantiate @eq-dgp with one of two covariate structures:

- **S1**, one covariate: $p = 1$, $\pi(X) = 0.8X$, $g_0(X) = X$. Every model
  in play is correctly specified, so only the pooling is under test.
- **S5**, five covariates: $p = 5$, $\pi(X) = X_4 + \cos X_5$,
  $g_0(X) = X_1^2 + \sin X_2 + 0.5 X_3$. The outcome structure is nonlinear
  in $X$, and treatment depends on covariates absent from $g_0$.

| Design | Structure | $F$ | $\kappa$ | $\theta_0$ | $n$ | Replications |
|---|---|---|---|---|---|---|
| Inference ladder (@tbl-ladder) | S1 | logistic | $(0)$ | 1.5 | 5000 | 500 |
| Ordinal parametric (@tbl-ordinal, top) | S1 | logistic | $(-1, 1)$ | 1.0 | 3000 | 500 |
| Binary ML | S5 | logistic | $(0)$ | 0.5, 3 | 1000, 2000 | 200 |
| Ordinal flexible adjustment (@tbl-ordinal, bottom) | S5 | logistic | $(0, 2)$ | 1.0 | 1000, 2000 | 200 |
| Link misspecification | S1 | Gumbel | $(0)$ or $(0, 2)$ | 1.0 | 5000 | 200 |
| Black-box imputation | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 100 |
| PL learner comparison (@tbl-pl-learners) | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 50 |
| DGP variation, continuous $D$ | S5, $D = 0.7\pi(X) + N(0,1)$ | logistic | $(0)$ or $(0, 2)$ | 1, 2.5 | 2000 | 200 |

: Data-generating processes. $\kappa$ lists the interior cutpoints, so
$(0)$ is binary and a pair gives $J = 3$. {#tbl-dgp}

**Models fitted.** The ladder uses $\mathrm{glm}(Y \sim D + X)$ as the
imputation model with $B \in \{10, 25, 50\}$ draws (the table shows
$B = 25$). The ordinal parametric design uses `ordinal::clm`, whose
covariance covers the thresholds, with $B = 25$. The ordinal
flexible-adjustment design uses parametric `clm` on $D$ and natural-spline
bases $\mathrm{ns}(X_k, 4)$ with `mgcv::gam`
partialling nuisances and $B = 25$.

**Link misspecification.** This design retains the one-covariate structure
but replaces logistic $\varepsilon$ with an extreme-value error: Gumbel-max
for the binary case, matching `glm` with a cloglog link, and Gumbel-min for
the ordinal case, matching `ordinal::clm` with a cloglog link. The analyst
fits either logit (misspecified) or cloglog (correct). Holding the simple
covariate structure fixed isolates error in the link.

**Partially-linear imputation models.** The original black-box result uses
$\hat\eta = \hat\alpha D + \hat f(X)$ with $\hat f$ a single-hidden-layer
neural net (`nnet`, `size = 4`, `decay = 0.3`, five backfitting iterations,
selected on target-level Monte Carlo bias) and 100 full-pipeline
bootstrap resamples. The common comparison uses 50 resamples per arm. The
GAM includes a linear treatment term and smooths of $X$. MARS builds
degree-two bases from $X$ only before a joint logistic fit with mandatory
linear $D$.

**Wine application.** Red and white wines are fit separately. The outcome
levels are 3--8 for red wine and 3--9 for white. Volatile acidity is the
standardized treatment. The ten standardized adjustment variables are fixed
acidity, citric acid, residual sugar, chlorides, free and total sulfur
dioxide, density, pH, sulphates, and alcohol. The proportional-odds logit
full model is `ordinal::clm` with index
$\alpha D+\beta^\top X$ (linear) or
$\alpha D+\sum_k \mathrm{ns}(X_k,3)$ (spline). Partialling nuisances
$E[S \mid X]$ and $E[D \mid X]$ use `mgcv::gam` (Gaussian for continuous
$D$), with five-fold cross-fitting, $B=50$ proper draws, and Rubin pooling.

## Nuisances, cross-fitting, and draws

Partialling nuisances $\hat\ell(X) = E[S \mid X]$ and $\hat m(X) = E[D \mid X]$
are `mgcv::gam` throughout, fit with five-fold cross-fitting. Proper draws
sample the full-model parameters from $N(\hat\beta, \widehat{\mathrm{Cov}}(\hat\beta))$
(thresholds included for `clm`) before each completion. Diagnostic draws use
a random number generator (RNG) stream independent of the estimation draws.

Binary experiments found per-completion refitting of $\hat\ell$ negligible,
so the binary estimator reuses a plug-in fit. The ordinal estimator refits
$\hat\ell$ for every completion because reuse inflated its within-draw
variance.

## Monte Carlo precision

Coverage is a binomial proportion, so its Monte Carlo standard error at a
true 0.95 is about $\sqrt{0.95 \cdot 0.05 / R}$: 0.010 at $R = 500$
replications, 0.015 at $R = 200$, and 0.022 at $R = 100$. Per-cell bias and
coverage Monte Carlo standard errors are in the committed result files.

# References

::: {#refs}
:::
