---
title: "Surrogate-Assisted Double Machine Learning for Discrete Outcomes"
author:
  - name: Brandon M. Greenwell
    email: greenwbm@ucmail.uc.edu
    affiliations:
      - name: "University of Cincinnati Carl H. Lindner College of Business"
        department: Department of Operations, Business Analytics, and Information Systems
        address: 2906 Woodside Drive
        city: Cincinnati, Ohio
        country: United States
        postal-code: 45221
    orcid: 0000-0002-8120-0084
  - name: Other Authors
    email: other.authors@email.com
    affiliations:
      - name: "University of XYZ"
authorcols: true
date: today
abstract: |
  Applying partialling-out directly to a discrete label targets the label's
  numeric scale, not a cumulative-link or log-mean coefficient.
  Surrogate-Assisted Double Machine Learning (SUDO) restores the target scale
  before partialling-out. A fitted distribution supplies a scalar index and
  conditional CDF; randomization within the observed CDF jump then produces a
  continuous surrogate whose conditional mean equals the true index when the
  distribution is correct. Standard Frisch-Waugh-Lovell partialling-out can
  then recover the treatment coefficient. Proper multiple-imputation draws
  provide inference for parametric fits, while flexible learners require a
  full-pipeline bootstrap. Simulations establish nominal coverage for the
  validated binary and ordinal designs and show the central limitation:
  double-machine-learning orthogonality does not protect against error in the
  imputation distribution. Scalar recalibration removes regularization bias
  from a coefficientless additive learner in three designs, but fails at a
  high-dimensional boundary. SUDO agrees with correct direct and tailored
  binary estimators; its practical value is a common route to latent-scale
  estimation, especially for ordinal outcomes. A wine-quality analysis
  illustrates that use. An oracle experiment validates count completion, but
  no fitted count learner is yet supported.
format:
  arxiv-pdf:
    keep-tex: false
bibliography: references.bib
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}
---

# Introduction

Many important outcomes are discrete: relapse is binary, survey responses
are ordinal, and trips are counts. Double Machine Learning [DML,
@chernozhukov2018] handles a continuous partially linear outcome by removing
flexibly estimated covariate effects and regressing one residual on another.
The same calculation can be applied to a discrete label, but it then targets
a contrast in $E[Y\mid D,X]$. It does not recover a cumulative-link
coefficient or a log conditional-mean effect. The obstacle is the outcome
scale.

SUDO changes the outcome before applying DML. It draws a continuous value
inside the fitted CDF jump for the observed category and adds the fitted
scalar index. If the conditional distribution is correct, the completed
outcome has that index as its conditional mean. The treatment coefficient is
therefore available through ordinary partialling-out.

This construction turns the surrogate-variable method of @liu2018 from a
diagnostic into an estimator. @cheng2021 extend surrogate residuals to
unordered discrete-choice models with vector utilities, and @greenwell2018
implement binary and ordinal diagnostics in the `sure` package. SUDO instead
focuses on scalar-index distributions. The learner supplies an index and CDF;
SUDO supplies completion, partialling-out, and uncertainty propagation.

The common interface is useful, but it is not an efficiency claim. Tailored
orthogonal scores already exist for logistic binary models [@liu2021lplr]
and Poisson log-mean effects [@drukker2022]. SUDO agrees with those estimators
when their models are correct. Its clearest use is ordinal regression, where
the same construction targets a latent proportional-odds coefficient without
deriving a new score. The simulations also establish the method's main
boundary: DML protects the adjustment regressions, not the distribution that
creates the surrogate. New imputation learners therefore require direct bias
and coverage validation.

# From a discrete outcome to the target scale

Let $D$ be a treatment, $X \in \mathbb{R}^p$ covariates, and $Y$ a
discrete outcome with conditional CDF $G_0(y\mid D,X)$. The true
distribution has a scalar additive index
$$\eta_0(D,X)=\theta_0D+g_0(X).$$ {#eq-index}
For binary and ordinal outcomes this is a latent-utility index: binary
$Y = \mathbf{1}\{Z > 0\}$, or ordinal,
$Y = j \iff \kappa_{j-1} < Z \le \kappa_j$ for thresholds
$-\infty = \kappa_0 < \kappa_1 < \dots < \kappa_{J-1} < \kappa_J = \infty$.
The latent utility follows a partially linear model

$$Z = \theta_0 D + g_0(X) + \varepsilon, \qquad
\varepsilon \mid D,X \sim F$$ {#eq-latent}
with $g_0$ an arbitrary nuisance function and $F$ a known error distribution
(logistic for proportional odds, extreme-value for complementary log-log,
and so on). The target $\theta_0$ is the treatment effect on this latent
scale, with $g_0$ left nonparametric. It is not a probability-scale average
treatment effect. Under a logistic link, $\exp(\theta_0)$ is a
proportional-odds ratio.
For Poisson or negative-binomial counts we use
$E[Y\mid D,X]=\exp\{\eta_0(D,X)\}$, making $\theta_0$ a log conditional-mean
effect and $\exp(\theta_0)$ a conditional rate ratio.

This notation matches the surrogate literature up to sign convention.
@liu2018 write the ordinal latent variable as
$Z=-f(X,\beta)+\varepsilon$, draw the surrogate from
$S\mid(Y,X)\sim Z\mid(Y,X)$, and define the surrogate residual
$R^S=S-E_0(S\mid X)$. Here $\eta=-f$, treatment is included in the index,
and the reference error is centered, so the corresponding correct-model
residual is $R^S=S-\eta(D,X)$. SUDO uses the surrogate $S$ as an outcome for
estimation rather than using $R^S$ only for diagnostics. The vector surrogate
of @cheng2021 is broader, but is not required for the scalar estimands studied
here.

Identification requires four conditions.

1. *Selection on observables.* No unobserved confounder acts beyond $X$.
   Together with the conditional location-family statement in @eq-latent,
   this assumption requires the latent error to have the same normalized law
   across treatment values given $X$.
2. *A constant index-scale effect.* The partially linear form in @eq-index
   makes $\theta_0$ a single number rather than a function of $X$.
3. *A correctly specified discrete conditional distribution.* Completion
   uses the entire fitted CDF jump, including the link and thresholds for
   cumulative models or the family and dispersion for counts.
4. *Overlap.* Treatment retains support after adjustment, so its residual is
   not annihilated.

The first and fourth are standard causal conditions. The second and third
are modeling choices shared with any structured distributional analysis.

## Randomized completion

FWL estimates a treatment coefficient by regressing an outcome residual on a
treatment residual after removing $X$ from both. DML combines that identity
with cross-fitting and flexible nuisance regressions [@chernozhukov2018].
SUDO changes only the outcome supplied to this calculation.

Fit a structured distributional model of $Y$ on $(D,X)$. It must return the
scalar index $\hat\eta_i$, the fitted CDF immediately below and at the
observed outcome, and any global distribution parameters. Draw
$$
V_i\mid Y_i,D_i,X_i\sim
\operatorname{Unif}\{\hat G(Y_i-1\mid D_i,X_i),
                      \hat G(Y_i\mid D_i,X_i)\},
\qquad S_i=\hat\eta_i+H^{-1}(V_i),
$$ {#eq-randomized-pit}
where $H$ is a continuous mean-zero reference law. For binary and ordinal
cumulative links, we set $H$ to the centered link-error law. This gives the
truncated latent-variable surrogate of @liu2018, expressed through its CDF
jump, up to the corresponding intercept centering. For counts we use standard
normal $H$, following the randomized-quantile-residual convention of
@dunn1996. The reference law changes the completion variance but not the
target.

For the normal count reference, the completion has an analytic
Rao-Blackwell form. With $a=\hat G(Y-1\mid D,X)$,
$b=\hat G(Y\mid D,X)$, and standard-normal density $\phi$,
$$
E\{\Phi^{-1}(V)\mid Y,D,X\}
=\frac{\phi\{\Phi^{-1}(a)\}-\phi\{\Phi^{-1}(b)\}}{b-a},
$$
with the continuous limiting value for a zero-width numerical jump. The
count stages report this analytic completion beside the $B=25$ randomized
completion.

**Proposition 1 (randomized-CDF completion).** *If $G=G_0$ and
$\eta=\eta_0$, and if $H$ is continuous with
$E\{H^{-1}(V)\}=0$ for $V\sim\operatorname{Unif}(0,1)$, then the completion
in @eq-randomized-pit satisfies $E[S\mid D,X]=\eta_0(D,X)$. Consequently,*

$$S = \theta_0 D + g_0(X) + \eta, \qquad E[\eta \mid D, X] = 0$$

*and population FWL recovers $\theta_0$.*

*Proof.* Conditional on $(D,X)$, each discrete outcome selects one disjoint
CDF jump with probability equal to that jump's width. Uniform randomization
inside the selected jump therefore makes $V\mid D,X$ uniform on $(0,1)$.
The reference quantile has conditional mean zero, leaving $\eta_0(D,X)$.
$\square$

For cumulative links, this is the conditioning identity behind the
surrogate-variable results of @liu2018 and @cheng2021. The estimation step is
to use $S$, rather than only its diagnostic residual, as the outcome in FWL
or DML.

With an estimated distribution, the equality is asymptotic under consistency
and approximate otherwise. The imputation model is not protected by DML
orthogonality.

The following base R example compares the fitted logistic coefficient, FWL
on the unobserved utility, and FWL on one surrogate completion.

```{r}
#| label: surrogate-fwl-demo
#| echo: true
set.seed(1)
n <- 40000
theta <- 1.5
X <- rnorm(n)
D <- rbinom(n, 1, plogis(0.8 * X))
Z <- theta * D + X + rlogis(n)
Y <- as.integer(Z > 0)

imputation_fit <- glm(Y ~ D + X, family = binomial())
index_hat <- predict(imputation_fit, type = "link")
cdf_at_zero <- plogis(-index_hat)

lower <- ifelse(Y == 1, cdf_at_zero, 0)
upper <- ifelse(Y == 1, 1, cdf_at_zero)
uniform_draw <- runif(n, min = lower, max = upper)
uniform_draw <- pmin(pmax(uniform_draw, 1e-9), 1 - 1e-9)
surrogate <- index_hat + qlogis(uniform_draw)

treatment_residual <- resid(lm(D ~ X))
fwl_coefficient <- function(outcome) {
  sum(treatment_residual * resid(lm(outcome ~ X))) /
    sum(treatment_residual^2)
}
round(c(
  glm = unname(coef(imputation_fit)["D"]),
  oracle = fwl_coefficient(Z),
  surrogate = fwl_coefficient(surrogate)
), 3)
```

All three estimates are close to $\theta_0 = 1.5$. Using only the binary
labels, the surrogate completion recovers the coefficient that FWL would
estimate if the latent utility were observed. Replacing the binary index
with an ordinal cumulative-link index and truncating to category-specific
intervals gives the ordinal construction. The remainder of the paper adds
two elements to this basic estimator: machine-learning partialling and
inference that accounts for the uncertainty hidden by a single completion.

# Estimation and inference

Cross-fit the two adjustment nuisances
$\hat m(X) \approx E[D \mid X]$ and $\hat\ell(X) \approx E[S \mid X]$, and
form the residual-on-residual estimate

$$\hat\theta = \frac{\sum_i (S_i - \hat\ell(X_i))(D_i - \hat m(X_i))}
{\sum_i (D_i - \hat m(X_i))^2}$$ {#eq-theta}

The $\hat\ell$ regression uses $X$ only: the index carried $D$'s signal
into $S$, and partialling on $X$ alone isolates $\theta_0$. The score
$\varphi = (S - \ell(X) - \theta(D - m(X)))(D - m(X))$ is Neyman-orthogonal in
$(\ell, m)$: its derivatives in both nuisance directions vanish by the
tower property, so first-order error in $\hat\ell$ or $\hat m$ does not bias
$\hat\theta$. This protection covers the two adjustment nuisances, not the
imputation model that produced the surrogate.

One completion understates uncertainty. We therefore draw $B$ surrogates and
pool them with Rubin's rules
[@rubin1987]: the estimate is $\bar\theta = B^{-1}\sum_b \hat\theta^{(b)}$
and the total variance is $T = \bar W + (1 + 1/B) B_{\!*}$, with a
Barnard-Rubin small-sample $t$ reference [@barnard1999]. Reusing one fitted
$\hat\beta$ across all completions leaves its shared estimation error
invisible to the between-draw variance.
A proper draw also redraws the imputation model's parameters,
$\beta^{(b)} \sim N(\hat\beta, \widehat{\mathrm{Cov}}(\hat\beta))$
(thresholds included for ordinal outcomes), before each completion. This is
the large-sample normal form of a proper posterior-predictive draw. It
accounts for sampling noise, surrogate randomness, and imputation-model
uncertainty. @tbl-ladder shows why all three matter.

For ordinal outcomes, thresholds are parameters too and must be perturbed
jointly with the coefficients. Surrogate-residual mean and variance checks
provide a diagnostic for the assumed link and mean structure [@liu2018], but
they do not replace target-level validation.

## What completion contributes

Let $R_D=D-m_0(X)$ and let a fitted distribution learner return the pointwise
index $\hat\eta(D,X)$ without requiring a treatment coefficient. Write
$\mu_Y=E\{H^{-1}(V)\mid Y,D,X,\hat G\}$ for the conditional mean reference
correction. The randomized estimator converges to the same completion mean
as its Rao-Blackwellized form.

**Proposition 2 (index projection plus correction).** *With the population
treatment nuisance, the population completion target is*
$$
\theta_{\mathrm{SUDO}}
=\frac{E\{R_D\hat\eta(D,X)\}}{E(R_D^2)}
 +\frac{E(R_D\mu_Y)}{E(R_D^2)}.
$$ {#eq-coefficient-correction}

*Proof.* Substitute $S=\hat\eta(D,X)+\mu_Y$ into the population FWL
numerator and separate its two terms. If
$\hat\eta(D,X)=\hat\alpha D+\hat f(X)$, then
$E\{R_D\hat f(X)\}=0$ and $E(R_D D)=E(R_D^2)$, so the first term reduces to
$\hat\alpha$. $\square$

The first term projects the fitted index onto residualized treatment; the
second uses the observed outcome to correct that projection. For an additive
fit, the first term is its treatment coefficient. The correction can offset
treatment shrinkage, transmit covariate-direction error, or vanish. Under a
binary logit model, it is an entropy-weighted residual moment.

## The unprotected imputation model

Orthogonality protects $\hat\theta$ from first-order error in the adjustment
nuisances. It does not protect the imputation model: its error becomes part
of the completed outcome supplied to DML.

Here, a *black-box imputation model* means a flexible
distributional learner that supplies a pointwise scalar index and conditional
CDF. It need not parameterize or expose a treatment coefficient. The count
validation isolates $D$ from the joint $X$ interaction block. An arbitrary
probability vector without a scalar index on the target scale remains outside
scope. Flexible imputation models are cross-fit so each supplied index is an
out-of-fold prediction.

Conditional on the fitted model, $\hat\eta$ is fixed while
$Y \mid D, X$ is generated by the truth. Since
$S$ is $\hat\eta$ plus truncated noise,

$$E[S \mid D, X] = \psi(\hat\eta, \eta), \quad
\psi(v, e) = v + F(e)\,\mu_+(v) + (1 - F(e))\,\mu_-(v)$$ {#eq-ess}

with $\mu_\pm$ the truncated-noise mean corrections. The correctly centered
draw is exact, $\psi(e, e) = e$, the correct-model surrogate identity in
@liu2018 and @cheng2021,
but index error passes through with a factor strictly below one,

$$c(e) = \left.\frac{\partial \psi}{\partial v}\right|_{v=e}
= 1 - f(-e)\,\bigl(\mu_+(e) - \mu_-(e)\bigr), \qquad f = F'$$ {#eq-ce}

Read $c$ as the fraction of index error that survives the truncation. With
no truncation the surrogate would be index plus noise and its mean would move
one-for-one with $\hat\eta$, giving $c = 1$; the observed category drags it
back, and $c$ is what is left.

Under a logistic link, $c$ rises from $1-\log 2=0.31$ at $e=0$ toward one
in the tails (@fig-passthrough). In fact,
$$c(e) = 1 - H\bigl(F(e)\bigr), \qquad H(p) = -p\log p - (1-p)\log(1-p),$$ {#eq-centropy}
so self-correction equals outcome entropy. An uncertain label sharply
constrains the latent utility and corrects more of the fitted-index error. A
nearly deterministic label adds little information and corrects almost none.

```{r}
#| label: fig-passthrough
#| echo: false
#| fig-width: 5
#| fig-height: 2.6
#| fig-cap: "Fraction of fitted-index error that survives completion under a logistic link. An informative label supplies the strongest correction near the center; the correction fades as the label becomes nearly deterministic."
mu_plus <- function(index) {
  integrate(function(error) error * dlogis(error), -index, Inf)$value /
    plogis(index)
}
mu_minus <- function(index) {
  integrate(function(error) error * dlogis(error), -Inf, -index)$value /
    plogis(-index)
}
pass_through <- function(index) {
  1 - dlogis(index) * (mu_plus(index) - mu_minus(index))
}
index <- seq(0, 6, by = 0.05)
pass_through_value <- vapply(index, pass_through, numeric(1))
par(mar = c(4, 4, 1, 1), cex = 0.9)
plot(
  index, pass_through_value,
  type = "l", lwd = 2, ylim = c(0, 1), las = 1,
  xlab = "index level e",
  ylab = expression("pass-through factor " * c(e))
)
abline(h = 1, lty = 3, col = "grey60")
origin_value <- pass_through(0)
points(0, origin_value, pch = 19)
text(
  0.1, origin_value + 0.1, sprintf("%.2f", origin_value),
  pos = 4, cex = 0.9
)
```

Fitted-index error therefore enters the SUDO target at first order. If the
effective treatment component is displaced by $\delta_D$, the leading bias is
approximately $\bar c_1\delta_D$, where $\bar c_1$ averages $c(e)$ over
the treatment-relevant index distribution. Error in the covariate component
can also survive because completion is nonlinear in the fitted index.
Partialling-out does not erase either channel. Predictive fit alone is not a
sufficient validation criterion; the treatment target must be checked
directly.

# Flexible distribution learners

A parametric model supplies an index and covariance matrix, making proper
draws straightforward. A flexible learner may supply neither a treatment
coefficient nor a usable joint covariance. SUDO asks it only for pointwise
index and CDF predictions, then handles structure and inference separately.

For the validated binary experiments, we restrict the imputation index to
$\hat\alpha D+\hat f(X)$. A GAM, MARS bases built from $X$ only, and
neural-network backfitting all implement this structure. The restriction
prevents treatment-by-covariate interactions, but it does not guarantee a
good fit or a small target bias. Two other partially linear learners failed
the bias criterion, including one tuned by predictive risk.

When a learner has no usable covariance, a full-pipeline bootstrap
resamples the observations and reruns the entire procedure, including the
imputation learner, folds, completions, and adjustment nuisances. A
within-fold refit misses important sources of variation. In the validated
binary designs, normal intervals based on the full-pipeline bootstrap
standard deviation attain nominal coverage.

The resulting workflow is deliberately strict: impose a scalar-index
structure suited to the estimand, tune on target-level simulation bias, and
check coverage before using a learner for inference.

## Scale-calibrated SUDO

Regularization can preserve the shape of an index while compressing its
location and scale. Scalar recalibration corrects that error before
completion. Let a learner supply an out-of-fold logit index
$\hat\eta(D,X)$. On the training observations for a held-out fold, fit
$$
P(Y=1\mid\hat\eta)=\Lambda(a+b\hat\eta)
$$ {#eq-scale-calibration}
and apply $\tilde\eta=a+b\hat\eta$ to the held-out predictions. The estimator
does not read a treatment coefficient from the learner, and neither treatment
nor its residual enters the calibration regression.

For binary logit, write $p=\Lambda(\tilde\eta)$ and
$h(p)=-p\log p-(1-p)\log(1-p)$. The Rao-Blackwell surrogate is
$$
\tilde S
=\tilde\eta+
  \frac{h(p)\{Y-p\}}{p(1-p)}.
$$ {#eq-rb-binary}
It equals the conditional mean of a truncated-logistic surrogate and removes
completion Monte Carlo noise. Randomized completion remains available by
drawing the logistic reference quantile inside the fitted Bernoulli CDF jump.

**Algorithm 1: Cross-fitted scale-calibrated SUDO.**

1. Split the data into $K$ folds and fix the tuning rule.
2. For each held-out fold, fit the distribution learner and treatment
   nuisance $m(X)=E(D\mid X)$ on the other folds. Predict the scalar index in
   the training and held-out observations.
3. Fit @eq-scale-calibration on the training predictions and recalibrate the
   held-out index.
4. Use @eq-rb-binary to construct training and held-out surrogates. Fit
   $\ell(X)=E(\tilde S\mid X)$ on the training observations and predict it in
   the held-out fold.
5. Stack held-out predictions and compute @eq-theta with $S=\tilde S$. For
   randomized completion, repeat the completion and partialling steps, then
   average the treatment estimates.
6. With a flexible learner, bootstrap the observations and rerun all five
   steps. Form a normal interval from the bootstrap standard deviation,
   $\hat\theta\mathbin{\pm}1.96\widehat{\operatorname{sd}}_{\!*}$.

The reason for the calibration is exact in a useful special case. If, on a
new population, $\eta_0(D,X)=a_0+b_0\hat\eta(D,X)$ and scalar recalibration
consistently estimates $(a_0,b_0)$, then @eq-rb-binary gives
$E(\tilde S\mid D,X)=\eta_0(D,X)$. Population FWL therefore recovers
$\theta_0$. With a remaining error $r(D,X)$, bias depends on its projection
onto $D-m_0(X)$ after completion. Scalar calibration does not control an
arbitrary $r$ and is not a substitute for target-level validation.

For sensitivity analysis, an affine fluctuation can add a treatment-residual
direction to @eq-scale-calibration. That version partly targets the treatment
effect before completion. Scale-only calibration more cleanly isolates what
the surrogate and FWL steps contribute.

# Simulation evidence {#sec-sims}

The simulations draw independent standard-normal covariates, assign treatment
from those covariates, and coarsen a partially linear latent utility:

$$X_1, \dots, X_p \stackrel{\text{iid}}{\sim} N(0, 1), \qquad
D \mid X \sim \mathrm{Bernoulli}\bigl(\Lambda(\pi(X))\bigr), \qquad
Z = \theta_0 D + g_0(X) + \varepsilon,$$ {#eq-dgp}
$$\varepsilon \sim F, \qquad
Y = j \iff \kappa_{j-1} < Z \le \kappa_j,$$

One correctly specified covariate isolates the inference calculation. The
five-covariate machine-learning design uses
$$\pi(X) = X_4 + \cos X_5, \qquad g_0(X) = X_1^2 + \sin X_2 + 0.5 X_3,$$
with binary or three-category ordinal outcomes. @tbl-dgp lists the full
designs.

Coverage is evaluated against 0.95. Standard deviation (SD) measures
run-to-run variation; mean standard error (mean SE) averages the reported
uncertainty. Every table is generated from a committed result file.

```{r}
#| label: results-setup
#| include: false
results_dir <- "../../R/results"
read_result <- function(filename) {
  read.csv(file.path(results_dir, filename), check.names = FALSE)
}
sg <- function(x, d = 3) sprintf(paste0("%+.", d, "f"), x)
nn <- function(x, d = 3) sprintf(paste0("%.", d, "f"), x)
```

**Proper draws deliver coverage.** In the simple binary design, the naive
interval covers 0.80 and repeated completion with fixed imputation parameters
covers 0.92. Redrawing those parameters raises coverage to 0.95
(@tbl-ladder).

```{r}
#| label: tbl-ladder
#| echo: false
#| tbl-cap: "Inference schemes, binary outcome, $n = 5000$, $\\theta_0 = 1.5$, 500 replications. $B$ is the number of surrogate draws."
s2 <- read_result("stage2_summary.csv")
estimators <- c("naive", "improper_B25", "proper_B25")
result <- s2[match(estimators, s2$estimator), ]
table_data <- data.frame(
  Scheme     = c("Naive single draw", "Improper, $B=25$", "Proper, $B=25$"),
  Bias       = sg(result$bias),
  SD         = nn(result$sd),
  `Mean SE`  = nn(result$mean_se),
  Coverage   = nn(result$coverage),
  check.names = FALSE)
knitr::kable(table_data, align = c("l", "r", "r", "r", "r"))
```

**Ordinal outcomes.** With three categories and machine-learning adjustment,
SUDO has bias 0.016 and coverage 0.960 in 200 replications
(@tbl-ordinal). Refitting $E[S\mid X]$ for each completion is necessary;
reusing one fit inflated the within-draw variance.

```{r}
#| label: tbl-ordinal
#| echo: false
#| tbl-cap: "Ordinal outcome, $J = 3$, $\\theta_0 = 1.0$. Parametric arm 500 replications; corrected spline + ML arm 200, with the outcome nuisance refit per completion."
s4 <- read_result("stage4_summary.csv")
s5 <- read_result("stage5r_summary.csv")
parametric_result <- s4[s4$estimator == "sudo_ordinal_clm", ]
refit_result <- s5[s5$estimator == "ordinal_refit_n2000_B25", ]
table_data <- data.frame(
  Configuration = c("Parametric, $n=3000$", "Spline + ML nuisances, $n=2000$"),
  Bias      = sg(c(parametric_result$bias, refit_result$bias)),
  SD        = nn(c(parametric_result$sd, refit_result$sd)),
  `Mean SE` = nn(c(parametric_result$mean_se, refit_result$mean_se)),
  Coverage  = nn(c(parametric_result$coverage, refit_result$coverage)),
  check.names = FALSE)
knitr::kable(table_data, align = c("l", "r", "r", "r", "r"))
```

**Direct and tailored comparisons.** In the correctly specified ordinal
design, SUDO and the cumulative-link coefficient agree. Under a cloglog
truth, SUDO, the tailored orthogonal score, and the direct cloglog coefficient
have indistinguishable bias. In the high-dimensional binary design, SUDO
reduces treatment-regularization bias but does not remove it
(@tbl-no-superiority). These comparisons establish agreement with correct
outcome-specific estimators, not an efficiency advantage.

```{r}
#| label: tbl-no-superiority
#| echo: false
#| tbl-cap: "Direct and tailored-score comparisons. The table reports the validated common-sample cells most relevant to the scope claim."
s8 <- read_result("stage8_summary.csv")
s11 <- read_result("stage11_debiased_comparison.csv")
s12 <- read_result("stage12_link_flexibility.csv")
o <- s8[s8$design == "D1_correct" &
          s8$estimator %in% c("clm_direct", "sudo"), ]
h <- s11[s11$n == 2000 & s11$arm == "pen" &
           s11$estimator %in% c("PL_OS", "sudo"), ]
l <- s12[s12$estimator %in% c("sudo_cloglog", "plos_cloglog",
                              "glm_cloglog"), ]
table_data <- rbind(
  data.frame(Design = ifelse(o$design == "D1_correct",
                             "Ordinal, correct", "Ordinal, misspecified"),
             Estimator = ifelse(o$estimator == "clm_direct", "Direct", "SUDO"),
             Bias = sg(o$bias), SD = nn(o$sd)),
  data.frame(Design = paste0("Binary high-dimensional, ", h$arm),
             Estimator = ifelse(h$estimator == "PL_OS",
                                "Tailored orthogonal", "SUDO"),
             Bias = sg(h$bias), SD = nn(h$sd)),
  data.frame(Design = "Binary cloglog, correct link",
             Estimator = c("Direct", "Tailored orthogonal", "SUDO")[
               match(l$estimator, c("glm_cloglog", "plos_cloglog",
                                    "sudo_cloglog"))],
             Bias = sg(l$bias), SD = nn(l$sd))
)
knitr::kable(table_data, align = c("l", "l", "r", "r"))
```

**Link misspecification.** Assuming a logistic link under Gumbel noise biases
the latent-scale effect by +0.77 for binary outcomes and +0.58 for ordinal
outcomes, with zero coverage. Bias is below 0.005 under the correct link.
Conditional second-moment drift in the surrogate residual flags the poor fit,
but is not specific to the link and remains a diagnostic [@liu2018].

**Partially-linear imputation learners.** GAM, MARS, and neural backfitting
fit the same $\alpha D+f(X)$ structure. All three pass the prespecified
bias criterion, and full-pipeline-bootstrap normal coverage ranges from 0.94
to 0.98 (@tbl-pl-learners). With 50 replications, these differences are too
small to rank the learners.

```{r}
#| label: tbl-pl-learners
#| echo: false
#| tbl-cap: "Partially-linear imputation learners with full-pipeline bootstrap, $n=2000$, 50 replications, 50 outer resamples, and 15 surrogate completions per resample."
s3s <- read_result("stage3s_summary.csv")
learner_name <- c(
  pl_gam = "GAM",
  pl_mars = "MARS",
  pl_backfit = "Neural backfit"
)
table_data <- data.frame(
  Learner = unname(learner_name[s3s$learner]),
  `$\\theta_0$` = nn(s3s$theta, 1),
  Bias = sg(s3s$bias),
  `SD/Mean SE` = nn(s3s$sd_over_mean_se),
  `Normal coverage` = nn(s3s$coverage_normal),
  `Percentile coverage` = nn(s3s$coverage_percentile),
  check.names = FALSE
)
knitr::kable(table_data, align = c("l", "r", "r", "r", "r", "r"))
```

Partial linearity alone is not enough. Two other partially linear learners
fail the bias screen, including one selected by cross-validated predictive
risk. The successful configurations were screened on target-level Monte
Carlo bias and evaluated on separate seeds. A data-only selection rule remains
open.

**Coefficientless additive learner.** GAMSEL [@chouldechova2015] fits the
binary index with a linear treatment term while allowing each $X_j$ to enter
linearly, smoothly, or not at all. SUDO uses only its index predictions. In
three $p=80$ additive designs, full-affine calibrated SUDO passes every
prespecified bias, coverage, SD-to-SE, and bootstrap-success gate
(@tbl-calibrated-sudo). Ordinary SUDO retains material regularization bias.

```{r}
#| label: tbl-calibrated-sudo
#| echo: false
#| tbl-cap: "Coefficientless GAMSEL learner with treatment coefficient 0.75. Full-affine results use 50 fresh datasets and 99 full-pipeline bootstrap resamples."
s16g <- read_result("stage16g_affine_calibrated_sudo_coverage.csv")
s16h <- read_result("stage16h_calibration_ablation.csv")
cell_order <- c("randomized", "confounded", "high_signal")
full <- s16g[
  s16g$estimator == "affine_calibrated_sudo",
  c("cell", "bias", "coverage", "sd_to_se")
]
ordinary <- s16g[
  s16g$estimator == "sudo", c("cell", "bias")
]
full <- full[match(cell_order, full$cell), ]
ordinary <- ordinary[match(cell_order, ordinary$cell), ]
table_data <- data.frame(
  Design = c("Randomized", "Confounded", "High signal"),
  `Ordinary bias` = sg(ordinary$bias),
  `Calibrated bias` = sg(full$bias),
  Coverage = nn(full$coverage),
  `SD/Mean SE` = nn(full$sd_to_se),
  check.names = FALSE
)
knitr::kable(table_data, align = c("l", rep("r", 4)))
```

```{r}
#| label: tbl-scale-ablation
#| echo: false
#| tbl-cap: "Scale-only mechanism screen on separate seeds. Entries are bias over 10 fresh datasets."
scale_direct <- s16h[
  s16h$estimator == "pit_affine_scale", c("cell", "bias")
]
scale_sudo <- s16h[
  s16h$estimator == "scale_calibrated_sudo", c("cell", "bias")
]
scale_direct <- scale_direct[match(cell_order, scale_direct$cell), ]
scale_sudo <- scale_sudo[match(cell_order, scale_sudo$cell), ]
table_data <- data.frame(
  Design = c("Randomized", "Confounded", "High signal"),
  `Direct read-off` = sg(scale_direct$bias),
  `Scale-calibrated SUDO` = sg(scale_sudo$bias),
  check.names = FALSE
)
knitr::kable(table_data, align = c("l", "r", "r"))
```

The scale-only ablation separates calibration from treatment targeting. Its
biases are $+0.014$, $+0.009$, and $-0.009$. Under confounding, the coefficient
read directly from the recalibrated index has bias $+0.039$; completion and
FWL reduce it to $+0.009$ (@tbl-scale-ablation). Tilt-only calibration fails
under confounding and high signal. Scale-only coverage has not been tested,
so the two tables deliberately separate its point-bias result from the
full-affine coverage result. At $p=500$, even full-affine calibrated SUDO has
bias $-0.048$ and fails the point gate.

With continuous treatment, bias remains below 1.5% of the truth in the binary
and ordinal designs. Strong-signal coverage is less stable, reinforcing the
need to rerun the full-pipeline bootstrap rather than transfer an uncertainty
result from a binary-treatment experiment.

## Count oracle bridge

To separate count completion from fitted-learner error, we supply the true
scalar index and count CDF. Six prespecified cells cross Poisson and
negative-binomial laws, binary and continuous treatments, and two signal
strengths where applicable. At
$n=2000$, with 100 replications, $B=25$ randomized completions, five folds,
and 99 bootstrap resamples, both analytic Rao-Blackwell and randomized
completion pass every bias, coverage, and SD-to-SE gate
(@tbl-count-oracle). Their paired differences are within two completion
Monte Carlo standard errors in every cell.

```{r}
#| label: tbl-count-oracle
#| echo: false
#| tbl-cap: "Count oracle bridge with the true index and CDF supplied. Entries pair Rao-Blackwell and randomized completion. Each cell uses $n=2000$, 100 replications, five folds, and 99 bootstrap resamples; randomized SUDO uses $B=25$."
s15o <- read_result("stage15o_count_oracle.csv")
cell_order <- c(
  "continuous_poisson_tlog1p5", "continuous_negbin_tlog1p5",
  "continuous_poisson_tlog2", "continuous_negbin_tlog2",
  "binary_poisson_tlog2", "binary_negbin_tlog2"
)
estimator_order <- c("rao_blackwell", "randomized")
s15o$cell <- factor(s15o$cell, levels = cell_order)
s15o$estimator <- factor(s15o$estimator, levels = estimator_order)
result <- s15o[order(s15o$cell, s15o$estimator), ]
rao_blackwell <- result[result$estimator == "rao_blackwell", ]
randomized <- result[result$estimator == "randomized", ]
stopifnot(identical(as.character(rao_blackwell$cell),
                    as.character(randomized$cell)))
table_data <- data.frame(
  Design = paste(
    ifelse(rao_blackwell$treatment == "continuous", "Continuous", "Binary"),
    ifelse(rao_blackwell$family == "poisson", "Poisson", "NB"),
    sep = ", "
  ),
  `$\\theta_0$` = ifelse(
    abs(rao_blackwell$truth - log(1.5)) < 1e-8,
    "$\\log(1.5)$", "$\\log(2)$"
  ),
  `Bias (RB / randomized)` = paste(
    sg(rao_blackwell$bias), sg(randomized$bias), sep = " / "
  ),
  `SD/SE (RB / randomized)` = paste(
    nn(rao_blackwell$sd / rao_blackwell$mean_se),
    nn(randomized$sd / randomized$mean_se),
    sep = " / "
  ),
  `Coverage (RB / randomized)` = paste(
    nn(rao_blackwell$coverage), nn(randomized$coverage), sep = " / "
  ),
  check.names = FALSE
)
knitr::kable(table_data, align = c("l", "c", "r", "r", "r"))
```

The oracle results validate count completion and bootstrap inference under a
correct supplied distribution. They do not validate estimation of that
distribution. Constrained XGBoost failed the point-bias screen, so the paper
makes no claim for fitted count learners or the planned bikeshare analysis.

# Wine quality application

We apply ordinal SUDO to the UCI wine-quality data [@cortez2009]. The outcome
is the median ordinal score from blind tasters. We analyze 1,599 red and 4,898
white wines separately; scores range from 3 to 8 for red wine and 3 to 9 for
white. The exposure is standardized volatile acidity, an acetic-acid fault
associated with vinegar-like aroma. The other ten standardized
physicochemical measurements form the adjustment set.

This is an observational analysis. A causal interpretation would require the
measured chemistry to suffice for adjustment and winemaking practice to
affect quality only through that chemistry and volatile acidity
(@fig-dag). Those assumptions are not established, so we report conditional
associations.

```{mermaid}
%%| label: fig-dag
%%| fig-width: 5.2
%%| fig-pos: H
%%| fig-cap: "Assumed application DAG. Volatile acidity (the treatment) affects latent quality directly; unobserved winemaking practice confounds both through the measured chemistry, the adjustment set."
flowchart LR
  P((winemaking<br/>practice)):::u --> VA[volatile acidity]
  P --> X[other chemistry]
  X --> VA
  VA --> Q[latent quality]
  X --> Q
  classDef u fill:#eee,stroke-dasharray:4 3;
```

\FloatBarrier

Volatile acidity retains 75% of its marginal standard deviation after
adjustment in red wine and 94% in white wine. Both imputation models are
proportional-odds logits with volatile acidity entering linearly. One uses
the adjustment variables linearly; the other uses a three-degree-of-freedom
natural spline for each variable, with no treatment-by-covariate interactions.
Ordinal SUDO uses 50 proper draws, five-fold cross-fitted GAM nuisances, and
Rubin pooling. The two specifications give similar estimates (@tbl-wine).

The coefficients correspond to one standard deviation of volatile acidity,
about 0.18 g/dm$^3$ for red wine and 0.10 for white. A one-SD increase is
associated with odds ratios of about 0.59 and 0.64, respectively, for scoring
above any fixed quality level. The direct cumulative-link coefficient is a
model-based comparator.

```{r}
#| label: tbl-wine
#| echo: false
#| tbl-cap: "Effect of a one-SD increase in volatile acidity on latent wine quality. Model-based coefficient from the linear cumulative-link fit (standard error in parentheses); SUDO estimates with 95% intervals in brackets."
wine_results <- read_result("wine_application.csv")
wine_row <- function(wine_name, label) {
  linear <- wine_results[
    wine_results$wine == wine_name & wine_results$full_model == "linear",
  ]
  spline <- wine_results[
    wine_results$wine == wine_name & wine_results$full_model == "pl_spline",
  ]
  data.frame(
    Wine = label,
    `Overlap $R^2(D \\sim X)$` = nn(linear$overlap_r2, 2),
    `Cumulative-link` = sprintf(
      "$%.2f$ (%.2f)", linear$naive_theta, linear$naive_se
    ),
    `SUDO (linear)` = sprintf(
      "$%.2f$ [$%.2f,%.2f$]",
      linear$sudo_theta, linear$ci_lo, linear$ci_hi
    ),
    `SUDO (spline)` = sprintf(
      "$%.2f$ [$%.2f,%.2f$]",
      spline$sudo_theta, spline$ci_lo, spline$ci_hi
    ),
    check.names = FALSE)
}
table_data <- rbind(wine_row("red", "Red"), wine_row("white", "White"))
knitr::kable(table_data, align = c("l", "r", "r", "r", "r"))
```

The negative sign is stable when any one adjustment covariate is removed.
An omitted-variable calibration suggests that a confounder would need to
explain roughly 15% of the residual treatment and outcome variation for red
wine and 10% for white wine to move the estimate to zero [@cinelli2020].
Because that calibration was derived for linear regression with an observed
outcome, we report it as a heuristic rather than a formal SUDO bound.

The fitted pass-through averages 0.21 for red wine and 0.17 for white. Thus
the observed categories correct much of a local fitted-index error in these
data. This calculation does not establish robustness to arbitrary
misspecification.

# Discussion

SUDO separates two tasks that are often conflated. The fitted distribution
places the outcome on the target scale, and DML removes flexible covariate
effects on that scale. This gives binary and ordinal models the same
partialling-out interface. It does not supersede a correct direct estimator or
a tailored orthogonal score. Proper draws or a full-pipeline bootstrap are
part of the procedure because a single completion hides uncertainty.

The separation also identifies the weak point. Orthogonality covers the two
adjustment regressions, not the distribution that creates the surrogate.
Partial linearity rules out treatment-by-covariate interactions but does not
ensure a small target bias. Scalar recalibration repairs one important case,
namely an index with the right shape and the wrong location or scale. It does
not repair arbitrary index error, as the $p=500$ failure shows.

Current evidence supports binary and ordinal SUDO, including a
coefficientless calibrated binary learner. Count completion works with an
oracle index and CDF, but fitted count learning remains unresolved. The next
inferential check is full coverage validation of scale-only calibration. More
broadly, every new imputation learner should be evaluated through
target-level bias and coverage experiments.

# Data and software

The wine-quality data come from the University of California, Irvine Machine
Learning Repository [@cortez2009]. The `sudo` Python package provides the
parametric estimator for binary and ordinal outcomes, including proper draws
and the per-completion ordinal outcome-nuisance refit. The partially-linear
machine-learning imputation learners and the full-pipeline bootstrap are
implemented in R, which is also where every method is prototyped and
validated before release. Count and flexible ordinal adapters remain R-only
until their fitted-learner confirmation stages pass. The full-affine
calibrated estimator also remains R-only, and scale-only calibration still
awaits coverage validation. The open-source
repository contains the code and committed result files behind every table
and figure under the MIT License:
<https://github.com/bgreenwell/sudo>.

# Declaration of artificial intelligence use

We used generative artificial intelligence (AI) to assist with code and
summaries of reference materials. The authors reviewed and validated all
outputs, retain full ownership of the final manuscript, and accept sole
responsibility for any errors.

# Appendix: simulation and application details {#sec-appendix}

Each design draws a latent utility
$Z = \theta_0 D + g_0(X) + \varepsilon$ and converts it to an observed
category. Most designs use logistic $\varepsilon$; the exceptions are
specified below. The designs vary the covariate structure, cutpoints, and
fitted models; covariates remain standard normal throughout. The `R/` and
`R/results/` directories contain the full data-generating code and committed
per-cell results.

## Data-generating processes

All designs instantiate @eq-dgp with one of two covariate structures:

- **S1**, one covariate: $p = 1$, $\pi(X) = 0.8X$, $g_0(X) = X$. Every model
  in play is correctly specified, so only the pooling is under test.
- **S5**, five covariates: $p = 5$, $\pi(X) = X_4 + \cos X_5$,
  $g_0(X) = X_1^2 + \sin X_2 + 0.5 X_3$. The outcome structure is nonlinear
  in $X$, and treatment depends on covariates absent from $g_0$.

| Design | Structure | $F$ | $\kappa$ | $\theta_0$ | $n$ | Replications |
|---|---|---|---|---|---|---|
| Inference ladder (@tbl-ladder) | S1 | logistic | $(0)$ | 1.5 | 5000 | 500 |
| Ordinal parametric (@tbl-ordinal, top) | S1 | logistic | $(-1, 1)$ | 1.0 | 3000 | 500 |
| Binary ML | S5 | logistic | $(0)$ | 0.5, 3 | 1000, 2000 | 200 |
| Ordinal flexible adjustment (@tbl-ordinal, bottom) | S5 | logistic | $(0, 2)$ | 1.0 | 1000, 2000 | 200 |
| Link misspecification | S1 | Gumbel | $(0)$ or $(0, 2)$ | 1.0 | 5000 | 200 |
| Black-box imputation | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 100 |
| PL learner comparison (@tbl-pl-learners) | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 50 |
| Calibration (@tbl-calibrated-sudo) | $p=80$ additive | logistic | $(0)$ | 0.75 | 3000 | 50 |
| DGP variation, continuous $D$ | S5, $D = 0.7\pi(X) + N(0,1)$ | logistic | $(0)$ or $(0, 2)$ | 1, 2.5 | 2000 | 200 |

: Data-generating processes. $\kappa$ lists the interior cutpoints, so
$(0)$ is binary and a pair gives $J = 3$. {#tbl-dgp}

**Models fitted.** The ladder uses $\mathrm{glm}(Y \sim D + X)$ as the
imputation model with $B \in \{10, 25, 50\}$ draws (the table shows
$B = 25$). The ordinal parametric design uses `ordinal::clm`, whose
covariance covers the thresholds, with $B = 25$. The ordinal
flexible-adjustment design uses parametric `clm` on $D$ and natural-spline
bases $\mathrm{ns}(X_k, 4)$ with `mgcv::gam`
partialling nuisances and $B = 25$.

**Link misspecification.** This design retains the one-covariate structure
but replaces logistic $\varepsilon$ with an extreme-value error: Gumbel-max
for the binary case, matching `glm` with a cloglog link, and Gumbel-min for
the ordinal case, matching `ordinal::clm` with a cloglog link. The analyst
fits either logit (misspecified) or cloglog (correct). Holding the simple
covariate structure fixed isolates error in the link.

**Partially-linear imputation models.** The original black-box result uses
$\hat\eta = \hat\alpha D + \hat f(X)$ with $\hat f$ a single-hidden-layer
neural net (`nnet`, `size = 4`, `decay = 0.3`, five backfitting iterations,
selected on target-level Monte Carlo bias) and 100 full-pipeline
bootstrap resamples. The common comparison uses 50 resamples per arm. The
GAM includes a linear treatment term and smooths of $X$. MARS builds
degree-two bases from $X$ only before a joint logistic fit with mandatory
linear $D$.

**Calibration experiment.** The calibration designs use $p=80$ independent
standard-normal covariates and
$$
g_0(X)=0.65X_1+0.55(X_2^2-1)+0.45X_3-0.35(X_4^2-1)
       +0.30X_5-0.25X_6.
$$
Treatment is $N(0,1)$ in the randomized design. In the confounded design,
$D=m_0(X)+N(0,1)$ with
$m_0(X)=0.50X_1-0.40X_3+0.30(X_2^2-1)$. The high-signal design uses
$D\sim N(0,4)$. Outcomes follow
$Y\sim\operatorname{Bernoulli}[\Lambda\{0.75D+g_0(X)\}]$.
GAMSEL uses cubic candidate smooths for $X$, a linear treatment term, and
cross-validated binomial deviance. Five outer and five inner folds are used.
The point ablation uses separate seeds from tuning and coverage.

**Wine application.** Red and white wines are fit separately. The outcome
levels are 3--8 for red wine and 3--9 for white. Volatile acidity is the
standardized treatment. The ten standardized adjustment variables are fixed
acidity, citric acid, residual sugar, chlorides, free and total sulfur
dioxide, density, pH, sulphates, and alcohol. The proportional-odds logit
full model is `ordinal::clm` with index
$\alpha D+\beta^\top X$ (linear) or
$\alpha D+\sum_k \mathrm{ns}(X_k,3)$ (spline). Partialling nuisances
$E[S \mid X]$ and $E[D \mid X]$ use `mgcv::gam` (Gaussian for continuous
$D$), with five-fold cross-fitting, $B=50$ proper draws, and Rubin pooling.

## Nuisances, cross-fitting, and draws

Partialling nuisances $\hat\ell(X) = E[S \mid X]$ and $\hat m(X) = E[D \mid X]$
are `mgcv::gam` throughout, fit with five-fold cross-fitting. Proper draws
sample the full-model parameters from $N(\hat\beta, \widehat{\mathrm{Cov}}(\hat\beta))$
(thresholds included for `clm`) before each completion. Diagnostic draws use
a random number generator (RNG) stream independent of the estimation draws.

Binary experiments found per-completion refitting of $\hat\ell$ negligible,
so the binary estimator reuses a plug-in fit. The ordinal estimator refits
$\hat\ell$ for every completion because reuse inflated its within-draw
variance.

## Monte Carlo precision

Coverage is a binomial proportion, so its Monte Carlo standard error at a
true 0.95 is about $\sqrt{0.95 \cdot 0.05 / R}$: 0.010 at $R = 500$
replications, 0.015 at $R = 200$, and 0.022 at $R = 100$. Per-cell bias and
coverage Monte Carlo standard errors are in the committed result files.

# References

::: {#refs}
:::
