---
title: "Surrogate-Assisted Double Machine Learning: Causal Inference with Binary and Ordinal Outcomes"
author:
  - name: Brandon M. Greenwell
    email: greenwbm@ucmail.uc.edu
    affiliations:
      - name: "University of Cincinnati Carl H. Lindner College of Business"
        department: Department of OBAIS
        address: 2906 Woodside Drive
        city: Cincinnati, OH
        country: USA
        postal-code: 45221
    orcid: 0000-0002-8120-0084
  - name: Other Authors
    email: other.authors@email.com
    affiliations:
      - name: "University of XYZ"
authorcols: true
date: today
abstract: |
  Double Machine Learning (DML) delivers root-$n$ consistent, asymptotically normal causal estimates by pairing Neyman-orthogonal scores with cross-fitting, but its partially linear model is confined to continuous outcomes. Binary outcomes are served only by a logit-hardwired logistic estimator; ordinal outcomes by no DML method we are aware of. We introduce SUDO (Surrogate-Assisted DML), which treats the latent outcome behind a categorical response as missing data and fills it in: each observation is completed to a continuous *surrogate* drawn from the assumed error distribution, truncated to the interval the observed category implies, after which ordinary partialling-out applies. The completions are multiple imputations, and pooling them by Rubin's rules is what restores nominal coverage, but only when the draws are *proper*, redrawing the imputation model's parameters (thresholds included) before each draw; naive and improper draws demonstrably under-cover. The method is link-flexible and the surrogate residuals double as a link diagnostic. We further give a closed-form account of how imputation-model error reaches the target: it passes through a truncation factor that rises from about 0.3 to 1 as the signal strengthens, so the imputation model's error is first-order, not the second-order nuisance error of standard DML. In the black-box designs studied, a partially-linear machine-learning imputation model with a full-pipeline bootstrap attains nominal coverage. We illustrate on wine-quality data, estimating the effect of volatile acidity on an ordinal quality score.
format:
  arxiv-pdf:
    keep-tex: false
bibliography: references.bib
---

# Introduction

Many outcomes that scientists want causal effects for are categorical.
A patient relapses or not; a wine is rated on a five- or ten-point scale;
a survey respondent is dissatisfied, neutral, or satisfied. For a
*continuous* outcome, Double Machine Learning [DML, @chernozhukov2018]
provides a now-standard route to a causal effect under flexible,
high-dimensional adjustment: predict the outcome from the covariates and
the treatment from the covariates with any machine-learning models, remove
both predictions, and read the effect from the residual-on-residual
regression. The removal, a Frisch-Waugh-Lovell (FWL) orthogonalization,
is what lets the nuisance predictions be flexible without contaminating the
effect, because the score is Neyman-orthogonal and, given standard
convergence-rate conditions, is insensitive to first-order nuisance error
[@chernozhukov2018].

That machinery is built for a continuous outcome. A binary or ordinal outcome
obeys the additive structure only on a *latent* scale: there is an
unobserved continuous utility $U = \theta_0 D + g_0(X) + \varepsilon$, and
we see only which interval $U$ fell in. Applying the continuous
partialling-out estimator to the raw labels still runs, but it changes the
estimand: it returns a probability-scale contrast, not $\theta_0$. An
available remedy for the binary case is the logistic
partially linear model of @liu2021lplr, the DML estimator for a binary
outcome with an official implementation in the DoubleML software. It is
effective but narrow: the logit link is built into its orthogonal score, so
it is confined to binary responses under a logistic distribution and has no ordinal
counterpart; for ordinal outcomes we are aware of no DML estimator. Ordinal
outcomes are therefore the sharpest gap, and the one this paper closes.
Under a logistic link, SUDO targets the same latent-scale coefficient as
@liu2021lplr through a different surrogate-completion construction; we do not
claim that the two estimators are identical or asymptotically equivalent.

Independently, @liu2018 and @cheng2021 developed *surrogate residuals* as a
diagnostic device for discrete-choice models. Grounded in random utility
theory [@mcfadden1974], a surrogate maps an observed category to a draw
from the conditional distribution of its latent utility, yielding a continuous,
homoscedastic residual with a known reference distribution
[@greenwell2018]. We repurpose that construction as an *estimation* device.
Substituting a surrogate $S$ for the unobserved $U$ turns the discrete
problem into a continuous one on which the FWL recipe runs unchanged. The
surrogate is a guess, so a single completion is not enough; the completions
are multiple imputations, pooled by Rubin's rules [@rubin1987]. Getting
that pooling right, imputing from the imputation model's *posterior
predictive* rather than a plug-in, is what separates near-nominal coverage
from clear under-coverage in every design we study, and is where a naive
application fails.

**Contributions.**

1. A DML estimator for binary and ordinal outcomes in one framework, not
   tied to the logit link, with a built-in link diagnostic.
2. The finding that only the proper multiple-imputation draw scheme attained
   nominal coverage in the designs we study, quantified against naive and
   improper alternatives.
3. A closed-form analysis of how imputation-model error propagates to the
   effect, revealing a *first-order* sensitivity absent from standard DML,
   confirmed by a direct perturbation experiment.
4. A route to nominal coverage with black-box machine-learning imputation
   models, via a partially-linear structure and a full-pipeline bootstrap.
5. An application to wine quality, an ordinal latent-scale estimand for
   which we are aware of no existing DML estimator.

# Setup and estimand

Let $D$ be a treatment, $X \in \mathbb{R}^p$ covariates, and $Y$ a
categorical outcome: binary, $Y = \mathbf{1}\{U > 0\}$, or ordinal,
$Y = j \iff \kappa_{j-1} < U \le \kappa_j$ for thresholds
$-\infty = \kappa_0 < \kappa_1 < \dots < \kappa_{J-1} < \kappa_J = \infty$
(we write $\kappa$ rather than $c$ to keep the thresholds distinct from the
pass-through factor $c(\cdot)$ of @sec-theory). The latent utility
follows a partially linear model

$$U = \theta_0 D + g_0(X) + \varepsilon, \qquad
\varepsilon \mid D,X \sim F$$ {#eq-latent}
with $g_0$ an arbitrary nuisance function and $F$ a known error distribution
(logistic, giving proportional-odds; extreme-value, giving
complementary-log-log; etc.). The target $\theta_0$ is the treatment effect
on the latent-utility scale, the same quantity a correctly specified
cumulative-link model targets, but with $g_0$ left nonparametric. It is not
the probability-scale average treatment effect; the two answer different
questions and should not be compared against a common truth. Because $F$
fixes the variance of $\varepsilon$, this scale is set by the link rather
than estimated, which is what gives $\theta_0$ its familiar reading: under
the logistic link a one-unit change in $D$ shifts the cumulative log-odds
of $Y$ by $\theta_0$, so $\exp(\theta_0)$ is a proportional-odds ratio.

Four assumptions identify $\theta_0$. (i) *Selection-on-observables*: no
unobserved confounder acts beyond $X$; together with the conditional
location-family statement in @eq-latent, the latent error has the same
normalized law across treatment values given $X$. (ii) *A constant latent-scale effect*: the
partially linear form (@eq-latent) makes $\theta_0$ a single number rather
than a function of $X$. (iii) *A correctly specified latent distribution*:
the surrogate completion draws from $F$ and the assumed cutpoints, so the
conditional law of $\varepsilon$, not merely its mean, must be right; this
is the assumption the link diagnostic monitors, and the one the propagation
analysis shows the estimate is most sensitive to. (iv) *Overlap*: the
treatment keeps support after adjustment, so its residual is not
annihilated. Assumptions (i) and (iv) are the standard causal conditions,
made explicit for the application through a stated causal graph; (ii) and
(iii) are modeling choices shared with any cumulative-link analysis.

# Background: FWL, DML, and surrogate residuals

The engine underneath DML is the Frisch-Waugh-Lovell (FWL) theorem: in a
linear model the coefficient on $D$ equals the coefficient from regressing
the $Y$-residuals on the $D$-residuals after partialling $X$ out of both.
DML makes this practical for high-dimensional $X$ by estimating the two
partialling regressions with flexible machine learning and $K$-fold
cross-fitting, and by using an orthogonal score so that first-order error
in those estimates does not bias the effect [@chernozhukov2018].

FWL still runs on a binary or ordinal label, but it targets the wrong thing:
the residual of a $0/1$ variable is bounded and heteroscedastic, and
regressing it on the treatment residual estimates a probability-scale
contrast rather than the latent-scale $\theta_0$. Recovering $\theta_0$ is
the obstacle SUDO removes.

The removal uses *surrogate residuals* [@liu2018; @cheng2021], developed
originally to diagnose discrete-choice models. Grounded in random utility
theory [@mcfadden1974], the construction maps an observed category to a
draw from the conditional distribution of its latent utility, producing a
continuous residual that is homoscedastic with a known reference
distribution [@greenwell2018]. We use it not to diagnose but to *complete*
the outcome, so that FWL, and then DML, can run.

# Surrogate FWL: the idea in its simplest form

The whole method reduces to one move. Fit an imputation model of the
outcome on $(D, X)$ to obtain an estimated latent index $\hat V_i$; for each
observation draw a surrogate $S_i$ from the assumed distribution $F$, located at $\hat V_i$ and truncated to the interval the observed category implies:
$(0,\infty)$ if $Y_i = 1$, $(-\infty, 0]$ if $Y_i = 0$, or
$(\hat\kappa_{Y_i-1}, \hat\kappa_{Y_i}]$ in the ordinal case. When the imputation model is correct and its parameters known, the
surrogate is drawn from the conditional law of $U$ given $(D, X)$, so
$E[S \mid D, X] = E[U \mid D, X]$ (Theorem 1 of @cheng2021) and the
completed data satisfy

$$S = \theta_0 D + g_0(X) + \eta, \qquad E[\eta \mid D, X] = 0$$ {#eq-completed}

with the same $g_0$ as in @eq-latent, so FWL recovers $\theta_0$. This is
the exact case. With an estimated index $\hat V$ the equality holds only
asymptotically under consistent estimation, and approximately otherwise;
the propagation analysis below makes the residual error precise. The
demonstration that follows fits the index by maximum likelihood, so it sits
in the asymptotic regime.

Nothing more is needed to see the idea work. The self-contained example
below (base R, no packages) draws a binary outcome from a logistic latent
model, fits the imputation model, completes the outcome to a surrogate,
and compares three estimates of $\theta_0$: the logistic-regression
coefficient on the treatment, the FWL estimate on the *true* (unobserved)
utility (the oracle), and the FWL estimate on the *surrogate* completion.

```{r}
#| label: surrogate-fwl-demo
#| echo: true
set.seed(1)
n <- 40000; theta <- 1.5
X <- rnorm(n)
D <- rbinom(n, 1, plogis(0.8 * X))        # treatment depends on a covariate
U <- theta * D + X + rlogis(n)            # latent utility (never observed)
Y <- as.integer(U > 0)                    # only the binary label is seen

V  <- predict(glm(Y ~ D + X, binomial))   # fitted latent index
Flo <- plogis(-V)                         # CDF of logistic(V) at the cutpoint 0
p  <- ifelse(Y == 1, Flo + runif(n) * (1 - Flo), runif(n) * Flo)
S  <- V + qlogis(pmin(pmax(p, 1e-9), 1 - 1e-9))   # truncated-logistic surrogate

rD  <- resid(lm(D ~ X))                   # partial the covariate out of D
fwl <- function(z) sum(rD * resid(lm(z ~ X))) / sum(rD^2)
round(c(glm       = unname(coef(glm(Y ~ D + X, binomial))["D"]),
        oracle    = fwl(U),               # FWL on the true utility
        surrogate = fwl(S)), 3)           # FWL on the surrogate completion
```

All three land on $\theta_0 = 1.5$: the surrogate completion recovers, from
the binary labels alone, the coefficient the latent model would have. The
identical construction, an ordinal cumulative-link index with interval
truncation, works for ordinal outcomes, and the same estimator is provided
as a Python package with a parallel set of R scripts. The rest of the paper does two
things to this simple estimator: it hands the partialling to machine
learning, and it makes the inference honest, since a single completion
understates uncertainty.

# Surrogate-assisted double machine learning

**The estimator.** Cross-fit the two adjustment nuisances
$\hat m(X) \approx E[D \mid X]$ and $\hat\ell(X) \approx E[S \mid X]$, and
form the residual-on-residual estimate

$$\hat\theta = \frac{\sum_i (S_i - \hat\ell(X_i))(D_i - \hat m(X_i))}
{\sum_i (D_i - \hat m(X_i))^2}$$ {#eq-theta}

The $\hat\ell$ regression uses $X$ only: the index carried $D$'s signal
into $S$, and partialling on $X$ alone strips it back out so the residual
regression isolates $\theta_0$. The score
$\varphi = (S - \ell(X) - \theta(D - m(X)))(D - m(X))$ is Neyman-orthogonal in
$(\ell, m)$: its derivatives in both nuisance directions vanish by the
tower property, so first-order error in $\hat\ell$ or $\hat m$ does not bias
$\hat\theta$. This protection covers the two *adjustment*
nuisances but not the imputation model that produced $\hat V$; the
next section analyzes the difference.

**Proper multiple-imputation inference.** A single completion understates
uncertainty, so we draw $B$ surrogates and pool with Rubin's rules
[@rubin1987]: the estimate is $\bar\theta = B^{-1}\sum_b \hat\theta^{(b)}$
and the total variance is $T = \bar W + (1 + 1/B) B_{\!*}$, the mean
within-draw sandwich variance plus the inflated between-draw variance, with
a Barnard-Rubin small-sample $t$ reference [@barnard1999]. What matters is
*how* the draws differ. Each surrogate is centered at an estimated index
$\hat V = x'\hat\beta$; that estimate carries sampling error, and reusing
one fitted $\hat\beta$ across all draws (an *improper* draw) leaves the
shared error invisible to the between-draw variance, so $T$ comes up short.
A *proper* draw redraws the imputation model's parameters,
$\beta^{(b)} \sim N(\hat\beta, \widehat{\mathrm{Cov}}(\hat\beta))$
(thresholds included in the ordinal case), before each completion, the
large-sample normal form of Rubin's posterior-predictive requirement. @tbl-ladder shows the three
schemes moving coverage from 0.80 (naive) to 0.93 (improper) to 0.96-0.97
(proper).

$T$ must account for three things that vary from draw to draw: the sampling
noise in $\hat\theta$ once a completion is fixed (the within-draw $\bar W$),
the surrogate randomness in $S$, and the uncertainty in the imputation model
itself. Improper draws discard the last of these.
The proper draw restores it by sampling
$\beta^{(b)} \sim N(\hat\beta, \widehat{\mathrm{Cov}}(\hat\beta))$, the
asymptotic sampling distribution of $\hat\beta$. Under standard regularity
this normal limit is shared by the parameter's sampling distribution, its
flat-prior posterior, and the parametric-bootstrap distribution, so the
proper draw is the large-sample form of a parametric bootstrap of the
imputation model, read off the covariance rather than recomputed by
refitting. It therefore needs a model that reports a
usable covariance, and that requirement is exactly what breaks for a
black-box learner, taken up in the machine-learning section below.

**Ordinal outcomes and the link.** The ordinal case is identical once a
cumulative-link model supplies the index and thresholds; because the
thresholds are parameters too, the proper draw perturbs them jointly with
the coefficients, which is why we fit with a model whose covariance covers
them. The error distribution $F$ is an assumption, checked by a diagnostic: under a correct link the surrogate residuals $R = S - \hat V$ are homoscedastic, so
a smooth of $R^2$ on $X$ (on an independent draw stream) flags a
misspecified link through variance drift that mean-based checks miss.

# How imputation-model error propagates

The orthogonality established above protects $\hat\theta$ from error in the
adjustment nuisances but says nothing about the imputation model, whose
error is baked into the completed data $S$ rather than differenced away.
This section quantifies that channel. It is the structural difference
between SUDO and standard DML: the imputation model, unlike the two
adjustment nuisances, sits outside the orthogonality guarantee, so the
method is DML-assisted rather than fully orthogonal, and its overall score
is not insensitive to every estimated component.

The argument below needs the fitted index to be free of the observation's
own outcome, and two routes deliver that. A *flexible* imputation model is
cross-fit, so $\hat V_i$ is an out-of-fold prediction and is independent of
$Y_i$ given $(D, X)$. A *parametric* one is fit once on all the data, where
$\hat V_i$ depends on $Y_i$ only at order $1/n$, which is asymptotically
negligible. Cross-fitting a parametric imputation model is not merely
unnecessary but harmful: the per-fold index noise inflates the between-draw
variance, and we measured coverage rising to 1.00 with the mean standard
error at 2.3 times the true standard deviation. Accordingly the binary
machine-learning design below uses a cross-fit `gam` imputation model, while
the ordinal designs and the wine application fit `clm` once.

Either way $\hat V$ is fixed with respect to the completion, while
$Y \mid D, X$ is generated by the truth. Since
$S$ is $\hat V$ plus truncated noise,

$$E[S \mid D, X] = \psi(\hat V, \eta), \quad
\psi(v, e) = v + F(e)\,\mu_+(v) + (1 - F(e))\,\mu_-(v)$$ {#eq-ess}

with $\mu_\pm$ the truncated-noise mean corrections. The correctly centered
draw is exact, $\psi(e, e) = e$ (recovering Theorem 1 of @cheng2021),
but index error passes through with a factor strictly below one,

$$c(e) = \left.\frac{\partial \psi}{\partial v}\right|_{v=e}
= 1 - f(-e)\,\bigl(\mu_+(e) - \mu_-(e)\bigr), \qquad f = F'$$ {#eq-ce}

Read $c$ as the fraction of index error that survives the truncation. With
no truncation the surrogate would be index plus noise and its mean would move
one-for-one with $\hat V$, giving $c = 1$; the observed category drags it
back, and $c$ is what is left.

Under a logistic link $c$ rises from $0.31$ at $e = 0$ to $1$ in the tails
(@fig-passthrough). The value at the origin is exactly $1 - \log 2$, because
for the logit
$$c(e) = 1 - H\bigl(F(e)\bigr), \qquad H(p) = -p\log p - (1-p)\log(1-p),$$ {#eq-centropy}
so the self-correction is precisely the entropy of the outcome. That is the
whole mechanism in one line: at $p = 1/2$ the label is maximally uncertain,
so learning it is maximally informative about where $U$ fell and the
correction is largest; as $p \to 1$ the label was already predictable, tells
us nothing we did not have from $\hat V$, and cannot correct anything. It is
also why strong effects are the hard case, and why an ordinal category
spanning many error standard deviations behaves badly: both make the observed
outcome uninformative about $U$.

The density in @eq-ce is evaluated at $-e$, which matters for an asymmetric
$F$ such as the extreme-value link; writing $f(e)$ is a shorthand valid only
when $F$ is symmetric. @sec-theory gives the general-$J$ form of both
@eq-ess and @eq-ce, where the ordinal thresholds contribute their own
derivative block. For the well-placed partitions studied here, adding
categories lowers the pass-through at $e = 0$: from $0.307$ for binary to
$0.157$ and $0.090$ for the displayed three- and four-category
configurations. Category count alone is not monotone, however; widely spaced
thresholds can instead drive the pass-through toward one.

```{r}
#| label: fig-passthrough
#| echo: false
#| fig-width: 5
#| fig-height: 2.6
#| fig-cap: "Pass-through factor $c(e)$ under a logistic link. Imputation-model error at signal level $e$ reaches $\\hat\\theta$ multiplied by $c(e)$, which climbs from about 0.31 at $e=0$ toward 1 as the signal grows: the truncation self-corrects when $Y$ is informative and stops doing so once $Y$ is near-deterministic. Averaging $c$ over the index distribution gives the $\\bar c_1$ quoted in the text."
mu_plus  <- function(e) integrate(function(w) w * dlogis(w), -e, Inf)$value / plogis(e)
mu_minus <- function(e) integrate(function(w) w * dlogis(w), -Inf, -e)$value / plogis(-e)
ce <- function(e) 1 - dlogis(e) * (mu_plus(e) - mu_minus(e))
e <- seq(0, 6, by = 0.05)
cval <- sapply(e, ce)
par(mar = c(4, 4, 1, 1), cex = 0.9)
plot(e, cval, type = "l", lwd = 2, ylim = c(0, 1), las = 1,
     xlab = "index level e", ylab = expression("pass-through factor " * c(e)))
abline(h = 1, lty = 3, col = "grey60")
points(0, ce(0), pch = 19)
text(0.1, ce(0) + 0.1, sprintf("%.2f", ce(0)), pos = 4, cex = 0.9)
```

Linearizing $\psi$ around the truth, the bias of $\hat\theta$ is
approximately $\bar c_1\, \delta_D$ plus a covariate-direction leak given
below, where $\delta_D$ is the imputation model's error in its effective
treatment coefficient and $\bar c_1 = E[w\, c(\theta_0 + g)] / E[w]$,
$w = m(1-m)$.

Three consequences follow. First, the truncation *self-corrects*: because
$Y$ comes from the truth, it drags $S$ toward the correct interval, which
is why $c < 1$. But $c \to 1$ as the signal grows: once $Y$ is nearly
deterministic it adds no information beyond $\hat V$, and the self-
correction switches off. This, not any numerical instability, is why
inference is hardest at strong effect sizes.

Second, the covariate direction is not protected either, which is worth
spelling out because the opposite is the natural guess. Split the imputation
model's index error into a treatment part and a covariate part,
$\hat V - \eta = \delta_D D + \delta_X(X)$. Partialling-out regresses on
$R = D - m_0(X)$, and $E[R \mid X] = 0$ annihilates any function of $X$, so
$\delta_X$ ought to wash out. It does not, because $S$ is nonlinear in
$\hat V$: linearizing multiplies the index error by $c(\eta)$, and $\eta$
depends on $D$, so $\delta_X$ arrives multiplied by something that itself
varies with treatment and therefore survives the projection. For binary $D$,
writing $\eta_d = \theta_0 d + g_0(X)$ and $w = \mathrm{Var}(D \mid X)$, the
surviving term is
$$\text{covariate-direction leak} \;=\;
\frac{E\bigl[w\,\bigl(c(\eta_1) - c(\eta_0)\bigr)\,\delta_X\bigr]}{E[w]} .$$ {#eq-leak}
The leak is thus governed by *how differently the truncation damps the two
treatment arms*. Were $c$ the same in both, it would vanish exactly and the
naive argument would be right.

It is not always small. It does vanish when the imputation model can
represent $g_0$, which is why designs with every model correctly specified
show no covariate-direction bias. When it cannot, the leak bites: giving
$g_0$ an $X_1 X_2$ interaction that an additive imputation model cannot
express produces a bias of 5 to 6% of $\theta_0$ even though the treatment
enters that model correctly, so $\delta_D = 0$ and @eq-leak is the whole
story, and coverage falls to 0.75 at strong signal. Refitting the same design
with an imputation model that admits the interaction removes the bias,
confirming the imputation model rather than the adjustment nuisances as the
source.

Third, in the sharpest
contrast with standard DML, the imputation model's error enters
$\hat\theta$ *linearly*. In DML the outcome is observed and nuisance error
enters only through a *product* of two errors [@chernozhukov2018;
@robinson1988]: writing $\ell = \ell_0 + a$ and $m = m_0 + b$ for the errors
in the two adjustment models, the remaining bias is
$E[ab] - \theta_0 E[b^2]$, which is second order in the pair. Getting the
treatment model right ($b = 0$) removes it entirely; getting only the
outcome model right does not. The score is Neyman-orthogonal, first-order
insensitive to nuisance error, rather than doubly robust in the usual sense.
SUDO adds a third model whose treatment-direction error is unprotected and
first-order, damped only by $\bar c_1 < 1$. @tbl-sensitivity confirms this
numerically: a perturbation applied to the outcome nuisance in DML leaves
the estimate unbiased at every level, whereas the same-sized index
perturbation in SUDO produces a linear bias with slope exactly $\bar c_1$.

# Machine-learning imputation models

@eq-ess makes imputation-model quality matter more than
adjustment-model quality, the reverse of the usual DML intuition. A
well-specified parametric imputation model (logistic or cumulative-link)
needs no extra machinery, since its covariance supplies the proper draw
directly. Such draws propagate parameter uncertainty and attain near-nominal
coverage in our parametric experiments; @sec-theory characterizes the small
Rubin-variance discrepancy that remains. Two obstacles arise with a black-box learner: it can carry a
treatment-direction error $\delta_D$ (a bias problem), and it has no
covariance matrix from which to draw (an inference problem).

The bias is addressed by a *partially-linear* imputation model,
$\hat V = \hat\alpha D + \hat f(X)$, in which the treatment enters linearly
and machine learning handles only $X$, so no treatment-by-covariate
artifact can arise. This is a structural learner class, not a commitment to
one training algorithm. We implement it with a GAM, MARS bases followed by a
joint logistic fit, and neural-network backfitting. All three are unbiased
across effect sizes in the common comparison in @sec-sims. The inference problem is
addressed by a *full-pipeline bootstrap*, the nonparametric counterpart of
the proper draw: where a parametric model injects imputation-model
uncertainty by sampling $\beta$ from its covariance, here we resample the
whole data set and rerun the entire procedure, cross-fitting and all, using
the spread of the results as the standard error. Both target the same
variance component; the bootstrap simply obtains it by refitting when no
covariance is available to sample from. It has to wrap the *whole* pipeline,
because a within-fold refit of the model alone under-propagates that same
imputation-model variability. Together these restore nominal coverage
with a black-box imputation model (@sec-sims).

We report the bootstrap as an empirical result rather than a theorem. What
would be needed is asymptotic linearity of the imputation-model estimator
uniformly over resamples, the @sec-theory rate conditions holding
conditionally on the data, fold assignment redrawn within each resample, and
differentiability of the generated-outcome map. The first three are close to
available: @tang2024 give conditions for bootstrap consistency of
asymptotically linear estimators with machine-learned nuisances, and
@lin2026 establish bootstrap validity for general DML estimators under
exchangeably weighted resampling, under the same conditions that make DML
itself valid. Neither covers the generated-outcome step, which is the gap.

A Bayesian ensemble is the obvious alternative, since each posterior draw is
a complete alternative imputation model and so is a proper draw by
construction. The mechanism is even tidier than it first appears: under a
probit imputation model the Albert-Chib augmentation variable is exactly the
surrogate completion, so a Gibbs sweep returns one for free. We tried it and
it did not pay off. On a probit analogue of the machine-learning design,
imputation by BART, both unconstrained and with the partially-linear
constraint above, carried a bias that grew with signal strength to about 5%
of $\theta_0$, where the backfit is unbiased, and its posterior propagated
variance no better than the bootstrap it was meant to replace. That is one
design and not a general verdict, but we report it because the elegance
argument is the natural thing to try next and it did not survive contact.
The run is in the repository. A forest's out-of-bag or
infinitesimal-jackknife variance [@wager2014] is not such a route at all: it
quantifies prediction uncertainty at a point, not a coherent joint draw of
the whole imputation model.

# Simulation study {#sec-sims}

Every design follows one recipe. Covariates are drawn independently,
treatment is assigned from them, a latent utility is formed, and the observed
category is read off:

$$X_1, \dots, X_p \stackrel{\text{iid}}{\sim} N(0, 1), \qquad
D \mid X \sim \mathrm{Bernoulli}\bigl(\Lambda(\pi(X))\bigr), \qquad
U = \theta_0 D + g_0(X) + \varepsilon,$$ {#eq-dgp}
$$\varepsilon \sim F, \qquad
Y = j \iff \kappa_{j-1} < U \le \kappa_j,$$

with $\Lambda$ the logistic CDF. A design is then fixed by choosing
$(p, \pi, g_0, F, \kappa, \theta_0, n)$. Two covariate structures are used
throughout: a **one-covariate** setting with $\pi(X) = 0.8X$ and
$g_0(X) = X$, where every model in play is correctly specified and the
question is purely about pooling; and a **five-covariate machine-learning**
setting with
$$\pi(X) = X_4 + \cos X_5, \qquad g_0(X) = X_1^2 + \sin X_2 + 0.5 X_3,$$
which is nonlinear in $X$ and makes the treatment depend on covariates that
do not enter $g_0$. Binary outcomes take $\kappa = (0)$; ordinal ones cut
$U$ at two interior thresholds into $J = 3$ categories. @tbl-dgp in the
appendix lists every design's parameters.

Coverage targets $0.95$ and Monte-Carlo error on a coverage figure is about
$\pm 0.02$. **SD** is the estimator's true run-to-run variability and
**Mean SE** the standard error it reports; Mean SE below SD means intervals
that are too short. Learner and tuning specifications, fold and draw counts,
and per-cell Monte-Carlo standard errors are in @sec-appendix and the
committed result files, from which the tables below are generated at render
time.

```{r}
#| label: results-setup
#| include: false
results_dir <- "../../R/results"
rd <- function(f) read.csv(file.path(results_dir, f), check.names = FALSE)
sg <- function(x, d = 3) sprintf(paste0("%+.", d, "f"), x)  # signed
nn <- function(x, d = 3) sprintf(paste0("%.", d, "f"), x)   # unsigned
```

**Proper draws are what deliver coverage.** With a simple binary design and
a correct parametric imputation model, only the proper scheme reaches
nominal coverage; the naive interval covers 0.80 and the improper 0.93
(@tbl-ladder).

```{r}
#| label: tbl-ladder
#| echo: false
#| tbl-cap: "Inference schemes, binary outcome, $n = 5000$, $\\theta_0 = 1.5$, 500 replications. $B$ is the number of surrogate draws."
s2 <- rd("stage2_summary.csv")
r <- function(e) s2[s2$estimator == e, ]
tab <- data.frame(
  Scheme     = c("Naive single draw", "Improper, $B=25$", "Proper, $B=25$"),
  Bias       = sg(c(r("naive")$bias, r("improper_B25")$bias, r("proper_B25")$bias)),
  SD         = nn(c(r("naive")$sd, r("improper_B25")$sd, r("proper_B25")$sd)),
  `Mean SE`  = nn(c(r("naive")$mean_se, r("improper_B25")$mean_se, r("proper_B25")$mean_se)),
  Coverage   = nn(c(r("naive")$coverage, r("improper_B25")$coverage, r("proper_B25")$coverage)),
  check.names = FALSE)
knitr::kable(tab, align = c("l", "r", "r", "r", "r"))
```

**Ordinal outcomes.** With a three-category outcome and machine-learning
adjustment, SUDO is essentially unbiased with valid-to-conservative
coverage (@tbl-ordinal), a regime for which we are aware of no DML
competitor. The original implementation was substantially conservative. A
term-level ordinal decomposition includes the threshold block and matches
Monte Carlo throughout a clean cumulative-logit testbed. It predicts only
2.8% variance overstatement at moderate signal and 2.0% understatement at
strong signal, so threshold estimation and ordinal pooling alone cannot
explain the 50% to 88% variance gap after flexible partialling nuisances are
introduced. A paired nuisance-isolation experiment locates the gap in the
fixed outcome nuisance: refitting $E[S\mid X]$ for every ordinal completion
reduces mean $T$ by 38% while leaving the empirical sampling variance
unchanged, moving $T/V$ from 1.505 to 0.942. The full corrected stage at
$B=25$ and 200 replications has bias 0.016, coverage 0.960, and $T/V=1.214$
with Monte-Carlo standard error 0.122. Per-draw refitting is therefore the
validated ordinal procedure, at the cost of fitting the outcome nuisance
$B$ times rather than once.
The residual variance ratio is compatible with one at this simulation's
precision and is not evidence of a remaining ordinal anomaly. Rubin's
variance is guaranteed to err safe only under conditions we have not
verified here (a valid imputation model together with a self-efficient
complete-data analysis), and under uncongeniality it can be biased in
either direction [@meng1994; @xie2017]. The over-coverage is a property of
these designs, not a general guarantee.

```{r}
#| label: tbl-ordinal
#| echo: false
#| tbl-cap: "Ordinal outcome, $J = 3$, $\\theta_0 = 1.0$. Parametric arm 500 replications; corrected spline + ML arm 200, with the outcome nuisance refit per completion."
s4 <- rd("stage4_summary.csv"); s5 <- rd("stage5r_summary.csv")
p <- s4[s4$estimator == "sudo_ordinal_clm", ]
q <- s5[s5$estimator == "ordinal_refit_n2000_B25", ]
tab <- data.frame(
  Configuration = c("Parametric, $n=3000$", "Spline + ML nuisances, $n=2000$"),
  Bias      = sg(c(p$bias, q$bias)),
  SD        = nn(c(p$sd, q$sd)),
  `Mean SE` = nn(c(p$mean_se, q$mean_se)),
  Coverage  = nn(c(p$coverage, q$coverage)),
  check.names = FALSE)
knitr::kable(tab, align = c("l", "r", "r", "r", "r"))
```

**Link misspecification.** Generating Gumbel noise but assuming a logistic
link biases the latent-scale effect substantially (+0.77 binary, +0.58 ordinal,
coverage 0); the correct link removes it (bias $<0.005$, nominal coverage),
and the variance-drift diagnostic flags the wrong link (p $\approx$ 2e-4)
while clearing the right one (p $\approx$ 0.43).

**First-order versus second-order sensitivity.** Perturbing one input by a
known function and tracing the bias (@tbl-sensitivity) confirms the
propagation analysis: standard
DML with the observed outcome is unbiased under an arbitrarily wrong outcome
nuisance provided the propensity is right, whereas SUDO's index error is
linear, with treatment-direction slope matching the derived $\bar c_1$
(0.669 at $\theta_0 = 1.5$, 0.861 at $\theta_0 = 3$) to within 0.001.

| Arm | Perturbed | Bias vs. perturbation |
|---|---|---|
| DML | outcome nuisance only | flat, $\approx 0$ |
| DML | both nuisances | quadratic |
| SUDO | index, treatment direction | linear, slope $=\bar c_1$ |

: Nuisance-sensitivity ladder. {#tbl-sensitivity}

**Partially-linear imputation learners.** A tuned partially-linear backfit
is unbiased at both effect sizes; pairing it with the full-pipeline
bootstrap brings coverage at the strong effect from 0.86 to 0.98 for the
normal interval and 0.95 for the percentile interval. The latter is the
coverage claim used here. With 100 replications, either coverage estimate has
Monte Carlo uncertainty of about $\pm 0.03$; the bootstrap SD also matches
the estimator's run-to-run variability at this resolution.

Backfitting is not essential. A common-sample comparison fits the same
$\alpha D + f(X)$ structure three ways: a GAM with a linear treatment term,
MARS bases constructed from $X$ only followed by a joint logistic fit with
mandatory linear $D$, and the tuned neural-network backfit. The MARS
configuration is selected on theta-level Monte Carlo bias, not predictive
GCV. All three pass the prespecified bias and percentile-coverage criteria
(@tbl-pl-learners). With 50 replications, these coverage cells have Monte
Carlo standard errors of roughly 0.02 to 0.05, so their small differences do
not rank the learners.

```{r}
#| label: tbl-pl-learners
#| echo: false
#| tbl-cap: "Partially-linear imputation learners with full-pipeline bootstrap, $n=2000$, 50 replications, 50 outer resamples, and 15 surrogate completions per resample."
s3s <- rd("stage3s_summary.csv")
learner_name <- c(
  pl_gam = "GAM",
  pl_mars = "MARS",
  pl_backfit = "Neural backfit"
)
tab <- data.frame(
  Learner = unname(learner_name[s3s$learner]),
  `$\\theta_0$` = nn(s3s$theta, 1),
  Bias = sg(s3s$bias),
  `SD/Mean SE` = nn(s3s$sd_over_mean_se),
  `Normal coverage` = nn(s3s$coverage_normal),
  `Percentile coverage` = nn(s3s$coverage_percentile),
  check.names = FALSE
)
knitr::kable(tab, align = c("l", "r", "r", "r", "r", "r"))
```

**Continuous treatment, and other variations.** The designs above all use a
binary treatment, while the application below does not, so we check the
transfer directly. Replacing binary $D$ with $D = 0.7\pi(X) + N(0,1)$ in the
S5 structure leaves the estimator unbiased for both a binary outcome
($-0.1\%$ at $\theta_0 = 1$, $+0.4\%$ at $2.5$) and a three-category ordinal
one ($+0.7\%$ and $+1.5\%$), so the latent-scale estimand and the machinery
carry over. Doubling the covariate count to ten, and scaling the propensity
so that overlap is markedly worse, likewise cost nothing in bias. The
variation that does matter is the one described above, an imputation model
unable to represent $g_0$. Coverage under a continuous treatment tracks the
binary case at moderate signal but is worse at strong signal, where the
reported SE falls furthest short of the truth; the full-pipeline bootstrap is
the remedy there, as it is for the black-box case. Details and the full
grid are in @tbl-dgp and the repository.

# Application: volatile acidity and wine quality

We estimate the effect of volatile acidity (the acetic-acid
"vinegar/nail-polish" fault) on wine quality, an ordinal sensory score
(the median of blind tasters) on the UCI wine-quality data [@cortez2009],
1,599 red and 4,898 white wines. Quality spans six to seven ordinal
levels.

This is observational data, so a causal reading rests on an assumed causal
structure (@fig-dag). Volatile acidity is a plausible treatment because its
effect on quality has a clear expected sign (tasters penalize the acetic
fault) and because it is a fault endpoint rather than obviously upstream of
the other chemistry. A causal reading still rests on strong assumptions:
that the remaining physicochemical measurements are pre-treatment
adjustments rather than mediators or proxies of volatile acidity, and that
winemaking practice has no direct effect on quality beyond the measured
chemistry, the exclusion restriction the graph encodes. We do not fully
defend these assumptions, but we probe the two most in doubt, robustness to
the adjustment set and to unmeasured confounding, below; the application is
an illustration on a genuinely ordinal outcome rather than a settled causal
finding.

```{mermaid}
%%| label: fig-dag
%%| fig-width: 3.2
%%| fig-cap: "Assumed causal structure. Volatile acidity (the treatment) affects latent quality directly; unobserved winemaking practice confounds both through the measured chemistry, the adjustment set."
flowchart TD
  P((winemaking<br/>practice)):::u --> VA[volatile acidity]
  P --> X[other chemistry]
  X --> VA
  VA --> Q[latent quality]
  X --> Q
  classDef u fill:#eee,stroke-dasharray:4 3;
```

There is adequate residual variation: after adjusting for the other
chemistry, volatile acidity retains 75% (red) and 94% (white) of its
standard deviation (residual $\sqrt{1 - R^2}$ from the $R^2$ column of
@tbl-wine), so its residual is not annihilated and the residual-on-residual
regression is well posed. Applying ordinal SUDO (cumulative-link imputation
model, proper draws, cross-fitted nuisances, Rubin pooling) gives a clearly
negative effect on latent quality, stable across a linear and a
partially-linear (spline) imputation model (@tbl-wine). The naive
alternative of reading the coefficient straight off the cumulative-link
model reports a much smaller standard error (white: 0.031 versus
0.053-0.074) because it treats the parametric model as exactly correct and
ignores the completion and cross-fitting uncertainty SUDO propagates; this
is the same inference gap the simulations isolate (@tbl-ladder).

Volatile acidity is standardized before fitting, so these coefficients are
per one standard deviation (about $0.18$ g/dm$^3$ for red, $0.10$ for
white), which keeps the effect comparable across the two wines instead of
tied to the raw acetic-acid scale. The proportional-odds reading then makes
the magnitude concrete: a one-SD rise in volatile acidity multiplies the
odds of a wine scoring above any fixed quality level by $\exp(\hat\theta)$,
about $0.59$ for red ($\hat\theta = -0.53$) and $0.62$ to $0.66$ for white,
a reduction of roughly a third to 40% in those odds per standard deviation.

```{r}
#| label: tbl-wine
#| echo: false
#| tbl-cap: "Effect of a one-SD increase in volatile acidity on latent wine quality. Naive read-off from the linear cumulative-link model (standard error in parentheses); SUDO estimates with 95% intervals in brackets."
w <- rd("wine_application.csv")
wrow <- function(which, label) {
  lin <- w[w$wine == which & w$full_model == "linear", ]
  spl <- w[w$wine == which & w$full_model == "pl_spline", ]
  data.frame(
    Wine = label,
    `Overlap $R^2(D \\sim X)$` = nn(lin$overlap_r2, 2),
    `Naive read-off` = sprintf("$%.2f$ (%.2f)", lin$naive_theta, lin$naive_se),
    `SUDO (linear)`  = sprintf("$%.2f$ [$%.2f,%.2f$]", lin$sudo_theta, lin$ci_lo, lin$ci_hi),
    `SUDO (spline)`  = sprintf("$%.2f$ [$%.2f,%.2f$]", spl$sudo_theta, spl$ci_lo, spl$ci_hi),
    check.names = FALSE)
}
tab <- rbind(wrow("red", "Red"), wrow("white", "White"))
knitr::kable(tab, align = c("l", "r", "r", "r", "r"))
```

The sign is robust. Dropping any single covariate from the adjustment set
leaves the effect in $[-0.67, -0.46]$ for red and $[-0.49, -0.37]$ for
white, every interval excluding zero, so the estimate does not hinge on one
adjustment choice. Read through the omitted-variable-bias framework
[@cinelli2020], an unobserved confounder would have to explain about 16%
(red) and 11% (white) of the residual variation in both volatile acidity and
latent quality to bring the effect to zero; weaker confounding leaves it
negative. We report these robustness values as a heuristic calibration
rather than a formal bound: the Cinelli-Hazlett formula is derived for
omitted-variable bias in linear regression with an observed outcome, and we
are applying it to a pooled, surrogate-completed estimate whose standard
error comes from Rubin's rules. Its validity in that setting has not been
established. These values and the leave-one-out ranges are in the committed
result files.

**How wrong could the imputation model be?** Unmeasured confounding is the
threat any observational analysis faces; the one specific to SUDO is error in
the imputation model, which @eq-leak and @eq-ce show reaches $\hat\theta$ at
first order. That channel admits its own sensitivity statement, and unlike
the robustness value above it follows directly from the estimator's own
theory rather than being imported. Since the bias is $\bar c_1 \delta_D$, the
imputation model's error in the volatile-acidity coefficient that would drive
the estimate to zero is $\delta_D = -\hat\theta / \bar c_1$. Evaluating
$\bar c_1$ at the fitted index with the continuous-treatment multiplier
$E[c(\eta) D R]/E[R^2]$ and the general-$J$ pass-through of Proposition A1
gives $\bar c_1 = 0.21$ (red) and $0.17$ (white). The cumulative-link model
would therefore have to be wrong about volatile acidity by $+2.55$ and
$+2.38$ respectively, which is 4.2 and 4.8 times the coefficient it actually
fits. This calculation addresses treatment-direction error only. Conditional
on covariate-direction misspecification being negligible, it shows that a
large treatment-coefficient error would be required to erase the estimate;
it does not establish robustness to general imputation-model error.

Two things make it so, and both are worth noting because they are properties
of the design rather than luck. The pass-through is small, near $0.2$ rather
than the $0.67$ to $0.86$ of the binary simulations, because a six- or
seven-category outcome with well-spaced thresholds constrains the latent
utility tightly and the truncation self-corrects hard. And the estimate is
far from zero relative to its scale. Both would deteriorate for a coarse
outcome or a weak effect, which is exactly when this check is worth running.

# Discussion

SUDO extends double machine learning to binary and, to our knowledge for
the first time, ordinal latent-scale outcomes, by treating the latent
outcome as missing data and completing it. Under the logit link it targets
the same latent-scale coefficient as the binary-logistic DML estimator of
@liu2021lplr, while admitting other error distributions and ordinal as well
as binary responses. The
methodological core is inferential: the completions are multiple
imputations, and in our experiments only proper draws, which redraw the
imputation model's parameters (thresholds included), attain nominal
coverage. The method is link-flexible over any distribution for which the
truncated draw and reference are available, and its residuals give a
diagnostic that can flag, though not confirm, a wrong link.

The cost, made precise by the propagation analysis, is that the imputation model
sits outside the orthogonality guarantee: its error reaches the effect at
first order, through a truncation factor that weakens as the signal
strengthens. This differs from standard DML, where nuisance error is
second-order. It is also manageable: a well-specified parametric or a partially-linear
machine-learning imputation model, paired with a full-pipeline bootstrap,
attains nominal coverage throughout our experiments.

Several problems remain open, and we are explicit about what @sec-theory does
and does not settle. Its results are derivations with numerical
verification rather than fully discharged theorems: the regularity conditions
are stated but not minimized, the treatment-in-imputation-model step still
rests on simulation evidence, and the appendix covers only the parametric
imputation model fit once on all the data, not the cross-fit flexible regime
the machine-learning designs use. SUDO is therefore best read as a
simulation-supported estimator with a worked-out first-order theory, not a
fully proven inferential procedure. Extending the machine-learning results
beyond the data-generating structures studied, and to unordered (multinomial)
outcomes, which require a vector surrogate and have no single scalar
effect, are natural next steps. A Bayesian ensemble imputation model, whose
posterior draws are literal proper imputations, is the tidiest candidate for
replacing the bootstrap, but a BART version of it did not improve on the
bootstrap in the one design we tried; what would make that route work is
itself an open question.

# Data and software

The wine-quality data are from the UCI Machine Learning Repository
[@cortez2009]. The estimator is provided as a Python package (`sudo`)
together with a parallel set of R scripts that prototype and validate each
step; the code and the committed result files behind every table and figure
are in the open-source (MIT-licensed) project repository at
<https://github.com/bgreenwell/sudo>.

# Declaration of AI use

We used Generative AI to assist in writing code and summarizing reference materials. All outputs were carefully reviewed and validated by the authors, who retain full ownership of the final manuscript and accept sole responsibility for any errors.

# Appendix: asymptotic theory {#sec-theory}

This appendix collects the theory behind the claims in the body: the
pass-through map and its derivatives for general $J$, the moment properties
that separate the adjustment nuisances from the imputation model, the
asymptotic distribution of the pooled estimate at fixed $B$, and the
resulting account of Rubin's variance.

A word on status. We label the results propositions and theorems for ease of
reference, but they are derivations with numerical verification rather than
fully discharged theorems: proofs are given as sketches and the regularity
conditions are stated without being minimized. Every closed-form expression
below is checked numerically in the project repository, against numerical
differentiation for the pass-through derivatives and against Monte Carlo for
the variance decomposition, and we flag in place the two points where the
argument is incomplete.

## Notation and assumptions

The imputation model is a parametric family $P_\beta(Y \mid D, X)$ with
latent index $\eta(D, X; \beta)$; for a GLM with known cutpoints
$\eta = x'\beta$. When the cutpoints are estimated, as in a cumulative-link
model, we collect them with the index coefficients into a single parameter
$\beta = (\beta_\eta, \kappa)$. The thresholds are components of $\beta$ but
not of $\eta$: they enter only through the truncation interval, a
distinction that matters in @eq-gblocks below. Write $\hat\beta$ for the
maximum-likelihood estimate, $m_0(X) = E[D \mid X]$, $R = D - m_0(X)$, and
$J_\theta = E[R^2]$.

Given $\beta$, a completion is $S(W; \beta) = \eta(D, X; \beta) + \xi$ with
$\xi$ drawn from $F$ truncated to the interval $Y$ implies. The draw noise
splits into a data-measurable part and a pure Monte-Carlo part,

$$\xi_i^{(b)} = e_i + u_i^{(b)}, \qquad e_i = E[\xi_i \mid W_i; \beta],
\qquad E[u_i^{(b)} \mid W_i] = 0 ,$$ {#eq-split}

and this split is what makes the fixed-$B$ bookkeeping come out right.
Note that $\xi$ is *not* mean-zero given $W$: the truncation is selected by
the observed $Y$. Only after averaging over $Y \mid D, X$ at the true model
does the mean vanish, because composing $Y \sim P_{\beta_0}$ with the
matching truncated draw reproduces the untruncated law, so
$\xi \mid D, X \sim F$ exactly. Define

$$\sigma_e^2 = E[e^2 R^2], \quad
\sigma_u^2 = E[\operatorname{Var}(\xi \mid W) R^2], \quad
\sigma_\varphi^2 = E[\xi^2 R^2] = \sigma_e^2 + \sigma_u^2 .$$

We assume throughout: **(A1)** the latent model @eq-latent with
selection-on-observables, a constant latent-scale effect, conditional error
law $\varepsilon\mid D,X\sim F$, and
overlap ($J_\theta > 0$, $\operatorname{Var}(D \mid X)$ bounded away from 0
and $\infty$); **(A2)** a correctly specified, regular imputation model, so
$\eta(D, X; \beta_0) = \theta_0 D + g_0(X)$ and
$\sqrt n(\hat\beta - \beta_0) \to_d N(0, \Sigma_\beta)$ with
$\Sigma_\beta = I(\beta_0)^{-1}$ the inverse per-observation information and
influence function $\operatorname{IF}_\beta = I(\beta_0)^{-1} s_\beta(W)$;
**(A3)** cross-fitted adjustment nuisances consistent in $L_2$ with the
product rate
$\lVert \hat\ell - \ell_0 \rVert_2 \lVert \hat m - m_0 \rVert_2 = o_p(n^{-1/2})$,
each $o_p(n^{-1/4})$; **(A4)** finite fourth moments and dominated
differentiability of $\beta \mapsto E[\varphi]$ at $\beta_0$; and **(A5)**
proper draws $\beta^{(b)} \sim N(\hat\beta, \widehat\Sigma_\beta / n)$, so
that $\beta^{(b)} = \hat\beta + n^{-1/2} Z_b$ with
$Z_b \mid \text{data} \sim N(0, \Sigma_\beta)$.

The scaling in (A5) is worth stating explicitly, since $\Sigma_\beta$ is the
limit of $n \widehat{\operatorname{Cov}}(\hat\beta)$ rather than of
$\widehat{\operatorname{Cov}}(\hat\beta)$ itself; drawing with the
unnormalized matrix would inject variance that does not vanish with $n$.

**Scope.** (A2) describes the parametric case: one regular maximum-likelihood
fit on all the data, with a covariance to draw from. That is what the ordinal
designs and the wine application use, and it is the setting the results below
cover. It does not describe a cross-fit flexible imputation model, whose
index is assembled from per-fold fits with no single $\hat\beta$ and no
covariance of the required form. The binary machine-learning design and the
black-box program are therefore outside this appendix, and we report them
empirically rather than claiming the theory extends to them.

## The pass-through map for general $J$

For an outcome in category $j$ the draw truncates $\xi$ to $(a_j, b_j]$ with
$a_j = \kappa_{j-1} - v$ and $b_j = \kappa_j - v$ at index $v$. Write
$\mu_j(v, \kappa) = E[\xi \mid a_j < \xi \le b_j]$ and
$p_j = F(\kappa^0_j - e) - F(\kappa^0_{j-1} - e)$ for the true category
probability at true index $e$. Averaging over $Y \mid D, X$ gives the
general form of @eq-ess,

$$\psi(v, \kappa; e, \kappa^0) = v + \sum_{j=1}^{J} p_j\, \mu_j(v, \kappa).$$ {#eq-psiord}

**Proposition A1.** *At $(v, \kappa) = (e, \kappa^0)$, writing
$t_k = \kappa^0_k - e$,*

$$c(e) = \frac{\partial \psi}{\partial v}
= 1 - \sum_{k=1}^{J-1} f(t_k)\bigl[\mu_{k+1}(e) - \mu_k(e)\bigr],
\qquad
\frac{\partial \psi}{\partial \kappa_k} = f(t_k)\bigl[\mu_{k+1}(e) - \mu_k(e)\bigr].$$ {#eq-gblocks}

*Consequently $c(e) = 1 - \sum_k \partial \psi / \partial \kappa_k$: the
index derivative and the threshold derivatives are the same kernel.*

*Sketch.* Differentiate @eq-psiord. Both truncation limits move with slope
$-1$ in $v$, and the derivative of a truncated mean with respect to its
upper limit is $f(b)(b - \mu)/P$, with the analogous expression at the lower
limit. Evaluated at the truth the truncation probability $P_j$ equals $p_j$,
so the weights in @eq-psiord cancel the denominators and the surviving terms
regroup by threshold. The final identity follows because $\mu_j$ depends on
$(v, \kappa)$ only through $\kappa - v$, so shifting the index is the same
as shifting every threshold the other way. $\square$

Three consequences. First, the binary case is $J = 2$ with $\kappa_1 = 0$,
where the sum collapses to @eq-ce. Second, the threshold derivatives are
generally nonzero, so for an ordinal outcome both $\Sigma_\beta$ and the
gradient $G$ of the next section carry a threshold block. This is the formal
reason the ordinal full model must be fit with a routine whose covariance
covers the thresholds (`ordinal::clm`) rather than one whose does not
(`mgcv::ocat`). Third, what governs the pass-through is how much the
categorization tells you about $U$, which is *not* the same as the number of
categories. Refining a partition with well-placed thresholds does lower $c$
(at $e = 0$ under a logit, from $0.307$ binary to $0.157$ and $0.090$ for the
three- and four-category configurations of @sec-numverify), but widening the
same three-category partition reverses this completely, taking $c(0)$ above
the binary value and on toward 1. The reason is visible in @eq-gblocks: a
category spanning many error standard deviations has a nearly unbounded
truncation interval, its $\mu_j$ barely moves with the index, and the
self-correction switches off. We therefore claim no monotonicity in $J$. The
useful statement is that $c \to 1$ as the observed category stops
constraining $U$, which is the same mechanism that drives $c \to 1$ at strong
signal.

**The binary logit closed form.** For $J = 2$ under a logistic $F$, the
truncated means have an exact expression that turns @eq-ce into @eq-centropy.
Substituting $u = F(x)$, so that $x = \log\{u/(1-u)\}$ and $\mathrm{d}u = f(x)\,\mathrm{d}x$,
the partial mean is
$$\int_{-\infty}^{t} x f(x)\,\mathrm{d}x
 = \int_0^{F(t)} \log\frac{u}{1-u}\,\mathrm{d}u
 = q\log q + (1-q)\log(1-q) = -H(q), \qquad q = F(t),$$
so the partial mean of a logistic is minus the entropy at the corresponding
probability. Taking $t = -e$ and $p = F(e)$, and using $H(1-p) = H(p)$ with
the fact that the untruncated mean is zero,
$$\mu_-(e) = \frac{-H(p)}{1-p}, \qquad \mu_+(e) = \frac{H(p)}{p},
\qquad \mu_+ - \mu_- = \frac{H(p)}{p(1-p)} .$$
Since $f(-e) = f(e) = p(1-p)$ for the logistic, the factor cancels exactly in
@eq-ce and $c(e) = 1 - H(p)$, giving $1 - \log 2 = 0.307$ at $e = 0$. The
cancellation is what makes the self-correction *equal* to the outcome's
entropy rather than merely proportional to something like it.

## Moment properties

**Proposition A2.** *Under (A1)-(A2), $E[\varphi] = 0$ at the truth and the
moment is Neyman-orthogonal in the adjustment nuisances,
$\partial_\ell E[\varphi] = \partial_m E[\varphi] = 0$. It is not orthogonal
in $\beta$: $\partial_\beta E[\varphi] = G$ with blocks*

$$G_{\beta_\eta} = E\bigl[c(\eta_0)\, \partial_{\beta_\eta} \eta\, R\bigr],
\qquad
G_{\kappa_k} = E\bigl[f(\kappa^0_k - \eta_0)\,(\mu_{k+1} - \mu_k)\, R\bigr].$$

*Sketch.* At the truth $S - \ell_0 - \theta_0 R = \xi$, and averaging over
$Y \mid D, X$ makes $\xi \mid D, X \sim F$, so $E[\varphi] = E[\xi R] = 0$.
The mean-zero statement holds only after that averaging; given $(W, \beta)$
the draw has mean $e \ne 0$. The nuisance derivatives vanish by the tower
property. The $\beta$ derivative is @eq-gblocks composed with
$\partial_\beta \eta$. $\square$

Because any function of $X$ is annihilated by $R$, profiling $\ell$ out of
the moment leaves $G$ unchanged. That covers either implementation: refitting
$\hat\ell$ on every completion, or fitting it once from a plug-in completion
and holding it fixed, which is what the estimator actually does (see the
appendix section on nuisances and draws). Proposition A2 is the formal
content of the statement in the body that the imputation model sits outside
the orthogonality guarantee.

## Asymptotic distribution at fixed $B$

**Theorem A1.** *Under (A1)-(A5), with
$\bar\theta = B^{-1}\sum_b \hat\theta^{(b)}$,*

$$\sqrt n(\bar\theta - \theta_0) \to_d N(0, V_B), \qquad
V_B = J_\theta^{-2}\Bigl[\sigma_e^2 + G'\Sigma_\beta G + 2G'C
  + \frac{G'\Sigma_\beta G + \sigma_u^2}{B}\Bigr],$$ {#eq-vb}

*where $C = \operatorname{Cov}(eR, \operatorname{IF}_\beta)$.*

*Sketch.* The $Z$-estimator expansion gives
$\sqrt n(\hat\theta - \theta_0) = J_\theta^{-1}\sqrt n M_n(\theta_0, \cdot) + o_p(1)$,
the sign following from $\partial_\theta E[\varphi] = -J_\theta$. Expanding
around $(\beta_0, \ell_0, m_0)$, the adjustment-nuisance terms are $o_p(1)$
by Proposition A2 plus cross-fitting plus the (A3) product rate, as in
@chernozhukov2018. The $\beta$ term is
$G'\sqrt n(\beta^{(b)} - \beta_0) + o_p(1)$, and
$\beta^{(b)} - \beta_0 = (\hat\beta - \beta_0) + n^{-1/2} Z_b$ by (A5).
Applying @eq-split separates the data-measurable score $e_i R_i$ from the
draw noise; averaging over $b$ leaves parameter-draw and surrogate-draw
terms with conditional variances $\Sigma_\beta / B$ and $\sigma_u^2 / B$.
$\square$

Two features of @eq-vb deserve emphasis. First, this is not the
single-completion result plus an $O(1/B)$ correction: the leading term
itself depends on $B$. A single completion carries the full sandwich
$\sigma_e^2 + \sigma_u^2$ *and* twice the parameter contribution, once from
$\hat\beta$'s own sampling error and once from the single draw around it,
which is exactly what Rubin's $(1 + 1/B)$ factor exists to undo. As
$B \to \infty$ the draw noise averages away and the leading term is
$\sigma_e^2$, not $\sigma_e^2 + \sigma_u^2$. Second, the cross term $2G'C$
is not sign-restricted, since $\operatorname{IF}_\beta$ and the moment are
built from the same data. So $V_B$ is not the DML sandwich plus a
nonnegative imputation penalty; it is the sandwich with the draw noise
removed by pooling and $G'\Sigma_\beta G + 2G'C$ added.

## The pooled variance

Rubin's total is $T = \bar W + (1 + 1/B) B_{\!*}$. The two pieces behave
differently, and the distinction matters. $\bar W$ averages a sandwich
estimate over $n$ observations, so $n\bar W \to_p J_\theta^{-2}(\sigma_e^2 + \sigma_u^2)$.
$B_{\!*}$ is the sample variance of $B$ draws. Writing
$V_{\mathrm{draw}} = G'\Sigma_\beta G + \sigma_u^2$ for the per-draw
variance, @eq-vb gives
$$n B_{\!*} \to_d J_\theta^{-2}\, V_{\mathrm{draw}}\,\frac{\chi^2_{B-1}}{B-1}
\qquad (n \to \infty,\ B \text{ fixed}),$$ {#eq-bstar}
which is *not* a probability limit: at fixed $B$ the between-draw term keeps
sample-variance randomness however large $n$ becomes. It is conditionally
unbiased for $J_\theta^{-2}V_{\mathrm{draw}}$, which is what lets the
comparison below be stated in expectation.

Write $E_\Delta[\cdot] = E[\cdot \mid W_1, \dots, W_n]$ for the expectation
over the imputation draws with the data held fixed, so that $E_\Delta[T]$ is
still a random variable through the data.

**Theorem A2.** *Under (A1)-(A5), for fixed $B$,*
$$n\,E_\Delta[T] - V_B \;\to_p\; 2 J_\theta^{-2}\bigl(\sigma_u^2 - G'C\bigr),$$
*with $n\,T$ itself converging in distribution, not in probability, by
@eq-bstar. Taking expectations over the data as well gives the same limit for
the unconditional $n\,E[T] - V_B$, which is the form the Monte-Carlo check
below estimates, since each replication redraws both the sample and the
imputations. As $B \to \infty$ the randomness in $B_{\!*}$ averages away and
the statement strengthens to a probability limit for $T$ itself,*
$$n\,T - V_\infty \;\to_p\; 2 J_\theta^{-2}\bigl(\sigma_u^2 - G'C\bigr),
\qquad V_\infty = J_\theta^{-2}\bigl(\sigma_e^2 + G'\Sigma_\beta G + 2G'C\bigr).$$
*Either way the $1/B$ terms cancel, so the discrepancy is free of $B$, and
Rubin's variance is asymptotically exact if and only if $G'C = \sigma_u^2$,
conservative when $G'C < \sigma_u^2$, and anti-conservative when
$G'C > \sigma_u^2$.*

**Remark (what fixed $B$ costs).** With $B$ fixed, no amount of data removes
the $\chi^2_{B-1}$ fluctuation in the reported variance. This establishes
that *some* non-normal reference distribution is necessary at any $n$, not
merely as a small-sample refinement. It does not by itself establish that the
Barnard-Rubin degrees of freedom are the right one: that formula is derived
for a complete-data analysis whose within- and between-imputation components
are independent, whereas here $\bar W$ and $B_{\!*}$ are built from the same
draws and the same $\hat\beta$. Establishing the required dependence
structure is open. The predicted spread is large and is confirmed
numerically in @sec-numverify.

Neither direction is automatic, and we do not find exactness. On a
deliberately congenial testbed that isolates the pooling from every other
source of error (a correctly specified logistic imputation model, continuous
treatment so both adjustment nuisances are exactly linear), all six
quantities are available in closed form:

| $\theta_0$ | $\sigma_e^2$ | $\sigma_u^2$ | $G'\Sigma_\beta G$ | $G'C$ | $\sigma_u^2 - G'C$ | $T/V$ at $B = 20$ |
|---|---|---|---|---|---|---|
| 1 | 1.269 | 2.021 | 4.077 | 1.954 | $+0.067$ | 1.014 |
| 3 | 0.580 | 2.710 | 41.414 | 3.094 | $-0.384$ | 0.985 |

The sign of the discrepancy flips with signal strength: on average $T$
overstates the sampling variance by about 1.4% at $\theta_0 = 1$ and
understates it by about 1.5% at $\theta_0 = 3$. The $T/V$ column compares
$n\,E[T]$ with $V_B$, which is the quantity Theorem A2 governs at fixed $B$;
individual replications scatter around it according to @eq-bstar. Both
discrepancies are asymptotic rather than finite-sample effects; they persist
at $n = 20{,}000$ and, being free of $B$, do not shrink as $B$ grows. The
mechanism is visible in the table. As the signal grows the
observed $Y$ carries less information beyond the index, so $\sigma_e^2$
falls while $\sigma_u^2$ and $G'C$ rise, and $G'C$ overtakes $\sigma_u^2$.
This is a more precise account of why inference is hardest at strong effect
sizes than the informal argument in the body, and it corrects a natural but
mistaken reading: $c \to 1$ is the regime in which $\psi$ is *closest* to
the identity, so the linearization is at its most accurate there, and cannot
be what degrades.

The practical import is small: a 1.5% error in a variance is roughly 0.7% in
a standard error, well inside the Monte-Carlo error of the coverage figures
in @sec-sims. The theoretical import is that "proper draws plus Rubin's rules
are exact here" is not a claim we can make. @sec-numverify reports the
Monte-Carlo check of @eq-vb term by term.

**Corollary (improper draws under-cover).** If $\beta^{(b)} \equiv \hat\beta$,
the between-draw term loses the parameter contribution entirely, so $B_{\!*}$
is conditionally unbiased for $J_\theta^{-2}\sigma_u^2$ alone and $n\,E[T]$
falls short of $V_B$.
The deficit is dominated by the imputation channel $G'\Sigma_\beta G$
whenever $G \ne 0$ and the signal is not weak, though it is not exactly that
quantity, since the cross term and the draw noise also enter. This is the
0.93 versus 0.97 coverage gap in @tbl-ladder.

## Bias under a misspecified imputation model

**Proposition A3.** *Drop the correct-specification half of (A2) and suppose
$\hat\beta \to_p \beta_*$. Restrict to binary $D$ and to a pure
treatment-coefficient error, so that the index error
$\eta(\cdot; \beta_*) - \eta_0$ equals $\delta_D$ in the treated arm and $0$
in the control arm. Then*
$$\bar\theta - \theta_0 = \bar c_1 \delta_D
  + \textit{(covariate-direction leak)} + o_p(1), \qquad
\bar c_1 = \frac{E[w\, c(\theta_0 + g_0)]}{E[w]}, \quad w = \operatorname{Var}(D \mid X).$$

*Sketch.* Linearize
$E[S \mid D, X; \beta_*] \approx \eta_0 + c(\eta_0)(\eta(\cdot; \beta_*) - \eta_0)$
and push through $J_\theta^{-1}E[\,\cdot\,R]$. Conditioning on $X$ with
binary $D$, the moment weights the treated arm by $w$ and the control arm by
$-w$; the control-arm error is zero by assumption, so only
$c(\theta_0 + g_0)$ survives, and $J_\theta = E[w]$ supplies the
denominator. $\square$

Both restrictions matter. The displayed $\bar c_1$ evaluates $c$ on the
treated arm only, which is correct precisely because the control-arm error
was set to zero; a general perturbation weights both arms as
$m\,c(\theta_0 + g_0) + (1 - m)\,c(g_0)$. For continuous $D$ with index
error $\delta_D D$ the multiplier is instead
$\bar c_1 = E[c(\eta_0) D R] / E[R^2]$, which is a different expression. The
covariate-direction leak is @eq-leak, nonzero because $c$ depends on $D$
through $\eta$ and so is not annihilated by projecting on $R$. We do not
bound it. It is not negligible in general: with an imputation model unable to
represent $g_0$ it reaches 5 to 6% of $\theta_0$ while $\delta_D = 0$, so
nothing here guarantees it is small relative to $\bar c_1 \delta_D$.
@tbl-sensitivity is the empirical counterpart for the treatment direction,
recovering the predicted slope $\bar c_1$.

# Appendix: simulation and application details {#sec-appendix}

Every design draws a latent utility $U = \theta_0 D + g_0(X) + \varepsilon$
with logistic $\varepsilon$ and reads off the observed category; the designs
differ in the covariate structure, the cutpoints, and the models fit.
Covariates are standard normal throughout. Full data-generating code and the
committed per-cell result files are in `R/` and `R/results/`.

## Data-generating processes

All designs instantiate @eq-dgp with one of two covariate structures:

- **S1**, one covariate: $p = 1$, $\pi(X) = 0.8X$, $g_0(X) = X$. Every model
  in play is correctly specified, so only the pooling is under test.
- **S5**, five covariates: $p = 5$, $\pi(X) = X_4 + \cos X_5$,
  $g_0(X) = X_1^2 + \sin X_2 + 0.5 X_3$. Nonlinear in $X$, and the treatment
  depends on covariates absent from $g_0$.

| design | structure | $F$ | $\kappa$ | $\theta_0$ | $n$ | reps |
|---|---|---|---|---|---|---|
| Inference ladder (@tbl-ladder) | S1 | logistic | $(0)$ | 1.5 | 5000 | 500 |
| Ordinal parametric (@tbl-ordinal, top) | S1 | logistic | $(-1, 1)$ | 1.0 | 3000 | 500 |
| Binary ML | S5 | logistic | $(0)$ | 0.5, 3 | 1000, 2000 | 200 |
| Ordinal ML (@tbl-ordinal, bottom) | S5 | logistic | $(0, 2)$ | 1.0 | 1000, 2000 | 200 |
| Link misspecification | S1 | Gumbel | $(0)$ or $(0, 2)$ | 1.0 | 5000 | 200 |
| Sensitivity (@tbl-sensitivity) | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 200 |
| Black-box imputation | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 100 |
| PL learner comparison (@tbl-pl-learners) | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 50 |
| Bayesian imputation (repository only) | S5$'$ | normal | $(0)$ | 1.5, 3 | 2000 | 200 |
| DGP variation, continuous $D$ | S5, $D = 0.7\pi(X) + N(0,1)$ | logistic | $(0)$ or $(0, 2)$ | 1, 2.5 | 2000 | 200 |
| DGP variation, interaction | S5 plus $X_1 X_2$ in $g_0$ | logistic | $(0)$ | 1, 2.5 | 2000 | 200 |
| DGP variation, dimension | S5 with $p = 10$ | logistic | $(0)$ | 1, 2.5 | 2000 | 200 |
| DGP variation, overlap | S5, $\pi(X)$ scaled $2.5\times$ | logistic | $(0)$ | 1, 2.5 | 2000 | 200 |

: Data-generating processes. $\kappa$ lists the interior cutpoints, so
$(0)$ is binary and a pair gives $J = 3$. {#tbl-dgp}

S5$'$ is the probit analogue of S5 used for the Bayesian imputation model,
whose binary machinery is probit: the systematic part is divided by
$\mathrm{sd}(\text{logistic}) = \pi/\sqrt3$ before normal errors are added,
which matches the event rates of S5 to within 0.006 and puts $\theta_0$ on
the probit scale at $\theta_0/1.8138$. The $\theta_0$ column reports the
logit-scale values for comparability. That design supports no table here;
it is reported so the claim in the machine-learning section is checkable.

**Models fitted.** The ladder uses $\mathrm{glm}(Y \sim D + X)$ as the
imputation model with $B \in \{10, 25, 50\}$ draws (the table shows
$B = 25$). The ordinal parametric design uses `ordinal::clm`, whose
covariance covers the thresholds, with $B = 25$. The ordinal ML design uses
`clm` on $D$ and natural-spline bases $\mathrm{ns}(X_k, 4)$ with `mgcv::gam`
partialling nuisances and $B = 25$.

**Link misspecification.** The one-covariate structure with extreme-value
$\varepsilon$ rather than logistic: Gumbel-max for the binary case, matching
`glm` with a cloglog link, and Gumbel-min for the ordinal case, matching
`ordinal::clm` with a cloglog link. The analyst fits either logit
(misspecified) or cloglog (correct). Note this design deliberately keeps the
one-covariate structure, so that the only thing wrong is the link.

**Sensitivity ladder (@tbl-sensitivity).** Starts from the oracle index and
oracle partialling nuisances, then adds a known perturbation of size $\delta$
to one input: the outcome nuisance only (DML-1), both nuisances (DML-2), the
index in the covariate direction (SUDO-X), or the index in the treatment
direction (SUDO-D). Bias is traced against a common latent-scale RMSE so the
four are comparable. The fitted SUDO-D slope reproduces $\bar c_1$ (0.669 at
$\theta_0 = 1.5$, 0.861 at $\theta_0 = 3$) to within 0.001, which is the
independent check on @eq-ce.

**Partially-linear imputation models.** The original black-box result uses
$\hat V = \hat\alpha D + \hat f(X)$ with $\hat f$ a single-hidden-layer
neural net (`nnet`, `size = 4`, `decay = 0.3`, five backfitting iterations,
selected on theta-level MC bias) and 100 full-pipeline bootstrap resamples.
The common learner comparison uses 50 resamples for each arm. Its GAM has an
unpenalized linear treatment term and smooths of the five covariates. Its
MARS fit builds degree-two bases from $X$ only (`nk = 21`, penalty 3), then
jointly estimates those basis coefficients and a mandatory linear treatment
coefficient by logistic regression. This construction is needed because
`earth::linpreds` alone still permits the named predictor in interaction
terms.

**Wine application.** Full model `ordinal::clm` on $D$ and either the raw
covariates (linear) or $\mathrm{ns}(X_k, 3)$ (spline); partialling nuisances
$E[S \mid X]$ and $E[D \mid X]$ by `mgcv::gam` (Gaussian for the continuous
$D$); five-fold cross-fitting; $B = 50$ proper draws; Rubin pooling. Volatile
acidity is standardized.

## Nuisances, cross-fitting, and draws

Partialling nuisances $\hat\ell(X) = E[S \mid X]$ and $\hat m(X) = E[D \mid X]$
are `mgcv::gam` throughout, fit with five-fold cross-fitting. Proper draws
sample the full-model parameters from $N(\hat\beta, \widehat{\mathrm{Cov}}(\hat\beta))$
(thresholds included for `clm`) before each completion. Diagnostic draws use
an RNG stream independent of the estimation draws.

One algorithmic detail matters enough to state. The surrogate $S$ changes with
every draw, so its nuisance $\hat\ell$ can be refit each time. Earlier binary
experiments found refitting negligible, so the binary estimator fits
$\hat\ell$ once on a plug-in completion and reuses it. The ordinal
nuisance-isolation ladder finds the opposite: holding $\hat\ell$ fixed nearly
doubles the within-draw component and accounts for the stage-5 conservatism.
The corrected ordinal estimator therefore cross-fits $\hat\ell$ separately
on every completion. The full $B=25$, 200-replication validation attains
0.960 coverage. The correction costs roughly $B$ times as much for the
outcome nuisance, which is why the cheaper binary behavior is retained where
it has been validated.

## Monte-Carlo precision

Coverage is a binomial proportion, so its Monte-Carlo standard error at a
true 0.95 is about $\sqrt{0.95 \cdot 0.05 / R}$: 0.010 at $R = 500$
replications, 0.015 at $R = 200$, and 0.022 at $R = 100$. Per-cell bias and
coverage Monte-Carlo standard errors are in the committed result files.

# Supplement: numerical verification of the theory {#sec-numverify}

Every closed-form expression in @sec-theory is checked against an independent
numerical route. Both scripts are in the project repository and assert their
own pass conditions.

## Pass-through derivatives (Proposition A1)

`R/theory_ordinal_passthrough.R` computes $\psi$ by quadrature and compares
the analytic index and threshold derivatives of @eq-gblocks against central
differences, for $J = 2, 3, 4$ under logit, probit, and both Gumbel
conventions. Maximum discrepancy across all configurations is $3\times10^{-10}$
for the index block and $4\times10^{-10}$ for the threshold block; the
identity $c = 1 - \sum_k \partial_{\kappa_k}\psi$ holds to $10^{-15}$. A
Monte-Carlo check confirms `complete_surrogate()` samples the map it should
($|z| < 4$ at $4\times10^5$ draws), so the theory describes the implemented
estimator and not merely the model.

The threshold-placement effect discussed in @sec-theory, at $e = 0$ under a
logit:

| $\kappa$ | $(0)$ binary | $(-0.5, 0.8)$ | $(-1, 0.2, 1.3)$ | $(\mp 0.5)$ | $(\mp 1.5)$ | $(\mp 3)$ | $(\mp 6)$ |
|---|---|---|---|---|---|---|---|
| $c(0)$ | 0.307 | 0.157 | 0.090 | 0.175 | 0.223 | 0.636 | 0.966 |

The last four entries are all $J = 3$: widening a three-category partition
drives $c$ from well below the binary value to nearly 1.

## The variance decomposition (Theorems A1 and A2)

`R/theory_variance_terms.R` evaluates $J_\theta$, $\sigma_e^2$, $\sigma_u^2$,
$G$, $\Sigma_\beta$ and $C$ in closed form on the congenial testbed by
product-grid quadrature, then compares @eq-vb against 4000 replications at
$n = 20{,}000$ for $B \in \{1, 2, 5, 20\}$ and $\theta_0 \in \{1, 3\}$. All
eight cells agree within 1.2 Monte-Carlo standard errors. Comparing across
$B$ is what identifies the individual terms; a single pooled variance ratio
cannot, because its Monte-Carlo error is the same size as the 1 to 2 percent
effect being measured, which is why the earlier ratio-based reading of these
simulations was inconclusive rather than confirmatory.

The fixed-$B$ spread of @eq-bstar is confirmed directly: at $\theta_0 = 1$,
$B = 20$, $n = 20{,}000$, the observed standard deviation of $nT$ across
replications is 2.13 against a mean of 9.73, while @eq-bstar predicts
$(1 + 1/B)V_{\mathrm{draw}}\sqrt{2/(B-1)} = 2.08$. The residual spread in the
reported variance is therefore the between-draw sample variance, not
estimation error, and it does not shrink with $n$.

Both closed forms behind these quantities rest on one identity,
$p(1-p)(\mu_+ - \mu_-) = H(p)$ for the logistic law, which yields both
$c(e) = 1 - H(F(e))$ and $C = \Sigma_\beta(A - G)$ with $A = E[zR]$. It is
verified pointwise by quadrature to $8\times10^{-16}$.

The ordinal counterpart is checked by
`R/theory_ordinal_variance_terms.R` on a correctly specified
three-category cumulative-logit design with continuous treatment and exactly
linear partialling nuisances. The calculation includes both threshold
derivatives in $G$ and matches the fixed-$B$ sampling variance and expected
Rubin variance within 2.3 Monte-Carlo standard errors in every cell at
$n=8{,}000$, 500 replications, $B\in\{1,2,5,20\}$, and
$\theta_0\in\{1,2.5\}$. At $B=20$, the predicted $T/V$ is 1.028 and 0.980,
respectively. The ordinal threshold machinery therefore changes the same
small Rubin discrepancy identified in the binary testbed but does not explain
the much larger stage-5 gap.

`R/theory_ordinal_nuisance_ladder.R` then changes one partialling component at
a time on the exact stage-5 DGP, using common samples, folds, imputation fits,
parameter draws, and surrogates across arms. At $n=2{,}000$, $B=10$, and 100
replications, the current fixed-$\hat\ell$ arm has empirical variance 0.01380
and mean $T=0.02077$, while refitting $\hat\ell$ on every draw has empirical
variance 0.01375 and mean $T=0.01295$. Thus the point estimator is unchanged
but $T/V$ falls from 1.505 to 0.942. The difference is almost entirely in
$\bar W$, which falls from 0.01666 to 0.00876; the between-draw component is
unchanged. This isolates reuse of the plug-in outcome nuisance as the source
of the stage-5 variance inflation. `R/stage5r_ordinal_refit.R` performs the
full correction check at $n=2{,}000$, $B=25$, and 200 replications. Bias is
0.016 (Monte-Carlo SE 0.007), coverage is 0.960 (Monte-Carlo SE 0.014), and
$T/V=1.214$ (Monte-Carlo SE 0.122), satisfying the stage's bias, coverage,
and variance-calibration criteria. The corrected per-draw refit is therefore
the ordinal default.

# References

::: {#refs}
:::
