---
title: "Surrogate-Assisted Double Machine Learning: Causal Inference with Binary and Ordinal Outcomes"
author:
  - name: Brandon M. Greenwell
    email: greenwbm@ucmail.uc.edu
    affiliations:
      - name: "University of Cincinnati Carl H. Lindner College of Business"
        department: Department of Operations, Business Analytics, and Information Systems
        address: 2906 Woodside Drive
        city: Cincinnati, Ohio
        country: United States
        postal-code: 45221
    orcid: 0000-0002-8120-0084
  - name: Other Authors
    email: other.authors@email.com
    affiliations:
      - name: "University of XYZ"
authorcols: true
date: today
abstract: |
  Double Machine Learning (DML) pairs Neyman-orthogonal scores with cross-fitting to estimate causal effects under flexible adjustment. For categorical outcomes, existing DML methods can target probability- or distribution-scale contrasts, but the partially linear latent-utility coefficient requires different machinery. A binary logistic estimator supplies that coefficient under a logit link; we are aware of no ordinal counterpart. We introduce SUDO (Surrogate-Assisted DML), which treats the latent outcome behind a binary or ordinal response as missing data. Each observation is completed to a continuous *surrogate* drawn from the assumed error distribution and truncated to the interval implied by its observed category, after which ordinary partialling-out applies. The completions form multiple imputations. In the parametric designs studied, Rubin pooling attained near-nominal coverage only with *proper* draws that redraw the imputation model's parameters, including ordinal thresholds; naive and improper draws under-covered. The method accommodates multiple links, and its surrogate residuals provide a specification diagnostic. We also derive how imputation-model error reaches the target: it enters at first order through a truncation factor that rises from about 0.3 to 1 as the signal strengthens. For a deterministic logistic-series learner, we prove a projected expansion and full-pipeline bootstrap result under explicit series conditions. In the machine-learning designs studied, normal intervals based on a full-pipeline bootstrap attained nominal coverage for generalized additive model (GAM), multivariate adaptive regression spline (MARS), and neural-backfit imputation learners; extending the proof to these adaptive algorithms remains open. We illustrate the method on wine-quality data by estimating the latent-scale effect of volatile acidity on an ordinal quality score.
format:
  arxiv-pdf:
    keep-tex: false
bibliography: references.bib
---

# Introduction

Many outcomes that scientists want causal effects for are categorical.
A patient relapses or not; a wine is rated on a five- or ten-point scale;
a survey respondent is dissatisfied, neutral, or satisfied. For a
*continuous* outcome, Double Machine Learning [DML, @chernozhukov2018]
provides a now-standard route to a causal effect under flexible,
high-dimensional adjustment. It predicts the outcome and treatment from the
covariates with machine-learning (ML) models, removes both predictions, and
reads the effect from a residual-on-residual regression. This
Frisch-Waugh-Lovell (FWL) orthogonalization allows flexible nuisance
predictions because, under standard convergence-rate conditions, the
Neyman-orthogonal score is insensitive to their first-order estimation error
[@chernozhukov2018].

That partially linear machinery is built for an observed continuous outcome.
DML can be applied to categorical outcomes through other scores, but those
procedures target probability- or distribution-scale contrasts. A binary or
ordinal outcome obeys the additive structure used here only on a *latent*
scale. An unobserved continuous utility follows
$U = \theta_0 D + g_0(X) + \varepsilon$, while the data reveal only the
interval containing $U$. Applying continuous partialling-out to the category
labels is computationally possible, but it changes the estimand: the result
is a probability-scale contrast rather than $\theta_0$.

For binary outcomes, the logistic partially linear model of @liu2021lplr
provides one remedy and has an official implementation in the DoubleML
software. Its orthogonal score builds in the logit link, confining the method
to binary responses under a logistic distribution. We are aware of no
corresponding DML estimator for ordinal latent-utility effects. Ordinal
latent-scale outcomes are therefore the clearest gap addressed here.
Under a logistic link, SUDO targets the same latent-scale coefficient as
@liu2021lplr through a different surrogate-completion construction; we do not
claim that the two estimators are identical or asymptotically equivalent.

Independently, @liu2018 and @cheng2021 developed *surrogate residuals* as a
diagnostic device for discrete-choice models. Grounded in random utility
theory [@mcfadden1974], a surrogate maps an observed category to a draw
from the conditional distribution of its latent utility. The result is a
continuous, homoscedastic residual with a known reference distribution
[@greenwell2018]. We repurpose that construction as an *estimation* device.
Substituting a surrogate $S$ for the unobserved $U$ turns the discrete
problem into a continuous one on which the FWL recipe runs unchanged. The
surrogate remains a draw rather than an observed outcome, so one completion
is not enough. We instead treat repeated completions as multiple
imputations and pool them with Rubin's rules [@rubin1987]. In the parametric
designs that compare draw schemes, imputing from the model's *posterior
predictive* distribution rather than a plug-in separates near-nominal
coverage from clear under-coverage.

**Contributions.**

1. A DML-assisted estimator for binary and ordinal latent-utility effects in
   one framework, not tied to the logit link, with a built-in
   specification diagnostic.
2. The finding that only the proper multiple-imputation draw scheme attained
   nominal coverage in the designs we study, quantified against naive and
   improper alternatives.
3. A closed-form analysis of how imputation-model error propagates to the
   effect, revealing a *first-order* sensitivity absent from standard DML,
   confirmed by a direct perturbation experiment.
4. Design-specific evidence for nominal coverage with machine-learning
   imputation models, using a partially-linear structure and a full-pipeline
   bootstrap, together with a proved deterministic-series reference learner.
5. An application to wine quality, illustrating an ordinal latent-scale
   estimand for which we are aware of no existing DML estimator.

# Setup and estimand

Let $D$ be a treatment, $X \in \mathbb{R}^p$ covariates, and $Y$ a
categorical outcome: binary, $Y = \mathbf{1}\{U > 0\}$, or ordinal,
$Y = j \iff \kappa_{j-1} < U \le \kappa_j$ for thresholds
$-\infty = \kappa_0 < \kappa_1 < \dots < \kappa_{J-1} < \kappa_J = \infty$
(we write $\kappa$ rather than $c$ to keep the thresholds distinct from the
pass-through factor $c(\cdot)$ of @sec-theory). The latent utility
follows a partially linear model

$$U = \theta_0 D + g_0(X) + \varepsilon, \qquad
\varepsilon \mid D,X \sim F$$ {#eq-latent}
with $g_0$ an arbitrary nuisance function and $F$ a known error distribution
(logistic, giving proportional-odds; extreme-value, giving
complementary-log-log; etc.). The target $\theta_0$ is the treatment effect
on the latent-utility scale, the same quantity a correctly specified
cumulative-link model targets, but with $g_0$ left nonparametric. It is not
the probability-scale average treatment effect; the two answer different
questions and should not be compared against a common truth. Because $F$
fixes the variance of $\varepsilon$, this scale is set by the link rather
than estimated, which is what gives $\theta_0$ its familiar reading: under
the logistic link a one-unit change in $D$ shifts the cumulative log-odds
of $Y$ by $\theta_0$, so $\exp(\theta_0)$ is a proportional-odds ratio.

Four assumptions identify $\theta_0$.

1. *Selection on observables.* No unobserved confounder acts beyond $X$.
   Together with the conditional location-family statement in @eq-latent,
   this assumption requires the latent error to have the same normalized law
   across treatment values given $X$.
2. *A constant latent-scale effect.* The partially linear form in
   @eq-latent makes $\theta_0$ a single number rather than a function of
   $X$.
3. *A correctly specified latent distribution.* The completion draws from
   $F$ using the assumed cutpoints, so the conditional law of
   $\varepsilon$, not merely its mean, must be correct. The surrogate-residual
   diagnostic monitors the fitted specification, of which this is one part,
   and the propagation analysis shows that the estimate is especially
   sensitive to its failure.
4. *Overlap.* Treatment retains support after adjustment, so its residual is
   not annihilated.

The first and fourth assumptions are standard causal conditions, made
explicit through the directed acyclic graphs (DAGs) for the simulations and
application. The second and third are modeling choices shared with any
cumulative-link analysis.

The inline example below and every simulation use the same causal DAG
(@fig-sim-dag). Their functional forms, treatment type, error law, and
cutpoints vary, but their causal arrows do not. The exogenous latent error
affects the utility but is independent of treatment and covariates.

```{mermaid}
%%| label: fig-sim-dag
%%| fig-width: 4.3
%%| fig-cap: "Causal DAG shared by the inline example and all simulation data-generating processes. The covariates may be a vector, and the observed category is a deterministic coarsening of the latent utility."
flowchart LR
  X[covariates X] --> D[treatment D]
  X --> U[latent utility U]
  D --> U
  E((latent error)):::u --> U
  U --> Y[observed category Y]
  classDef u fill:#eee,stroke-dasharray:4 3;
```

# Background: FWL, DML, and surrogate residuals

The FWL theorem states that, in a linear model, the coefficient on $D$
equals the coefficient from regressing the outcome residual on the treatment
residual after partialling $X$ out of both. DML makes this identity useful
with high-dimensional covariates by combining flexible estimates of the two
partialling regressions, $K$-fold cross-fitting, and an orthogonal score
[@chernozhukov2018].

Applying that calculation directly to a binary or ordinal label targets a
different quantity. A binary residual, for example, is bounded and
heteroscedastic, so its regression on the treatment residual estimates a
probability-scale contrast. SUDO instead uses the surrogate-residual
construction [@liu2018; @cheng2021] to complete the latent outcome before
FWL is applied. The completion preserves the continuous latent scale needed
to target $\theta_0$.

# Surrogate FWL: the idea in its simplest form

Fit an imputation model of the outcome on $(D, X)$ to obtain an estimated
latent index $\hat V_i$. For each
observation, draw a surrogate $S_i$ from $F$, centered at $\hat V_i$ and
truncated to the interval implied by the observed category:
$(0,\infty)$ when $Y_i = 1$, $(-\infty, 0]$ when $Y_i = 0$, or
$(\hat\kappa_{Y_i-1}, \hat\kappa_{Y_i}]$ in the ordinal case. When the
imputation model is correct and its parameters are known, this is the
conditional law of $U$ given $(D, X)$. Therefore,
$E[S \mid D, X] = E[U \mid D, X]$ (Theorem 1 of @cheng2021) and the
completed data satisfy

$$S = \theta_0 D + g_0(X) + \eta, \qquad E[\eta \mid D, X] = 0$$

with the same $g_0$ as in @eq-latent, so FWL recovers $\theta_0$. This is
the exact case. With an estimated index $\hat V$ the equality holds only
asymptotically under consistent estimation, and approximately otherwise;
the propagation analysis below makes the residual error precise. The
demonstration that follows estimates the index by maximum likelihood, so it
illustrates the corresponding large-sample regime.

The following self-contained example uses base R to draw a binary outcome
from a logistic latent model, fit the imputation model, and complete the
outcome to a surrogate. It compares three estimates of $\theta_0$: the
logistic-regression
coefficient on the treatment, the FWL estimate on the *true* (unobserved)
utility (the oracle), and the FWL estimate on the *surrogate* completion.

```{r}
#| label: surrogate-fwl-demo
#| echo: true
set.seed(1)
n <- 40000
theta <- 1.5
X <- rnorm(n)
D <- rbinom(n, 1, plogis(0.8 * X))  # treatment depends on a covariate
U <- theta * D + X + rlogis(n)      # latent utility, never observed
Y <- as.integer(U > 0)              # observed binary label

imputation_fit <- glm(Y ~ D + X, binomial)
index_hat <- predict(imputation_fit)
cdf_at_zero <- plogis(-index_hat)
uniform_draw <- ifelse(
  Y == 1,
  cdf_at_zero + runif(n) * (1 - cdf_at_zero),
  runif(n) * cdf_at_zero
)
surrogate <- index_hat + qlogis(
  pmin(pmax(uniform_draw, 1e-9), 1 - 1e-9)
)

treatment_residual <- resid(lm(D ~ X))
fwl <- function(outcome) {
  sum(treatment_residual * resid(lm(outcome ~ X))) /
    sum(treatment_residual^2)
}
round(c(
  glm = unname(coef(imputation_fit)["D"]),
  oracle = fwl(U),              # FWL on the true utility
  surrogate = fwl(surrogate)    # FWL on the completed utility
), 3)
```

All three estimates are close to $\theta_0 = 1.5$. Using only the binary
labels, the surrogate completion recovers the coefficient that FWL would
estimate if the latent utility were observed. Replacing the binary index
with an ordinal cumulative-link index and truncating to category-specific
intervals gives the ordinal construction. The remainder of the paper adds
two elements to this basic estimator: machine-learning partialling and
inference that accounts for the uncertainty hidden by a single completion.

# Surrogate-assisted double machine learning

**The estimator.** Cross-fit the two adjustment nuisances
$\hat m(X) \approx E[D \mid X]$ and $\hat\ell(X) \approx E[S \mid X]$, and
form the residual-on-residual estimate

$$\hat\theta = \frac{\sum_i (S_i - \hat\ell(X_i))(D_i - \hat m(X_i))}
{\sum_i (D_i - \hat m(X_i))^2}$$ {#eq-theta}

The $\hat\ell$ regression uses $X$ only: the index carried $D$'s signal
into $S$, and partialling on $X$ alone strips it back out so the residual
regression isolates $\theta_0$. The score
$\varphi = (S - \ell(X) - \theta(D - m(X)))(D - m(X))$ is Neyman-orthogonal in
$(\ell, m)$: its derivatives in both nuisance directions vanish by the
tower property, so first-order error in $\hat\ell$ or $\hat m$ does not bias
$\hat\theta$. This protection covers the two *adjustment*
nuisances but not the imputation model that produced $\hat V$; the
next section analyzes the difference.

**Proper multiple-imputation inference.** A single completion understates
uncertainty, so we draw $B$ surrogates and pool with Rubin's rules
[@rubin1987]: the estimate is $\bar\theta = B^{-1}\sum_b \hat\theta^{(b)}$
and the total variance is $T = \bar W + (1 + 1/B) B_{\!*}$, the mean
within-draw sandwich variance plus the inflated between-draw variance, with
a Barnard-Rubin small-sample $t$ reference [@barnard1999]. What matters is
*how* the draws differ. Each surrogate is centered at an estimated index
$\hat V = x'\hat\beta$; that estimate carries sampling error, and reusing
one fitted $\hat\beta$ across all draws (an *improper* draw) leaves the
shared error invisible to the between-draw variance, so $T$ comes up short.
A *proper* draw redraws the imputation model's parameters,
$\beta^{(b)} \sim N(\hat\beta, \widehat{\mathrm{Cov}}(\hat\beta))$
(thresholds included in the ordinal case), before each completion, the
large-sample normal form of Rubin's posterior-predictive requirement. @tbl-ladder shows the three
schemes moving coverage from 0.80 (naive) to 0.92 (improper) to 0.96-0.97
(proper).

$T$ must account for three things that vary from draw to draw: the sampling
noise in $\hat\theta$ once a completion is fixed (the within-draw $\bar W$),
the surrogate randomness in $S$, and the uncertainty in the imputation model
itself. Improper draws discard the last of these.

Under standard regularity, the normal parameter draw above describes the
same large-sample limit as the sampling distribution of $\hat\beta$, its
flat-prior posterior, and a parametric bootstrap. A proper draw is therefore
the large-sample form of a parametric bootstrap of the imputation model,
obtained from its covariance rather than by refitting. This approach
requires a model that reports a usable covariance, a requirement that fails
for the black-box learners considered below.

**Ordinal outcomes and the link.** The ordinal calculation is unchanged once a
cumulative-link model supplies the index and thresholds; because the
thresholds are parameters too, the proper draw perturbs them jointly with
the coefficients, which is why we fit with a model whose covariance covers
them. The error distribution $F$ remains an assumption. Under a correct
link, the surrogate residuals $R = S - \hat V$ are homoscedastic. A smooth
of $R^2$ on $X$, computed with an independent draw stream, can therefore
flag link misspecification through variance drift that mean-based checks
miss.

# How imputation-model error propagates

The orthogonality established above protects $\hat\theta$ from error in the
adjustment nuisances but says nothing about the imputation model, whose
error is baked into the completed data $S$ rather than differenced away.
That channel distinguishes SUDO from standard DML. The imputation model,
unlike the two adjustment nuisances, sits outside the orthogonality
guarantee, so the method is DML-assisted rather than fully orthogonal; its
overall score is not insensitive to every estimated component.

The argument requires the fitted index to be effectively free of an
observation's own outcome. Two estimation strategies provide that
separation. A *flexible* imputation model is cross-fit, making $\hat V_i$ an
out-of-fold prediction that is independent of $Y_i$ given $(D, X)$. A
*parametric* model is fit once on the full sample; its $\hat V_i$ depends on
$Y_i$ only at order $1/n$, which is asymptotically negligible. In our
experiments, cross-fitting the parametric model was harmful: per-fold index
noise inflated the between-draw variance, raising coverage to 1.00 while the
mean standard error reached 2.3 times the true standard deviation. The
binary machine-learning design therefore uses a cross-fit `gam` imputation
model, whereas the ordinal designs and wine application fit `clm` once.

Either way $\hat V$ is fixed with respect to the completion, while
$Y \mid D, X$ is generated by the truth. Since
$S$ is $\hat V$ plus truncated noise,

$$E[S \mid D, X] = \psi(\hat V, \eta), \quad
\psi(v, e) = v + F(e)\,\mu_+(v) + (1 - F(e))\,\mu_-(v)$$ {#eq-ess}

with $\mu_\pm$ the truncated-noise mean corrections. The correctly centered
draw is exact, $\psi(e, e) = e$ (recovering Theorem 1 of @cheng2021),
but index error passes through with a factor strictly below one,

$$c(e) = \left.\frac{\partial \psi}{\partial v}\right|_{v=e}
= 1 - f(-e)\,\bigl(\mu_+(e) - \mu_-(e)\bigr), \qquad f = F'$$ {#eq-ce}

Read $c$ as the fraction of index error that survives the truncation. With
no truncation the surrogate would be index plus noise and its mean would move
one-for-one with $\hat V$, giving $c = 1$; the observed category drags it
back, and $c$ is what is left.

Under a logistic link $c$ rises from $0.31$ at $e = 0$ to $1$ in the tails
(@fig-passthrough). The value at the origin is exactly $1 - \log 2$, because
for the logit
$$c(e) = 1 - H\bigl(F(e)\bigr), \qquad H(p) = -p\log p - (1-p)\log(1-p),$$ {#eq-centropy}
so the self-correction is precisely the entropy of the outcome. This identity
summarizes the mechanism. At $p = 1/2$, the label is maximally uncertain, so
observing it is most informative about where $U$ fell and the correction is
largest. As $p \to 1$, the label was already predictable, adds little
information beyond $\hat V$, and provides almost no correction. Strong
effects are therefore difficult for the same reason that an ordinal
category spanning many error standard deviations is difficult: in both
settings, the observed category conveys little additional information about
$U$.

The density in @eq-ce is evaluated at $-e$, which matters for an asymmetric
$F$ such as the extreme-value link; writing $f(e)$ is a shorthand valid only
when $F$ is symmetric. @sec-theory gives the general-$J$ form of both
@eq-ess and @eq-ce, where the ordinal thresholds contribute their own
derivative block. For the well-placed partitions studied here, adding
categories lowers the pass-through at $e = 0$: from $0.307$ for binary to
$0.157$ and $0.090$ for the displayed three- and four-category
configurations. Category count alone is not monotone, however; widely spaced
thresholds can instead drive the pass-through toward one.

```{r}
#| label: fig-passthrough
#| echo: false
#| fig-width: 5
#| fig-height: 2.6
#| fig-cap: "Pass-through factor $c(e)$ under a logistic link. Error in the imputation index reaches $\\hat\\theta$ multiplied by $c(e)$. The factor rises from about 0.31 at $e=0$ toward 1 as the signal grows because truncation self-corrects when $Y$ is informative but not when $Y$ is nearly deterministic. Averaging over the index distribution gives the $\\bar c_1$ reported in the text."
mu_plus <- function(index) {
  integrate(function(error) error * dlogis(error), -index, Inf)$value /
    plogis(index)
}
mu_minus <- function(index) {
  integrate(function(error) error * dlogis(error), -Inf, -index)$value /
    plogis(-index)
}
pass_through <- function(index) {
  1 - dlogis(index) * (mu_plus(index) - mu_minus(index))
}
index <- seq(0, 6, by = 0.05)
pass_through_value <- vapply(index, pass_through, numeric(1))
par(mar = c(4, 4, 1, 1), cex = 0.9)
plot(
  index, pass_through_value,
  type = "l", lwd = 2, ylim = c(0, 1), las = 1,
  xlab = "index level e",
  ylab = expression("pass-through factor " * c(e))
)
abline(h = 1, lty = 3, col = "grey60")
origin_value <- pass_through(0)
points(0, origin_value, pch = 19)
text(
  0.1, origin_value + 0.1, sprintf("%.2f", origin_value),
  pos = 4, cex = 0.9
)
```

Linearizing $\psi$ around the truth, the bias of $\hat\theta$ is
approximately $\bar c_1\, \delta_D$ plus a covariate-direction leak given
below, where $\delta_D$ is the imputation model's error in its effective
treatment coefficient and $\bar c_1 = E[w\, c(\theta_0 + g)] / E[w]$,
$w = m(1-m)$.

**Self-correction weakens with signal.** Because $Y$ comes from the truth,
truncation pulls $S$ toward the correct interval, giving $c < 1$. As the
signal grows and $Y$ becomes nearly deterministic, however, it adds no
information beyond $\hat V$ and $c \to 1$. For a *fixed* index
misspecification, this loss of self-correction is what makes bias hardest to
control at strong effect sizes: less of the index error is damped away. It is
not the whole story when the imputation model is correctly specified, where
@sec-theory shows the variance components themselves move against us. Nor
does it license the reverse reading, since $c \to 1$ is also the regime in
which the linearization of $\psi$ is most accurate.

**Covariate-direction error can survive partialling.** The opposite might
seem natural because residualized treatment is orthogonal to any function of
$X$. To see why it fails here, split the imputation
model's index error into a treatment part and a covariate part,
$\hat V - \eta = \delta_D D + \delta_X(X)$. Partialling-out regresses on
$R = D - m_0(X)$, and $E[R \mid X] = 0$ annihilates any function of $X$, so
$\delta_X$ ought to wash out. It does not, because $S$ is nonlinear in
$\hat V$: linearizing multiplies the index error by $c(\eta)$, and $\eta$
depends on $D$, so $\delta_X$ arrives multiplied by something that itself
varies with treatment and therefore survives the projection. For binary $D$,
writing $\eta_d = \theta_0 d + g_0(X)$ and $w = \mathrm{Var}(D \mid X)$, the
surviving term is
$$\text{covariate-direction leak} \;=\;
\frac{E\bigl[w\,\bigl(c(\eta_1) - c(\eta_0)\bigr)\,\delta_X\bigr]}{E[w]} .$$ {#eq-leak}
Proposition A4 gives the exact path-integrated version of this first-order
form and a weighted norm bound. The leak is governed by *how differently
the truncation damps the two treatment arms*. Were the two path-averaged
factors the same, it would vanish exactly and the naive argument would be
right.

The leak vanishes when the imputation model can represent $g_0$, which is
why designs with every model correctly specified show no
covariate-direction bias. When it cannot, two channels can appear.
Giving $g_0$ an $X_1X_2$ interaction that an additive logistic imputation
model cannot express produces a bias of 5 to 6% of $\theta_0$ and coverage
of 0.75 at strong signal. Although treatment is constrained to enter
linearly, omitted logistic heterogeneity shifts its pseudo-true coefficient,
so this experiment is not a pure $\delta_D=0$ case. The population
decomposition in @sec-numverify separates the common covariate leak from
that induced treatment-direction error. Refitting the same design with an
imputation model that admits the interaction removes the bias, confirming
the imputation model rather than the adjustment nuisances as the source.

**Imputation-model error is first order.** In the sharpest contrast with
standard DML, the imputation model's error enters
$\hat\theta$ *linearly*. In DML the outcome is observed and nuisance error
enters only through a *product* of two errors [@chernozhukov2018;
@robinson1988]: writing $\ell = \ell_0 + a$ and $m = m_0 + b$ for the errors
in the two adjustment models, the remaining bias is
$E[ab] - \theta_0 E[b^2]$, which is second order in the pair. Getting the
treatment model right ($b = 0$) removes it entirely; getting only the
outcome model right does not. The score is Neyman-orthogonal, first-order
insensitive to nuisance error, rather than doubly robust in the usual sense.
SUDO adds a third model whose treatment-direction error is unprotected and
first-order, damped only by $\bar c_1 < 1$. @tbl-sensitivity confirms this
numerically: a perturbation applied to the outcome nuisance in DML leaves
the estimate unbiased at every level, whereas the same-sized index
perturbation in SUDO produces a linear bias with slope exactly $\bar c_1$.

# Machine-learning imputation models

@eq-ess reverses the usual DML emphasis: imputation-model quality matters
more than adjustment-model quality. A well-specified parametric imputation
model (logistic or cumulative-link)
needs no extra machinery, since its covariance supplies the proper draw
directly. Such draws propagate parameter uncertainty and attain near-nominal
coverage in our parametric experiments; @sec-theory characterizes the small
Rubin-variance discrepancy that remains. A black-box learner introduces two
obstacles: it can carry treatment-direction error $\delta_D$, creating bias,
and it has no covariance matrix from which to draw, complicating inference.

The bias is addressed by a *partially-linear* (PL) imputation model,
$\hat V = \hat\alpha D + \hat f(X)$, in which the treatment enters linearly
and machine learning handles only $X$, so no treatment-by-covariate
artifact can arise. This restriction blocks one important source of
treatment-direction error, but it does not guarantee that $\hat f$ is
accurate or that the learner satisfies the projected expansion needed for
inference. It defines a structural learner class, not a commitment to one
training algorithm. We implement it with a generalized additive model
(GAM), multivariate adaptive regression spline (MARS) bases followed by a
joint logistic fit, and neural-network backfitting. All three adaptive
learners meet the Monte Carlo-aware bias criterion across effect sizes in
the common comparison in @sec-sims. We also state the corresponding
results for a deterministic series dictionary: Theorem S1 gives its projected
first-order expansion and Theorem S2 its bootstrap stability, both under
explicit series conditions. The fixed degree-three Hermite learner we run is
a numerical reference for those results rather than an instance of them, for
the reason given in @sec-theory, and it is not the practical default.

For inference, we use a *full-pipeline bootstrap*, the nonparametric
counterpart of the proper draw. A parametric model injects imputation-model
uncertainty by sampling $\beta$ from its covariance. Here we instead
resample the whole data set and rerun the entire procedure, cross-fitting
and all, using the spread of the estimates as the standard error. Both
approaches target the same variance component; the bootstrap obtains it by
refitting when no covariance is available. It has to wrap the *whole* pipeline,
because a within-fold refit of the model alone under-propagates that same
imputation-model variability. In the designs studied, normal intervals using
this full-pipeline bootstrap attain nominal coverage for all three adaptive
learners (@sec-sims).

Implication M2 states a learner-class bootstrap result under a projected
bootstrap-stability condition and conditional nuisance rates. Under those
conditions, fixed and redrawn regular folds are both admissible, and only the
integrated generated-outcome map must be differentiable, not every raw
inverse-transform draw. Implication M2 also identifies a fixed-imputation
centering issue. Fresh surrogate randomness makes the bootstrap variance
consistent, but it does not justify the usual distribution centered at one
realized randomized estimate. The supported procedure is therefore the
normal interval based on the full-pipeline bootstrap standard deviation.

Learner-specific verification of the projected stability condition remains
open for the three adaptive algorithms. The deterministic series result
closes the proof route for one tractable partially-linear learner; the
adaptive-learner comparison below supports, but does not prove, the same
conclusion for GAM, MARS, or neural backfitting.

A Bayesian ensemble offers a direct alternative because each posterior draw
is a complete imputation model and therefore a proper draw. Under a probit
imputation model, the Albert-Chib augmentation variable is the surrogate
completion, so each Gibbs sweep supplies one. In the probit analogue we
studied, however, the approach did not improve performance. Bayesian
additive regression tree (BART) imputation, both unconstrained and under the
partially-linear restriction above, had bias that grew with signal strength
to about 5% of $\theta_0$, where the backfit meets the bias criterion. Its
posterior draws propagated variance no better than the bootstrap they were
intended to replace. This result comes from one design and is not a general
verdict. The run is in the repository. A forest's out-of-bag or
infinitesimal-jackknife variance [@wager2014] is not such a route at all: it
quantifies prediction uncertainty at a point, not a coherent joint draw of
the whole imputation model.

# Simulation study {#sec-sims}

Every simulation data-generating process (DGP) follows the causal structure
in @fig-sim-dag. Independently and identically distributed (iid) covariates
determine treatment assignment; treatment and covariates then determine a
latent utility, which is coarsened into the observed category:

$$X_1, \dots, X_p \stackrel{\text{iid}}{\sim} N(0, 1), \qquad
D \mid X \sim \mathrm{Bernoulli}\bigl(\Lambda(\pi(X))\bigr), \qquad
U = \theta_0 D + g_0(X) + \varepsilon,$$ {#eq-dgp}
$$\varepsilon \sim F, \qquad
Y = j \iff \kappa_{j-1} < U \le \kappa_j,$$

where $\Lambda$ is the logistic cumulative distribution function. Choosing
$(p, \pi, g_0, F, \kappa, \theta_0, n)$ then fixes a design. We use two
covariate structures throughout. The **one-covariate setting (S1)** has
$\pi(X) = 0.8X$ and $g_0(X) = X$; every fitted model is correctly specified,
so only pooling is under test. The **five-covariate machine-learning setting
(S5)** has
$$\pi(X) = X_4 + \cos X_5, \qquad g_0(X) = X_1^2 + \sin X_2 + 0.5 X_3,$$
which is nonlinear in $X$ and makes the treatment depend on covariates that
do not enter $g_0$. Binary outcomes take $\kappa = (0)$; ordinal ones cut
$U$ at two interior thresholds into $J = 3$ categories. @tbl-dgp in the
appendix lists every design's parameters.

The coverage target is $0.95$, and Monte Carlo error for a reported coverage
rate is about $\pm 0.02$. **Standard deviation (SD)** denotes the
estimator's run-to-run variability; **mean standard error (mean SE)** is the
average uncertainty it reports. A mean SE below the SD signals intervals
that are too short. @sec-appendix gives learner and tuning
specifications, fold and draw counts, and per-cell Monte Carlo standard
errors. The tables below are generated at render time from the committed
result files.

```{r}
#| label: results-setup
#| include: false
results_dir <- "../../R/results"
read_result <- function(filename) {
  read.csv(file.path(results_dir, filename), check.names = FALSE)
}
sg <- function(x, d = 3) sprintf(paste0("%+.", d, "f"), x)  # signed
nn <- function(x, d = 3) sprintf(paste0("%.", d, "f"), x)   # unsigned
```

**Proper draws are what deliver coverage.** With a simple binary design and
a correct parametric imputation model, only the proper scheme reaches
nominal coverage; the naive interval covers 0.80 and the improper 0.92
(@tbl-ladder).

```{r}
#| label: tbl-ladder
#| echo: false
#| tbl-cap: "Inference schemes, binary outcome, $n = 5000$, $\\theta_0 = 1.5$, 500 replications. $B$ is the number of surrogate draws."
s2 <- read_result("stage2_summary.csv")
estimators <- c("naive", "improper_B25", "proper_B25")
result <- s2[match(estimators, s2$estimator), ]
table_data <- data.frame(
  Scheme     = c("Naive single draw", "Improper, $B=25$", "Proper, $B=25$"),
  Bias       = sg(result$bias),
  SD         = nn(result$sd),
  `Mean SE`  = nn(result$mean_se),
  Coverage   = nn(result$coverage),
  check.names = FALSE)
knitr::kable(table_data, align = c("l", "r", "r", "r", "r"))
```

**Ordinal outcomes.** With a three-category outcome and machine-learning
adjustment, SUDO meets its bias criterion with valid-to-conservative
coverage (@tbl-ordinal), a regime for which we are aware of no DML
competitor for the latent-scale coefficient. The original implementation
was substantially conservative. A term-level ordinal decomposition includes
the threshold block and matches Monte Carlo throughout a clean
cumulative-logit testbed. The decomposition predicts only
2.8% variance overstatement at moderate signal and 2.0% understatement at
strong signal, so threshold estimation and ordinal pooling alone cannot
explain the 50% to 88% variance gap after flexible partialling nuisances are
introduced. A paired nuisance-isolation experiment locates the gap in the
fixed outcome nuisance: refitting $E[S\mid X]$ for every ordinal completion
reduces mean $T$ by 38% while leaving the empirical sampling variance
unchanged, moving $T/V$ from 1.505 to 0.942. The full corrected stage at
$B=25$ and 200 replications has bias 0.016, coverage 0.960, and $T/V=1.214$
with Monte Carlo standard error 0.122. Per-draw refitting is therefore the
validated ordinal procedure, at the cost of fitting the outcome nuisance
$B$ times rather than once.
The isolation experiment and the full stage report $T/V$ on opposite sides
of one, 0.942 and 1.214, but each is within two Monte Carlo standard errors
of one at its own resolution, so they do not conflict and neither is
evidence of a remaining ordinal anomaly. Rubin's
variance is guaranteed to err safe only under conditions we have not
verified here (a valid imputation model together with a self-efficient
complete-data analysis), and under uncongeniality it can be biased in
either direction [@meng1994; @xie2017]. The over-coverage is a property of
these designs, not a general guarantee.

```{r}
#| label: tbl-ordinal
#| echo: false
#| tbl-cap: "Ordinal outcome, $J = 3$, $\\theta_0 = 1.0$. Parametric arm 500 replications; corrected spline + ML arm 200, with the outcome nuisance refit per completion."
s4 <- read_result("stage4_summary.csv")
s5 <- read_result("stage5r_summary.csv")
parametric_result <- s4[s4$estimator == "sudo_ordinal_clm", ]
refit_result <- s5[s5$estimator == "ordinal_refit_n2000_B25", ]
table_data <- data.frame(
  Configuration = c("Parametric, $n=3000$", "Spline + ML nuisances, $n=2000$"),
  Bias      = sg(c(parametric_result$bias, refit_result$bias)),
  SD        = nn(c(parametric_result$sd, refit_result$sd)),
  `Mean SE` = nn(c(parametric_result$mean_se, refit_result$mean_se)),
  Coverage  = nn(c(parametric_result$coverage, refit_result$coverage)),
  check.names = FALSE)
knitr::kable(table_data, align = c("l", "r", "r", "r", "r"))
```

**Link misspecification.** Generating Gumbel noise but assuming a logistic
link substantially biases the latent-scale effect (+0.77 for binary, +0.58
for ordinal outcomes, with zero coverage). The correct link removes this
bias (bias $<0.005$, with valid-to-conservative coverage of 0.980 and
0.975). A surrogate-residual check behaves as expected: the conditional
second moment drifts with $X$ under the misspecified fit (nominal
$p \approx 2\times 10^{-4}$) and not under the correct one
($p \approx 0.46$). Under a correctly specified model the surrogate residual
has constant conditional mean and variance [@liu2018], so drift is evidence
against the fitted specification. It is not specific to the link, since
scale heterogeneity, omitted mean structure, or non-proportionality would
also produce drift, and the reported values aggregate over surrogate draws
and are not calibrated for parameter estimation. We therefore use this as a
diagnostic rather than a test.

**First-order versus second-order sensitivity.** @tbl-sensitivity
traces the bias after perturbing one input by a known function. Standard DML
with the observed outcome remains unbiased under an arbitrarily wrong
outcome nuisance when the propensity is correct. SUDO's index error instead
enters linearly: its treatment-direction slope matches the derived
$\bar c_1$ (0.669 at $\theta_0 = 1.5$, 0.861 at $\theta_0 = 3$) to within
0.001.

| Arm | Perturbed | Bias vs. perturbation |
|---|---|---|
| DML | outcome nuisance only | flat, $\approx 0$ |
| DML | both nuisances | quadratic |
| SUDO | index, treatment direction | linear, slope $=\bar c_1$ |

: Nuisance-sensitivity ladder. {#tbl-sensitivity}

**Partially-linear imputation learners.** A tuned partially-linear backfit
meets the bias criterion at both effect sizes. Pairing it with the
full-pipeline bootstrap brings strong-effect normal coverage from 0.86 to
0.98; the corresponding percentile coverage is 0.95. The normal interval is
the inferential procedure supported by Implication M2 under its stated
conditions. The percentile result is retained as an empirical comparison
from the original validation. With 100 replications, either coverage
estimate has Monte Carlo uncertainty of about $\pm 0.03$; the bootstrap SD
also matches the estimator's run-to-run variability at this resolution.

Backfitting is not essential. A common-sample comparison fits the same
$\alpha D + f(X)$ structure three ways: a GAM with a linear treatment term,
MARS bases constructed from $X$ only followed by a joint logistic fit with
mandatory linear $D$, and the tuned neural-network backfit. The MARS
configuration is selected on theta-level Monte Carlo bias, not predictive
generalized cross-validation (GCV). The three adaptive learners pass the
original prespecified bias and percentile-coverage criteria
(@tbl-pl-learners). Their normal coverage, which is the recommended
inferential comparison after the fixed-$B$ centering analysis, ranges from
0.94 to 0.98. With 50 replications, these coverage cells have Monte Carlo
standard errors of roughly 0.02 to 0.05, so their small differences do not
rank the learners.

```{r}
#| label: tbl-pl-learners
#| echo: false
#| tbl-cap: "Partially-linear imputation learners with full-pipeline bootstrap, $n=2000$, 50 replications, 50 outer resamples, and 15 surrogate completions per resample."
s3s <- read_result("stage3s_summary.csv")
learner_name <- c(
  pl_gam = "GAM",
  pl_mars = "MARS",
  pl_backfit = "Neural backfit"
)
table_data <- data.frame(
  Learner = unname(learner_name[s3s$learner]),
  `$\\theta_0$` = nn(s3s$theta, 1),
  Bias = sg(s3s$bias),
  `SD/Mean SE` = nn(s3s$sd_over_mean_se),
  `Normal coverage` = nn(s3s$coverage_normal),
  `Percentile coverage` = nn(s3s$coverage_percentile),
  check.names = FALSE
)
knitr::kable(table_data, align = c("l", "r", "r", "r", "r", "r"))
```

**Partial linearity alone is not sufficient.** Two further partially-linear
arms failed the bias criterion, both at $n = 2000$ with 100 replications. An
interaction-constrained gradient-boosted index, left deliberately under-tuned
to test whether the structural constraint protects a poorly fitted index on
its own, had bias $-0.09$ at $\theta_0 = 1.5$ and $-0.16$ at $\theta_0 = 3$.
A componentwise-boosted index whose stopping iteration was chosen by
cross-validated predictive risk was worse, at $-0.17$ and $-0.44$. The second
result is the informative one: predictive risk is not the objective that
governs first-order target bias, so a rule that is principled for prediction
need not control it. We ran no fixed-stopping comparator, so this does not
show that the cross-validated rule made that learner worse than an untuned
one. Both arms were scored with the recentered within-fold bootstrap rather
than the full-pipeline bootstrap, so their coverage (0.85 and 0.73; 0.70 and
0.11) is not comparable to @tbl-pl-learners and indicates only the scale of
the failure. Partial linearity excludes treatment-by-covariate artifacts; it
does not remove error in the treatment coefficient or in the covariate
component.

**How the learners were selected.** The MARS and neural-backfit
configurations were screened on target-level Monte Carlo bias against a known
$\theta_0$, then evaluated in separate runs. This shows that the selected
configurations met the prespecified bias and coverage criteria in these
designs. It is not a selection rule available to an analyst working from
data. A criterion
computable without $\theta_0$, together with verification that (M2) and (MB1)
survive the selection, remains open.

The bootstrap comparison uses paired generated datasets and reruns each
learner with either one fixed regular fold partition or folds redrawn in
every resample. @tbl-ml-bootstrap reports the direct comparison. The
within-fold method refits only the imputation learner inside the original
training folds, while the improper method holds its fitted index fixed.
Their SE ratios therefore locate which layer of variation is omitted. The
normal interval is the procedure supported by Implication M2 under its high-level
conditions; percentile and conventionally centered studentized intervals
are included as fixed-$B$ empirical benchmarks.

```{r}
#| label: tbl-ml-bootstrap
#| echo: false
#| tbl-cap: "Paired full-pipeline bootstrap comparison, $n=2000$, 30 replications, 15 outer resamples, and 15 surrogate completions per resample."
sboot <- read_result("theory_ml_bootstrap.csv")
table_data <- data.frame(
  Learner = unname(learner_name[sboot$learner]),
  `$\\theta_0$` = nn(sboot$theta, 1),
  `Fixed / redrawn SE` = nn(sboot$fixed_over_redrawn_se),
  `Within / full SE` = nn(sboot$within_over_full_se),
  `Improper / within SE` = nn(sboot$improper_over_within_se),
  `Normal coverage` = nn(sboot$coverage_redrawn_normal),
  `Percentile coverage` = nn(sboot$coverage_redrawn_percentile),
  `Studentized coverage` = nn(sboot$coverage_redrawn_student),
  check.names = FALSE
)
knitr::kable(table_data, align = c("l", rep("r", 7)))
```

**Continuous treatment, and other variations.** The designs above all use a
binary treatment, while the application below does not, so we check the
transfer directly. Replacing binary $D$ with $D = 0.7\pi(X) + N(0,1)$ in the
S5 structure leaves the estimator unbiased for both a binary outcome
($-0.1\%$ at $\theta_0 = 1$, $+0.4\%$ at $2.5$) and a three-category ordinal
one ($+0.7\%$ and $+1.5\%$), so the latent-scale estimand and the machinery
carry over. Doubling the covariate count to ten, and scaling the
treatment-assignment index so that overlap is markedly worse, likewise cost
nothing in bias. Bias does appear when the imputation model cannot represent
$g_0$, as described above. Coverage under a continuous treatment tracks the
binary case at moderate signal but is worse at strong signal, where the
reported SE falls furthest short of the truth. The full-pipeline bootstrap
is the remedy there, as it is for the black-box case. Details and the full
grid are in @tbl-dgp and the repository.

# Application: volatile acidity and wine quality

We estimate the effect of volatile acidity, the acetic-acid fault commonly
described as a vinegar or nail-polish aroma, on wine quality. In the
University of California, Irvine (UCI) wine-quality data [@cortez2009],
quality is the median ordinal score from blind tasters. The data contain
1,599 red and 4,898 white wines, with quality spanning six to seven levels.

Because these data are observational, causal interpretation depends on the
assumed application DAG in @fig-dag. Volatile acidity is a plausible
treatment: tasters are expected to penalize the acetic fault, and the
variable is a fault endpoint rather than an apparent cause of the remaining
chemistry. Even so, the interpretation requires the other physicochemical
measurements to be pre-treatment adjustments rather than mediators or
proxies of volatile acidity. It also requires winemaking practice to have no
direct effect on quality beyond measured chemistry, the exclusion
restriction encoded by the graph.

We do not claim that these assumptions are established. Instead, we examine
the two most questionable features through adjustment-set robustness and an
unmeasured-confounding calibration. The application illustrates the method
on a genuinely ordinal outcome; it is not a settled causal finding.

```{mermaid}
%%| label: fig-dag
%%| fig-width: 3.2
%%| fig-cap: "Assumed application DAG. Volatile acidity (the treatment) affects latent quality directly; unobserved winemaking practice confounds both through the measured chemistry, the adjustment set."
flowchart TD
  P((winemaking<br/>practice)):::u --> VA[volatile acidity]
  P --> X[other chemistry]
  X --> VA
  VA --> Q[latent quality]
  X --> Q
  classDef u fill:#eee,stroke-dasharray:4 3;
```

Volatile acidity retains adequate residual variation after adjustment: 75%
of its standard deviation for red wine and 94% for white wine. These values
are residual $\sqrt{1 - R^2}$ computed from the $R^2$ column of @tbl-wine.
The treatment residual is therefore not annihilated, and the
residual-on-residual regression is well posed.

Ordinal SUDO, using a cumulative-link imputation model, proper draws,
cross-fitted nuisances, and Rubin pooling, gives a negative effect on latent
quality. The estimate is stable across linear and partially-linear spline
imputation models (@tbl-wine). As a model-based comparator, we also report
the treatment coefficient from the cumulative-link fit itself. Its standard
error is smaller (white: 0.031
versus 0.064-0.065) because it relies directly on the full parametric
outcome model, whereas SUDO leaves the partialling regressions flexible and
propagates the completion and cross-fitting variability. The narrower
model-based interval can be efficient when its stronger functional-form
assumptions are correct; it is not evidence that SUDO's additional
uncertainty is spurious.

Volatile acidity is standardized before fitting, so these coefficients are
per one standard deviation (about $0.18$ g/dm$^3$ for red, $0.10$ for
white), which keeps the effect comparable across the two wines instead of
tied to the raw acetic-acid scale. The proportional-odds reading then makes
the magnitude concrete: a one-SD rise in volatile acidity multiplies the
odds of a wine scoring above any fixed quality level by $\exp(\hat\theta)$,
about $0.59$ for red ($\hat\theta = -0.52$) and $0.64$ for white
($\hat\theta = -0.45$), a reduction of roughly 36 to 40% in those odds per
standard deviation.

```{r}
#| label: tbl-wine
#| echo: false
#| tbl-cap: "Effect of a one-SD increase in volatile acidity on latent wine quality. Model-based coefficient from the linear cumulative-link fit (standard error in parentheses); SUDO estimates with 95% intervals in brackets."
wine_results <- read_result("wine_application.csv")
wine_row <- function(wine_name, label) {
  linear <- wine_results[
    wine_results$wine == wine_name & wine_results$full_model == "linear",
  ]
  spline <- wine_results[
    wine_results$wine == wine_name & wine_results$full_model == "pl_spline",
  ]
  data.frame(
    Wine = label,
    `Overlap $R^2(D \\sim X)$` = nn(linear$overlap_r2, 2),
    `Cumulative-link` = sprintf(
      "$%.2f$ (%.2f)", linear$naive_theta, linear$naive_se
    ),
    `SUDO (linear)` = sprintf(
      "$%.2f$ [$%.2f,%.2f$]",
      linear$sudo_theta, linear$ci_lo, linear$ci_hi
    ),
    `SUDO (spline)` = sprintf(
      "$%.2f$ [$%.2f,%.2f$]",
      spline$sudo_theta, spline$ci_lo, spline$ci_hi
    ),
    check.names = FALSE)
}
table_data <- rbind(wine_row("red", "Red"), wine_row("white", "White"))
knitr::kable(table_data, align = c("l", "r", "r", "r", "r"))
```

The negative sign is stable across the reported sensitivity checks. Dropping
any single covariate from the adjustment set leaves the effect in
$[-0.68, -0.47]$ for red and $[-0.52, -0.42]$ for
white, every interval excluding zero, so the estimate does not hinge on one
adjustment choice. Read through the omitted-variable-bias framework
[@cinelli2020], an unobserved confounder would have to explain about 15%
(red) and 10% (white) of the residual variation in both volatile acidity and
latent quality to bring the effect to zero; weaker confounding leaves it
negative. We report these robustness values as a heuristic calibration
rather than a formal bound: the Cinelli-Hazlett formula is derived for
omitted-variable bias in linear regression with an observed outcome, and we
are applying it to a pooled, surrogate-completed estimate whose standard
error comes from Rubin's rules. Its validity in that setting has not been
established. These values and the leave-one-out ranges are in the committed
result files.

**How wrong could the imputation model be?** Unmeasured confounding is the
threat any observational analysis faces; the one specific to SUDO is error in
the imputation model, which @eq-leak and @eq-ce show reaches $\hat\theta$ at
first order. That channel admits its own sensitivity statement, and unlike
the robustness value above it follows directly from the estimator's own
theory rather than being imported. Since the bias is $\bar c_1 \delta_D$, the
imputation model's error in the volatile-acidity coefficient that would drive
the estimate to zero is $\delta_D = -\hat\theta / \bar c_1$. Evaluating
$\bar c_1$ at the fitted index with the continuous-treatment multiplier
$E[c(\eta) D R]/E[R^2]$ and the general-$J$ pass-through of Proposition A1
gives $\bar c_1 = 0.21$ (red) and $0.17$ (white). The cumulative-link model
would therefore have to be wrong about volatile acidity by $+2.51$ and
$+2.65$ respectively, which is 4.1 and 5.3 times the coefficient it actually
fits. This calculation addresses treatment-direction error only. Conditional
on covariate-direction misspecification being negligible, it shows that a
large treatment-coefficient error would be required to erase the estimate;
it does not establish robustness to general imputation-model error.

Two features of the design produce this result. First, the pass-through is
near $0.2$, rather than the $0.67$ to $0.86$ found in the binary
simulations. Six or seven categories with well-spaced thresholds constrain
the latent utility tightly, allowing truncation to provide substantial
self-correction. Second, the estimate is far from zero relative to its
scale. Both protections weaken for a coarser outcome or a smaller effect,
precisely the settings in which this sensitivity calculation becomes most
informative.

# Discussion

SUDO uses outcome completion to extend partially linear DML to binary and,
to our knowledge for the first time, ordinal latent-scale effects. Other DML
scores can target probability- or distribution-scale effects for
categorical outcomes; SUDO targets the coefficient of a latent-utility
model. Under the logit link it targets
the same latent-scale coefficient as the binary-logistic DML estimator of
@liu2021lplr, while admitting other error distributions and ordinal as well
as binary responses. The methodological core is inferential: the completions are multiple
imputations, and in our experiments only proper draws, which redraw the
imputation model's parameters (thresholds included), attain nominal
coverage. The method is link-flexible over any distribution for which the
truncated draw and reference are available, and its residuals give a
diagnostic that can flag a misspecified fit, though it neither confirms the
specification nor isolates the link as the cause.

The propagation analysis makes the main tradeoff explicit. The imputation
model sits outside the orthogonality guarantee, so its error reaches the
effect at first order through a truncation factor that increases as the
signal strengthens and the outcome's self-correction fades. Standard DML
nuisance error is instead second order. In the designs studied, this
sensitivity is manageable: well-specified parametric models with proper
draws and the three validated partially-linear learners with a
full-pipeline bootstrap attain nominal coverage. Partial linearity alone
does not provide this guarantee. The imputation learner must still estimate
the covariate component well and satisfy the projected stability conditions
needed for inference, and two partially-linear arms we tried did not. The
MARS and neural-backfit configurations were themselves selected against a
known $\theta_0$, so supplying a selection criterion an analyst could
compute from data remains open.

Several problems remain open. @sec-theory separates the proved results from
the conjectured ones. Theorem S1 proves the projected expansion for a
deterministic logistic-series learner, and Theorem S2 verifies its
full-pipeline bootstrap stability under explicit series and conditional
bootstrap conditions. The broader adaptive-learner statements are
Implications M1 and M2, which hold under high-level learner conditions that
are stated but not minimized. What remains conjectural is that GAM, MARS,
and neural backfitting satisfy those conditions, the projected
asymptotic-linearity requirement in particular. SUDO therefore has
a proved projected expansion and bootstrap result for the theorem-reference
specialization, while inference with the adaptive learners remains
simulation-supported. Natural next steps are to extend the adaptive results
beyond the data-generating structures studied and to develop a method for
unordered (multinomial) outcomes, which require a vector surrogate and have
no single scalar effect. A Bayesian ensemble could replace the bootstrap
because its posterior draws are literal proper imputations. A BART version
did not improve on the bootstrap in the one design we tried, so the
conditions under which that route succeeds remain open.

Two narrower theoretical questions also remain. Conjecture A2 leaves open when
$G'C = \sigma_u^2$: the identity fails in both directions in the validated
testbeds, though the resulting 1% to 2% discrepancy is smaller than the
fixed-$B$ chi-square variation at ordinary imputation counts. Strong-signal
theory separately needs primitive conditions for dominated differentiability
of the integrated completion map and for excluding asymptotic separation as
category probabilities approach zero or one.

# Data and software

The wine-quality data come from the University of California, Irvine Machine
Learning Repository [@cortez2009]. The `sudo` Python package provides the
parametric estimator for binary and ordinal outcomes, including proper draws
and the per-completion ordinal outcome-nuisance refit. The partially-linear
machine-learning imputation learners and the full-pipeline bootstrap are
implemented in R, which is also where every method is prototyped and
validated before release. The open-source project repository contains the code and committed
result files behind every table and figure under the Massachusetts Institute
of Technology (MIT) License:
<https://github.com/bgreenwell/sudo>.

# Declaration of artificial intelligence use

We used generative artificial intelligence (AI) to assist with code and
summaries of reference materials. The authors reviewed and validated all
outputs, retain full ownership of the final manuscript, and accept sole
responsibility for any errors.

# Appendix: asymptotic theory {#sec-theory}

This appendix develops four theoretical components used in the body: the
pass-through map and its derivatives for general $J$; the moment properties
that separate adjustment nuisances from the imputation model; the
fixed-$B$ asymptotic distribution of the pooled estimate; and the resulting
account of Rubin's variance.

The labels distinguish the status of each result. Direct differentiation,
conditioning, or inequalities establish the propositions. Conjectures A1 and
A2 are labeled as such because we give heuristic derivations and numerical
support rather than proofs. Implications M1 and M2 are conditional
calculations: they follow from high-level learner conditions, and what
remains conjectural is whether the implemented adaptive learners satisfy
those conditions, not the displayed conclusions given them. The
deterministic-series specialization is stated as a theorem because explicit
series and conditional-bootstrap conditions are given and the arguments are
carried out under them. The learner-specific projected asymptotic-linearity
and bootstrap-stability conditions remain unproved for the implemented
adaptive learners. The
project repository checks every closed-form expression numerically, using
numerical differentiation for the pass-through derivatives and Monte Carlo
for the variance decomposition.

## Notation and assumptions

Let the imputation model be a parametric family
$P_\beta(Y \mid D, X)$ with latent index $\eta(D, X; \beta)$. For a
generalized linear model (GLM) with known cutpoints, $\eta = x'\beta$. When
a cumulative-link model estimates the cutpoints, we collect them and the
index coefficients in $\beta = (\beta_\eta, \kappa)$. The thresholds are
components of $\beta$, but not of $\eta$; they enter only through the
truncation interval. This distinction matters in @eq-gblocks. Write
$\hat\beta$ for the maximum-likelihood estimate,
$m_0(X) = E[D \mid X]$, $R = D - m_0(X)$, and $J_\theta = E[R^2]$.

Given $\beta$, a completion is $S(W; \beta) = \eta(D, X; \beta) + \xi$ with
$\xi$ drawn from $F$ truncated to the interval $Y$ implies. The draw noise
splits into a data-measurable part and a pure Monte Carlo part,

$$\xi_i^{(b)} = e_i + u_i^{(b)}, \qquad e_i = E[\xi_i \mid W_i; \beta],
\qquad E[u_i^{(b)} \mid W_i] = 0 ,$$ {#eq-split}

This split determines the fixed-$B$ variance decomposition.
Note that $\xi$ is *not* mean-zero given $W$: the truncation is selected by
the observed $Y$. Only after averaging over $Y \mid D, X$ at the true model
does the mean vanish, because composing $Y \sim P_{\beta_0}$ with the
matching truncated draw reproduces the untruncated law, so
$\xi \mid D, X \sim F$ exactly. Define

$$\sigma_e^2 = E[e^2 R^2], \quad
\sigma_u^2 = E[\operatorname{Var}(\xi \mid W) R^2], \quad
\sigma_\varphi^2 = E[\xi^2 R^2] = \sigma_e^2 + \sigma_u^2 .$$

We assume throughout: **(A1)** the latent model @eq-latent with
selection-on-observables, a constant latent-scale effect, conditional error
law $\varepsilon\mid D,X\sim F$, and
overlap ($J_\theta > 0$, $\operatorname{Var}(D \mid X)$ bounded away from 0
and $\infty$); **(A2)** a correctly specified, regular imputation model, so
$\eta(D, X; \beta_0) = \theta_0 D + g_0(X)$ and
$\sqrt n(\hat\beta - \beta_0) \to_d N(0, \Sigma_\beta)$ with
$\Sigma_\beta = I(\beta_0)^{-1}$ the inverse per-observation information and
influence function $\operatorname{IF}_\beta = I(\beta_0)^{-1} s_\beta(W)$;
**(A3)** cross-fitted adjustment nuisances consistent in $L_2$ with the
product rate
$\lVert \hat\ell - \ell_0 \rVert_2 \lVert \hat m - m_0 \rVert_2 = o_p(n^{-1/2})$,
each $o_p(n^{-1/4})$; **(A4)** finite fourth moments and dominated
differentiability of $\beta \mapsto E[\varphi]$ at $\beta_0$; and **(A5)**
proper draws $\beta^{(b)} \sim N(\hat\beta, \widehat\Sigma_\beta / n)$, so
that $\beta^{(b)} = \hat\beta + n^{-1/2} Z_b$ with
$Z_b \mid \text{data} \sim N(0, \Sigma_\beta)$.

The scaling in (A5) must be explicit because $\Sigma_\beta$ is the
limit of $n \widehat{\operatorname{Cov}}(\hat\beta)$ rather than of
$\widehat{\operatorname{Cov}}(\hat\beta)$ itself; drawing with the
unnormalized matrix would inject variance that does not vanish with $n$.

**Scope.** Assumption (A2) describes the parametric case: one regular
maximum-likelihood fit on the full data, with an estimated covariance from
which to draw. The ordinal designs and wine application use this case.
Implication M1 separately treats a cross-fitted flexible imputation index and
does not require a single $\hat\beta$ or covariance matrix. Its central
condition is an asymptotically linear, residual-weighted projection of the
fitted index error. This learner-level condition does not follow from
cross-fitting or structural partial linearity alone. Theorems S1 and S2
establish it for the deterministic-series reference. For the three adaptive
learners, we verify the generated-outcome linearization but do not claim a
proof of their projected functional condition.

## The pass-through map for general $J$

For an outcome in category $j$ the draw truncates $\xi$ to $(a_j, b_j]$ with
$a_j = \kappa_{j-1} - v$ and $b_j = \kappa_j - v$ at index $v$. Write
$\mu_j(v, \kappa) = E[\xi \mid a_j < \xi \le b_j]$ and
$p_j = F(\kappa^0_j - e) - F(\kappa^0_{j-1} - e)$ for the true category
probability at true index $e$. Averaging over $Y \mid D, X$ gives the
general form of @eq-ess,

$$\psi(v, \kappa; e, \kappa^0) = v + \sum_{j=1}^{J} p_j\, \mu_j(v, \kappa).$$ {#eq-psiord}

**Proposition A1.** *At $(v, \kappa) = (e, \kappa^0)$, writing
$t_k = \kappa^0_k - e$,*

$$c(e) = \frac{\partial \psi}{\partial v}
= 1 - \sum_{k=1}^{J-1} f(t_k)\bigl[\mu_{k+1}(e) - \mu_k(e)\bigr],
\qquad
\frac{\partial \psi}{\partial \kappa_k} = f(t_k)\bigl[\mu_{k+1}(e) - \mu_k(e)\bigr].$$ {#eq-gblocks}

*Consequently $c(e) = 1 - \sum_k \partial \psi / \partial \kappa_k$: the
index derivative and the threshold derivatives are the same kernel.*

*Proof.* Differentiate @eq-psiord. Both truncation limits move with slope
$-1$ in $v$, and the derivative of a truncated mean with respect to its
upper limit is $f(b)(b - \mu)/P$, with the analogous expression at the lower
limit. Evaluated at the truth the truncation probability $P_j$ equals $p_j$,
so the weights in @eq-psiord cancel the denominators and the surviving terms
regroup by threshold. The final identity follows because $\mu_j$ depends on
$(v, \kappa)$ only through $\kappa - v$, so shifting the index is the same
as shifting every threshold the other way. $\square$

Proposition A1 has three practical implications. For binary outcomes,
$J = 2$ and $\kappa_1 = 0$, so the sum reduces to @eq-ce. For ordinal
outcomes, the generally nonzero threshold derivatives require both
$\Sigma_\beta$ and the gradient $G$ below to contain a threshold block.
This requirement is why the ordinal full model must use a routine whose
covariance covers the thresholds (`ordinal::clm`) rather than one whose
covariance does not (`mgcv::ocat`).

The final implication concerns interpretation. Pass-through depends on how
much the categorization reveals about $U$, not simply on the number of
categories. Refining a partition with well-placed thresholds lowers $c$ at
$e = 0$ under a logit, from $0.307$ for binary outcomes to $0.157$ and
$0.090$ for the three- and four-category configurations in
@sec-numverify. Widening the same three-category partition reverses this
pattern, raising $c(0)$ above the binary value and toward 1. The reason is
visible in @eq-gblocks: a category spanning many error standard
deviations has a nearly unbounded truncation interval, so its $\mu_j$ barely
moves with the index and self-correction switches off. We therefore make no
monotonicity claim in $J$. The relevant limit is $c \to 1$ as the observed
category ceases to constrain $U$, the same mechanism that operates at strong
signal.

**The binary logit closed form.** For $J = 2$ under a logistic $F$, the
truncated means have an exact expression that turns @eq-ce into @eq-centropy.
Substituting $u = F(x)$, so that $x = \log\{u/(1-u)\}$ and $\mathrm{d}u = f(x)\,\mathrm{d}x$,
the partial mean is
$$\int_{-\infty}^{t} x f(x)\,\mathrm{d}x
 = \int_0^{F(t)} \log\frac{u}{1-u}\,\mathrm{d}u
 = q\log q + (1-q)\log(1-q) = -H(q), \qquad q = F(t),$$
so the partial mean of a logistic is minus the entropy at the corresponding
probability. Taking $t = -e$ and $p = F(e)$, and using $H(1-p) = H(p)$ with
the fact that the untruncated mean is zero,
$$\mu_-(e) = \frac{-H(p)}{1-p}, \qquad \mu_+(e) = \frac{H(p)}{p},
\qquad \mu_+ - \mu_- = \frac{H(p)}{p(1-p)} .$$
For the logistic law, $f(-e) = f(e) = p(1-p)$, so this factor cancels
exactly in @eq-ce. Thus $c(e) = 1 - H(p)$ and
$c(0) = 1 - \log 2 = 0.307$. The cancellation makes self-correction equal to
the outcome entropy rather than merely proportional to it.

## Cross-fitted partially linear imputation learners

The machine-learning estimator does not have a finite-dimensional $\beta$.
Partition the observations into fixed-$K$ folds $I_k$. For $i\in I_k$, fit
the following index on $I_k^c$:
$$\hat v_{-k}(D,X)=\hat\alpha_{-k}D+\hat f_{-k}(X),$$
and form $B$ completions using independent auxiliary uniforms. Let
$v_0(D,X)=\theta_0D+g_0(X)$ and define the conditional generated-outcome
error
$$\Delta_k(D,X)
=\psi\{\hat v_{-k}(D,X);v_0(D,X)\}-v_0(D,X).$$ {#eq-ml-delta}
For ordinal outcomes, the thresholds are additional arguments of $\psi$.
The scalar-index result below applies directly to the binary
machine-learning implementation.

We use four learner-class conditions. **(M1)** Conditional on its training
sample, each $\hat v_{-k}$ is $L_2$ consistent and the Taylor remainder obeys
$$\sqrt n\sum_k\frac{|I_k|}{n}
E\!\left[|R|\{\hat v_{-k}(D,X)-v_0(D,X)\}^2\mid I_k^c\right]=o_p(1).$$
**(M2)** The residual-weighted projected error has an asymptotically linear
representation for some mean-zero, square-integrable $\rho(W)$,
$$\sqrt n\sum_k\frac{|I_k|}{n}E[R\Delta_k(D,X)\mid I_k^c]
=\frac1{\sqrt n}\sum_{i=1}^n\rho(W_i)+o_p(1).$$
**(M3)** The cross-fitted adjustment nuisances satisfy (A3), including after
replacing the oracle completion by the fitted completion, and their
interactions with $\hat v_{-k}-v_0$ are $o_p(n^{-1/2})$. **(M4)** The
pass-through map is twice continuously differentiable on a neighbourhood
reached with probability tending to one, with a dominated second derivative;
the auxiliary uniforms are independent of the sample and one another; and
the centered completion
$S\{\hat v_{-k}\}-\psi\{\hat v_{-k};v_0\}$ converges in
conditional $L_2(R^2)$ to its oracle counterpart.

Condition (M2) is the substantive requirement on the imputation learner.
Prediction consistency alone does not imply it because the imputation model
lies outside the orthogonality guarantee. For the partially linear class,
the first-order term on its left side is
$$E\!\left[R c(v_0)
\left\{(\hat\alpha-\theta_0)D+\hat f(X)-g_0(X)\right\}\right],$$ {#eq-ml-first}
up to fold averaging. The treatment-coefficient error and the flexible
$X$-direction error both enter at first order. The latter need not vanish:
$R$ is orthogonal to functions of $X$, but $R c(v_0)$ generally is not.

**High-level implication M1 (adaptive-learner expansion).** *Under (A1) and
(M1)-(M4), with fixed $K$ and fixed $B$,*
$$\sqrt n(\hat\theta_B-\theta_0)
=J_\theta^{-1}\frac1{\sqrt n}\sum_{i=1}^n
\left[e_iR_i+\rho(W_i)+R_i\bar u_i\right]+o_p(1),\qquad
\bar u_i=B^{-1}\sum_{b=1}^B u_i^{(b)},$$ {#eq-ml-expansion}
*and hence*
$$\sqrt n(\hat\theta_B-\theta_0)\ \to_d\
N(0,V_B^{\mathrm{ML}}),\qquad
V_B^{\mathrm{ML}}=J_\theta^{-2}
\left\{\operatorname{Var}(eR+\rho(W))+\frac{\sigma_u^2}{B}\right\}.$$ {#eq-ml-vb}

*Argument.* Conditional on the training sample for fold $k$,
Taylor-expand @eq-ml-delta at $v_0$. Proposition A1 gives the linear term in
@eq-ml-first; (M1) and (M4) control the quadratic remainder. Cross-fitting
makes the empirical fluctuation of the vanishing index error $o_p(1)$, while
(M2) supplies the first-order contribution from the training sample.
Neyman orthogonality and (M3) remove the adjustment-nuisance terms.

At the truth, the completion error is $e_i+u_i^{(b)}$. Averaging the $B$
completions leaves $\bar u_i$ with conditional variance $\sigma_u^2/B$. A
joint central limit theorem then gives @eq-ml-expansion and @eq-ml-vb. This
law is unconditional over the observed sample and auxiliary uniforms.
Conditional on the observed sample, the uniforms represent imputation Monte
Carlo error rather than repeated-sample uncertainty.

The unresolved learner-specific conjecture is that the implemented GAM,
MARS, and neural-backfit learners satisfy (M2) and the remaining uniform
remainder requirements in (M1), (M3), and (M4). The argument above also
compresses the held-out empirical fluctuation of $R\Delta_k$, the
denominator replacement, and the joint triangular-array central limit
theorem, none of which is formalized here.

The parametric expression follows as a corollary. If
$\hat v=v(\hat\beta)$ is regular, then
$\rho(W)=G'\operatorname{IF}_\beta(W)$ and
$\operatorname{Var}(eR+\rho)=\sigma_e^2+G'\Sigma_\beta G+2G'C$. If each
completion also redraws $\beta$, the independent draw contributes the
additional $G'\Sigma_\beta G/B$ term in @eq-vb. A flexible learner instead
keeps its fitted index fixed across the $B$ point-estimate completions and
uses the full-pipeline bootstrap to propagate its sampling variability.

### A proved reference learner

For a deterministic logistic-series learner, the conditions behind the
general implication can be verified, and it becomes a theorem. This learner serves as a proof reference rather than the
practical default. Let
$$z_n(D,X)=\{1,D,P_n(X)'\}',\qquad
\hat v_{-k}(D,X)=z_n(D,X)'\hat\beta_{-k},$$
where the dictionary $P_n$ is fixed before observing $Y$ and contains
functions of $X$ only. The implemented reference uses additive normalized
Hermite polynomials and keeps $D$ as a mandatory, unpenalized linear term.
The population logistic projection $\beta_n$ solves
$$E[z_n\{Y-\Lambda(z_n'\beta_n)\}]=0,$$
and we write $v_n=z_n'\beta_n$,
$$
A_n=E[\Lambda(v_n)\{1-\Lambda(v_n)\}z_nz_n'],\qquad
a_n=E[R\,\partial_v\psi(v_n;v_0)z_n],
$$
$$
\rho_n(W)=a_n'A_n^{-1}z_n\{Y-\Lambda(v_n)\},\qquad
h_n(\beta)=E[R\{\psi(z_n'\beta;v_0)-v_0\}].
$$

The theorem uses six series-learner (SL) conditions. **(SL1)** The dictionary is
deterministic, has dimension $q_n$, the eigenvalues of $A_n$ are bounded away
from zero and infinity, and the largest eigenvalue of $E[z_nz_n']$ is
bounded. **(SL2)** Uniformly over unit vectors $t$,
$E|t'z_n|^4$ is bounded, $q_n^3/n\to0$, and
$\max_i\|z_{n,i}\|=o_p\{(n/q_n)^{1/2}\}$.
**(SL3)** The fold-specific logistic maximum-likelihood estimates exist
uniquely with probability tending to one, so asymptotic separation is
excluded. **(SL4)** The approximation is undersmoothed for the target:
$\sqrt n\,h_n(\beta_n)\to0$ and
$$\sqrt n\,E[|R|\{v_n(D,X)-v_0(D,X)\}^2]\to0.$$
**(SL5)** $\rho_n$ obeys Lindeberg's condition,
$\operatorname{Var}\{\rho_n(W)\}\to\tau^2<\infty$, and the adjustment
nuisances satisfy (M3). **(SL6)** With fixed $K$, the fold partition is
deterministic or drawn independently of the observations, and
$$\max_{1\le k\le K}\bigl||I_k|/n - 1/K\bigr|\to0,$$
in probability if the partition is random. These are standard series
restrictions specialized
to the residual-weighted functional. The rate and functional conditions are
in the spirit of @newey1997series; bootstrap theory for semiparametric
generalized additive models provides a related but different result
[@hardle2004bootstrap].

**Theorem S1 (series expansion).** *Under (A1), (M4), and (SL1)-(SL6), with
fixed $K$ and fixed $B$, the deterministic series learner satisfies (M2)
with influence $\rho_n$. Consequently*
$$
\sqrt n(\hat\theta_B-\theta_0)
=J_\theta^{-1}\frac1{\sqrt n}\sum_{i=1}^n
\{e_iR_i+\rho_n(W_i)+R_i\bar u_i\}+o_p(1),
$$ {#eq-series-expansion}
*and the limit is normal whenever the variance of the displayed triangular
array converges.*

*Proof.* Uniformly over the fixed number of folds, the fold training score
and a mean-value expansion give
$$
\hat\beta_{-k}-\beta_n
=A_n^{-1}\frac{1}{|I_k^c|}
\sum_{i\in I_k^c}z_{n,i}\{Y_i-\Lambda(v_{n,i})\}
+r_{n,k},\qquad
\max_k|a_n'r_{n,k}|=o_p(n^{-1/2}).
$$
Expanding the logistic score at $\beta_n$ makes the remainder explicit.
Conditions (SL1)-(SL3) give uniform convergence of the fold Hessians to
$A_n$. The logistic third derivative is bounded, and the envelope and
dimension rates in (SL2) make the quadratic score remainder
$o_p(n^{-1/2})$. This step therefore does not assume the projected
expansion.

A Taylor expansion of $h_n$ at $\beta_n$ has derivative $a_n$ and quadratic
remainder bounded by a constant times
$E[|R|\{z_n'(\hat\beta_{-k}-\beta_n)\}^2]$. Conditions (SL1)-(SL2) make the
fold-averaged remainder $o_p(n^{-1/2})$. Write $p_k=|I_k|/n$. For
$i\in I_j$ the fold-averaged weight on $\rho_n(W_i)$ is $g_j/n$ with
$g_j=\sum_{k\ne j}p_k/(1-p_k)$, and exactly balanced folds give $g_j=1$,
so that
$$
\sum_k\frac{|I_k|}{n}\frac{\mathbf1\{i\in I_k^c\}}{|I_k^c|}=\frac1n.
$$
In general, because (SL6) makes the partition independent of the
observations, $\operatorname{Var}\{|I_j|^{-1/2}\sum_{i\in I_j}\rho_n(W_i)\}$
is $O(1)$ by the variance clause of (SL5), so
$$
\sqrt n\sum_i\Bigl(\frac{g_{j(i)}}{n}-\frac1n\Bigr)\rho_n(W_i)
=\sum_{j=1}^K(g_j-1)\frac1{\sqrt n}\sum_{i\in I_j}\rho_n(W_i)
=O_p\bigl(\max_j|g_j-1|\bigr),
$$
which is $o_p(1)$ under (SL6) since $p\mapsto p/(1-p)$ is continuous at
$1/K$. Asymptotic balance therefore suffices; no rate on the imbalance is
needed. The fold-averaged linear terms equal
$n^{-1}\sum_i\rho_n(W_i)+o_p(n^{-1/2})$. Condition (SL4) removes the series
approximation shift at the target scale. The generated-outcome Taylor
remainder is also negligible by (SL4) and (M4). Finally, (M3) removes the
adjustment-nuisance terms, while the completion split in @eq-split supplies
$e_iR_i+R_i\bar u_i$. Lindeberg's theorem and (SL5) give the stated normal
limit. $\square$

The target-undersmoothing clause in (SL4) is essential. If only its
quadratic-approximation clause holds, the same argument centers the estimator
at the series-projection target
$$\theta_n=\theta_0+J_\theta^{-1}h_n(\beta_n),$$
not at $\theta_0$. The fixed degree-three implementation below measures that
shift and finds it nonzero, so it checks the projected expansion around its
own series target rather than instantiating the $\theta_0$-centered
conclusion. Absent exact representation of the target, satisfying (SL4)
requires a growing dictionary, and (SL2) then binds: its fourth-moment clause is dimension-free, whereas normalized
Hermite polynomials have fourth moments growing exponentially in the
polynomial degree, and local spline or Fourier bases normalized to keep the
Gram eigenvalues bounded have fourth moments growing with $q_n$. Standard
series analysis controls a growing basis envelope instead
[@newey1997series]. Restating (SL2) in envelope form, and rechecking the
Hessian and remainder rates that rest on it, is left open. Theorem S1 should
therefore be read as a conditional result, not one with a demonstrated
growing-dictionary instance.

## Bootstrap propagation for the learner class

Because the fixed-$B$ estimator is randomized, its bootstrap statement
differs slightly from that for a deterministic asymptotically linear
estimator. Let
$\hat\theta_B^*$ rerun the entire procedure on an Efron resample, including
fresh surrogate uniforms, and let
$$\mu_n^*=E^*(\hat\theta_B^*\mid\mathcal D_n)$$
average over resampling, any redrawn fold partition, and the fresh
surrogate randomness. We add four machine-learning bootstrap (MB)
conditions. **(MB1)** The
projected learner expansion in (M2) is bootstrap stable: after centering, its
resampled version has influence
$n^{-1/2}\sum_i(N_i-1)\rho(W_i)+o_{P^*}(1)$ in outer probability.
**(MB2)** The adjustment-nuisance rates in (M3) hold conditionally under
resampling. **(MB3)** The integrated map $\psi$ has the dominated
differentiability in (M4), and the centered randomized completion is
conditionally $L_2(R^2)$ continuous in the fitted index. **(MB4)** Fold
sizes are asymptotically balanced and the learner conditions hold uniformly
over either a fixed partition independent of the observations or a fresh
regular partition in every resample.

The conditions clarify how two existing bootstrap results apply here.
@tang2024 give general consistency conditions for estimators with
machine-learned nuisances, and @lin2026 establish exchangeably weighted
bootstrap validity for DML under its usual conditions. SUDO additionally
requires the generated-outcome conditions (MB1) and (MB3). A raw
inverse-transform draw need not be differentiable for every fixed uniform at
a truncation boundary. Its integrated score $\psi$ must be differentiable,
and conditional $L_2$ continuity controls the centered-draw remainder.

**High-level implication M2 (full-pipeline bootstrap).** *Under (A1),
(M1)-(M4), and (MB1)-(MB4), with fixed $K$
and fixed $B$,*
$$
\sup_t\left|
 P^*\{\sqrt n(\hat\theta_B^*-\mu_n^*)\leq t\}
 -\Phi\{t/\sqrt{V_B^{\mathrm{ML}}}\}
\right|\to_p0,
\qquad
n\,\operatorname{Var}^*(\hat\theta_B^*)\to_pV_B^{\mathrm{ML}}.
$$ {#eq-ml-bootstrap}

*Argument.* Apply the exchangeably weighted expansion to
$eR+\rho(W)$ using (MB1)-(MB2). Conditional on each resampled observation,
the fresh centered completion terms are independent across observations and
draws, with limiting contribution $\sigma_u^2/B$; (MB3) permits replacing
the fitted index in that term by its limit. The conditional Lindeberg
central limit theorem and (MB4) give the same law for fixed or redrawn
regular folds. Conditional second-moment convergence gives the variance
statement.

The unresolved bootstrap conjecture is that the implemented adaptive learners
satisfy (MB1) and the conditional uniformity requirements in (MB2)-(MB4).
The conditional Lindeberg and second-moment steps are likewise stated rather
than formalized.

**Theorem S2 (series full-pipeline bootstrap).** *Suppose the conditions of
Theorem S1 hold uniformly on Efron resamples, the empirical Hessian and score
moments in (SL1)-(SL3) converge conditionally, and (MB2)-(MB4) hold. Also
suppose the empirical second moment of
$e_iR_i+\rho_n(W_i)$ converges to its population counterpart and its maximum
absolute value is $o_p(\sqrt n)$. Then the deterministic series learner
satisfies (MB1). Consequently
@eq-ml-bootstrap holds for this learner, for either fixed or independently
redrawn regular folds.*

*Proof.* On a resample with multiplicities $N_i$, expand each fold training
score at its original-sample solution. Conditional convergence of the
resampled Hessian and the same bounded logistic derivative used in Theorem
S1 give
$$
\sum_k\frac{|I_k|}{n}
\{h_n(\hat\beta_{-k}^*)-h_n(\hat\beta_{-k})\}
=\frac1n\sum_{i=1}^n(N_i-1)\rho_n(W_i)+o_{P^*}(n^{-1/2})
$$
in outer probability. The balanced-fold identity from Theorem S1 applies
conditionally for a fixed partition. For an independently redrawn regular
partition, (MB4) makes the difference between its inclusion weights and
$1/n$ negligible at the same scale. This proves (MB1) rather than assuming
it. The exchangeably weighted central limit theorem applies to
$eR+\rho_n$; (MB2) removes the adjustment-nuisance remainder, and (MB3)
permits replacement of the fitted index in the centered completion term by
its limit. Fresh completion terms contribute $\sigma_u^2/B$ conditionally.
Conditional second-moment convergence yields bootstrap-variance
consistency. $\square$

The centering in @eq-ml-bootstrap affects the inferential conclusion. At fixed $B$,
$\hat\theta_B-\mu_n^*$ contains $O_p(n^{-1/2})$ surrogate Monte Carlo noise.
Consequently the usual claim with
$\hat\theta_B^*-\hat\theta_B$ does not follow. The bootstrap standard
deviation is still consistent and therefore supports the normal interval
$\hat\theta_B\pm z_{1-\alpha/2}\hat s_{\mathrm{boot}}$. A centered
bootstrap distribution may instead estimate $\mu_n^*$ by the mean of the
outer replicates. Raw percentile and conventionally centered studentized
intervals require either $B\to\infty$ fast enough to remove this centering
gap or an additional randomized-estimator argument. We report them as
empirical comparisons, not consequences of Implication M2.

Neither (M2) nor its bootstrap analogue (MB1) has been proved separately for
the implemented GAM, MARS, or neural-backfit algorithms. Implications M1 and
M2 therefore isolate the learner-specific obligations; numerical performance
does not replace them. Redrawing folds is also unnecessary when
(MB4) holds, although it remains the implementation default because it
propagates partition variability. At its Monte Carlo resolution, the paired
check in @sec-numverify finds fixed and redrawn folds indistinguishable.

## Moment properties

**Proposition A2.** *Under (A1)-(A2), $E[\varphi] = 0$ at the truth and the
moment is Neyman-orthogonal in the adjustment nuisances,
$\partial_\ell E[\varphi] = \partial_m E[\varphi] = 0$. It is not orthogonal
in $\beta$: $\partial_\beta E[\varphi] = G$ with blocks*

$$G_{\beta_\eta} = E\bigl[c(\eta_0)\, \partial_{\beta_\eta} \eta\, R\bigr],
\qquad
G_{\kappa_k} = E\bigl[f(\kappa^0_k - \eta_0)\,(\mu_{k+1} - \mu_k)\, R\bigr].$$

*Proof.* At the truth $S - \ell_0 - \theta_0 R = \xi$, and averaging over
$Y \mid D, X$ makes $\xi \mid D, X \sim F$, so $E[\varphi] = E[\xi R] = 0$.
The mean-zero statement holds only after that averaging; given $(W, \beta)$
the draw has mean $e \ne 0$. The nuisance derivatives vanish by the tower
property. The $\beta$ derivative is @eq-gblocks composed with
$\partial_\beta \eta$. $\square$

Because $R$ annihilates any function of $X$, profiling $\ell$ out of the
moment leaves $G$ unchanged. This result covers both implementations:
refitting $\hat\ell$ for every completion, or fitting it once from a plug-in
completion and holding it fixed, as the binary estimator does. The appendix
section on nuisances and draws explains this distinction. Proposition A2
formalizes the claim in the body that the imputation model lies outside the
orthogonality guarantee.

## Asymptotic distribution at fixed $B$

**Conjecture A1.** *Under (A1)-(A5), with
$\bar\theta = B^{-1}\sum_b \hat\theta^{(b)}$,*

$$\sqrt n(\bar\theta - \theta_0) \to_d N(0, V_B), \qquad
V_B = J_\theta^{-2}\Bigl[\sigma_e^2 + G'\Sigma_\beta G + 2G'C
  + \frac{G'\Sigma_\beta G + \sigma_u^2}{B}\Bigr],$$ {#eq-vb}

*where $C = \operatorname{Cov}(eR, \operatorname{IF}_\beta)$.*

*Heuristic derivation.* The $Z$-estimator expansion gives
$\sqrt n(\hat\theta - \theta_0) = J_\theta^{-1}\sqrt n M_n(\theta_0, \cdot) + o_p(1)$,
the sign following from $\partial_\theta E[\varphi] = -J_\theta$. Expanding
around $(\beta_0, \ell_0, m_0)$, the adjustment-nuisance terms are $o_p(1)$
by Proposition A2 plus cross-fitting plus the (A3) product rate, as in
@chernozhukov2018. The $\beta$ term is
$G'\sqrt n(\beta^{(b)} - \beta_0) + o_p(1)$, and
$\beta^{(b)} - \beta_0 = (\hat\beta - \beta_0) + n^{-1/2} Z_b$ by (A5).
Applying @eq-split separates the data-measurable score $e_i R_i$ from the
draw noise; averaging over $b$ leaves parameter-draw and surrogate-draw
terms with conditional variances $\Sigma_\beta / B$ and $\sigma_u^2 / B$.

Two features of @eq-vb are important for interpretation. First, it is not
the single-completion result plus an $O(1/B)$ correction because the leading
term itself depends on $B$. A single completion carries the full sandwich
$\sigma_e^2 + \sigma_u^2$ *and* twice the parameter contribution, once from
$\hat\beta$'s own sampling error and once from the single draw around it,
which is exactly what Rubin's $(1 + 1/B)$ factor exists to undo. As
$B \to \infty$ the draw noise averages away and the leading term is
$\sigma_e^2$, not $\sigma_e^2 + \sigma_u^2$. Second, the cross term $2G'C$
is not sign-restricted because $\operatorname{IF}_\beta$ and the moment are
built from the same data. Consequently, $V_B$ is not the DML sandwich plus a
nonnegative imputation penalty. Pooling removes the draw noise from the
sandwich and adds $G'\Sigma_\beta G + 2G'C$.

## The pooled variance

Rubin's total is $T = \bar W + (1 + 1/B) B_{\!*}$, but its two components
behave differently. The within-draw term $\bar W$ averages a sandwich
estimate over $n$ observations, so
$n\bar W \to_p J_\theta^{-2}(\sigma_e^2 + \sigma_u^2)$. The between-draw
term $B_{\!*}$ is the sample variance of $B$ draws. Writing
$V_{\mathrm{draw}} = G'\Sigma_\beta G + \sigma_u^2$ for the per-draw
variance, @eq-vb gives
$$n B_{\!*} \to_d J_\theta^{-2}\, V_{\mathrm{draw}}\,\frac{\chi^2_{B-1}}{B-1}
\qquad (n \to \infty,\ B \text{ fixed}),$$ {#eq-bstar}
which is *not* a probability limit. With $B$ fixed, the between-draw term
retains sample-variance randomness regardless of the sample size. It is
conditionally unbiased for $J_\theta^{-2}V_{\mathrm{draw}}$, allowing the
comparison below to be stated in expectation.

Write $E_\Delta[\cdot] = E[\cdot \mid W_1, \dots, W_n]$ for the expectation
over the imputation draws with the data held fixed, so that $E_\Delta[T]$ is
still a random variable through the data.

**Conjecture A2.** *Under (A1)-(A5), for fixed $B$,*
$$n\,E_\Delta[T] - V_B \;\to_p\; 2 J_\theta^{-2}\bigl(\sigma_u^2 - G'C\bigr),$$
*with $n\,T$ itself converging in distribution, not in probability, by
@eq-bstar. Taking expectations over the data as well gives the same limit for
the unconditional $n\,E[T] - V_B$, which is the form the Monte Carlo check
below estimates, since each replication redraws both the sample and the
imputations. As $B \to \infty$ the randomness in $B_{\!*}$ averages away and
the statement strengthens to a probability limit for $T$ itself,*
$$n\,T - V_\infty \;\to_p\; 2 J_\theta^{-2}\bigl(\sigma_u^2 - G'C\bigr),
\qquad V_\infty = J_\theta^{-2}\bigl(\sigma_e^2 + G'\Sigma_\beta G + 2G'C\bigr).$$
*Either way the $1/B$ terms cancel, so the discrepancy is free of $B$.
Rubin's variance is asymptotically exact in expectation over the imputation
draws if and only if $G'C = \sigma_u^2$, conservative in that sense when
$G'C < \sigma_u^2$, and anti-conservative when $G'C > \sigma_u^2$.*

At fixed $B$ this is a statement about the mean of $T$, not about $T$
itself. By @eq-bstar, $n\,T$ retains a nondegenerate chi-square component,
so $T$ is not ratio-consistent for $V_B/n$ even when the identity holds.
That finite-$B$ randomness does not invalidate Barnard-Rubin inference,
whose reference distribution exists precisely to absorb uncertainty in the
between-draw variance; @sec-numverify measures how well it does so.

*Heuristic derivation.* Substitute the probability limit of $n\bar W$ and
the conditional mean of @eq-bstar into Rubin's total, then subtract
@eq-vb. The $1/B$ terms cancel. The $B\to\infty$ statement additionally
uses concentration of the between-draw sample variance.

The joint law determines the appropriate reference distribution. Define
$$
V_{\mathrm{data}}=\sigma_e^2+G'\Sigma_\beta G+2G'C,\quad
W_0=J_\theta^{-2}(\sigma_e^2+\sigma_u^2),\quad
Q_0=J_\theta^{-2}(1+1/B)V_{\mathrm{draw}},
$$
and let $\nu=B-1$.

**Conjecture A3 (fixed-$B$ reference law).** *Under (A1)-(A5),*
$$
\left\{\sqrt n(\bar\theta-\theta_0),\,nT\right\}
\ \to_d\
\left\{\sqrt{V_B}\,Z,\,
W_0+Q_0\frac{X}{\nu}\right\},
\qquad Z\sim N(0,1),\quad X\sim\chi^2_\nu,\quad Z\perp X.
$$ {#eq-fixed-b-joint}
*Consequently*
$$
\frac{\bar\theta-\theta_0}{\sqrt T}
\ \to_d\
\frac{\sqrt{V_B}\,Z}
{\sqrt{W_0+Q_0X/\nu}}.
$$ {#eq-fixed-b-reference}

*Heuristic derivation.* The joint central limit theorem posited in
Conjecture A1 writes the
$b$th completed-data limit as
$J_\theta^{-1}(A+Q_b)$, where
$A\sim N(0,V_{\mathrm{data}})$ is common to the draws and the
$Q_b\stackrel{\mathrm{iid}}{\sim}N(0,V_{\mathrm{draw}})$ are independent of
$A$. Their sample mean enters $\bar\theta$ and their sample variance enters
$B_{\!*}$. Normal sample means and sample variances are independent, while
$n\bar W\to_pW_0$, which gives @eq-fixed-b-joint.

The exact limiting critical value is therefore a quantile of a
normal-over-shifted-chi-square mixture, not generally a Student $t$. It is
obtained by solving
$$
E_X\!\left[
2\Phi\!\left\{q_B
\sqrt{\frac{W_0+Q_0X/\nu}{V_B}}\right\}-1
\right]=1-\alpha.
$$
A convenient approximation matches the first two moments of the random
denominator. With
$$
\mu_T=W_0+Q_0,\qquad
\nu_S=\nu(\mu_T/Q_0)^2,
$$
the variance-calibrated Satterthwaite critical value is
$q_S=\sqrt{V_B/\mu_T}\,t_{1-\alpha/2,\nu_S}$.
Barnard-Rubin (BR) instead uses the observed
$r=(Q_0X/\nu)/W_0$ and
$\nu_{\mathrm{BR}}=\nu(1+1/r)^2$ in the limit. It is not exactly
@eq-fixed-b-reference, but the comparison in @sec-numverify shows that it is
close.

The discrepancy need not have a fixed sign, and the calculation does not
yield exactness. On a deliberately congenial testbed that isolates pooling
from every other source of error (a correctly specified logistic imputation
model, continuous treatment so both adjustment nuisances are exactly
linear), all six quantities are available in closed form:

| $\theta_0$ | $\sigma_e^2$ | $\sigma_u^2$ | $G'\Sigma_\beta G$ | $G'C$ | $\sigma_u^2 - G'C$ | $T/V$ at $B = 20$ |
|---|---|---|---|---|---|---|
| 1 | 1.269 | 2.021 | 4.077 | 1.954 | $+0.067$ | 1.014 |
| 3 | 0.580 | 2.710 | 41.414 | 3.094 | $-0.384$ | 0.985 |

The sign of the discrepancy flips with signal strength: on average $T$
overstates the sampling variance by about 1.4% at $\theta_0 = 1$ and
understates it by about 1.5% at $\theta_0 = 3$. The $T/V$ column compares
$n\,E[T]$ with $V_B$, which is the quantity Conjecture A2 governs at fixed $B$;
individual replications scatter around it according to @eq-bstar. Both
discrepancies are asymptotic rather than finite-sample effects; they persist
at $n = 20{,}000$ and, being free of $B$, do not shrink as $B$ grows. The
mechanism is visible in the table. As the signal grows the
observed $Y$ carries less information beyond the index, so $\sigma_e^2$
falls while $\sigma_u^2$ and $G'C$ rise, and $G'C$ overtakes $\sigma_u^2$.
This is the variance-side account of why inference is hardest at strong
effect sizes, and it is separate from the bias-side pass-through argument in
the body. The two must not be conflated. Rising $c$ explains why a *given*
index misspecification survives more nearly intact; it cannot itself explain
degradation under a correctly specified imputation model, because $c \to 1$
is precisely the regime in which $\psi$ is closest to the identity and its
linearization is most accurate. Under correct specification the degradation
is in the variance components above and in imputation-model estimation
error.

The practical effect is small: a 1.5% variance error corresponds to roughly
0.7% in a standard error, well within the Monte Carlo error of the coverage
figures in @sec-sims. The theoretical conclusion is nevertheless important:
proper draws plus Rubin's rules are not exactly variance-calibrated here.
@sec-numverify reports the term-by-term Monte Carlo check of
@eq-vb.

**Implication for improper draws.** If $\beta^{(b)} \equiv \hat\beta$,
the between-draw term loses the parameter contribution entirely, so $B_{\!*}$
is conditionally unbiased for $J_\theta^{-2}\sigma_u^2$ alone and $n\,E[T]$
falls short of $V_B$.
The deficit is dominated by the imputation channel $G'\Sigma_\beta G$
whenever $G \ne 0$ and the signal is not weak, though it is not exactly that
quantity, since the cross term and the draw noise also enter. This is the
0.92 versus 0.97 coverage gap in @tbl-ladder.

## Bias under a misspecified imputation model

**Proposition A3.** *Consider a local sequence of binary-treatment
imputation models whose index error equals $\delta_D$ in the treated arm and
$0$ in the control arm, with $\delta_D\to0$. The corresponding population
SUDO target $\theta(\delta_D)$ satisfies*
$$\theta(\delta_D) - \theta_0 = \bar c_1 \delta_D
  + o(\delta_D), \qquad
\bar c_1 = \frac{E[w\, c(\theta_0 + g_0)]}{E[w]}, \quad w = \operatorname{Var}(D \mid X).$$

*Proof.* Differentiate the generated-outcome mean
$E[S \mid D, X]$ with respect to the local index perturbation at zero.
Proposition A1 gives derivative $c(\eta_0)$. Applying
$J_\theta^{-1}E[\,\cdot\,R]$ and conditioning on $X$ with
binary $D$, the moment weights the treated arm by $w$ and the control arm by
$-w$; the control-arm error is zero by assumption, so only
$c(\theta_0 + g_0)$ survives, and $J_\theta = E[w]$ supplies the
denominator. $\square$

Both restrictions matter. Write $\eta_d = \theta_0 d + g_0(X)$ and consider a
local path whose treated- and control-arm index errors are
$\varepsilon h_1(X)$ and $\varepsilon h_0(X)$. Conditioning
$E[R\,\Delta_D]$ on $X$ gives $w(\Delta_1 - \Delta_0)$, so
$$\theta(\varepsilon) - \theta_0
= \varepsilon\,
\frac{E\bigl[w\{c(\eta_1)h_1 - c(\eta_0)h_0\}\bigr]}{E[w]}
+ o(\varepsilon).$$
Proposition A3 is the case $h_1 = 1$, $h_0 = 0$, which is why the displayed
$\bar c_1$ evaluates $c$ on the treated arm only. Taking $h_1 = a + h$ and
$h_0 = h$ adds the first-order covariate-direction leak of @eq-leak to that
treated-arm term, and the leak is nonzero because $c$ depends on $D$ through
$\eta$ and so is not annihilated by projecting on $R$. For continuous $D$
with index error $\delta_D D$ the multiplier is instead
$\bar c_1 = E[c\{\theta_0 D + g_0(X)\} D R] / E[R^2]$, which is a different
expression.
@tbl-sensitivity is the empirical counterpart for the treatment direction,
recovering the predicted slope $\bar c_1$.

For a finite perturbation, define the arm-specific path derivative and its
average
$$
c_d(t;h)=
\left.\partial_v\psi(v;\eta_d)\right|_{v=\eta_d+t h},
\qquad
\bar c_d(h)=\int_0^1c_d(t;h)\,dt ,
$$
where $\psi(v;\eta_d)$ uses fitted index $v$ while the observed category is
still generated at the true index $\eta_d$. At $h=0$,
$c_d(0;0)=c(\eta_d)$ from Proposition A1. Write
$\|a\|_w=\{E[wa^2]\}^{1/2}$ and $J_\theta=E[w]$.

**Proposition A4 (covariate-leakage bound).** *Suppose $D$ is binary and the
limiting imputation-index error is the same $h(X)$ in both arms. Then the
population SUDO target has the exact covariate-direction bias*
$$
L_X(h)=J_\theta^{-1}
E\!\left[w\,h\{\bar c_1(h)-\bar c_0(h)\}\right],
$$ {#eq-leak-exact}
*and hence*
$$
|L_X(h)|\leq
J_\theta^{-1}\|h\|_w\,
\|\bar c_1(h)-\bar c_0(h)\|_w .
$$
*Writing the first-order term as*
$$L_X^{(1)}(h)=J_\theta^{-1}
E[w\,h\{c(\eta_1)-c(\eta_0)\}],$$
*the sharper local bound is*
$$
|L_X(h)-L_X^{(1)}(h)|
\leq J_\theta^{-1}\|h\|_w\,
\left\|
\{\bar c_1(h)-c(\eta_1)\}
-\{\bar c_0(h)-c(\eta_0)\}
\right\|_w .
$$

*Proof.* The fundamental theorem of calculus gives
$\psi(\eta_d+h;\eta_d)-\eta_d=h\bar c_d(h)$. Conditioning
$E[R\Delta_D]$ on $X$ gives
$w(\Delta_1-\Delta_0)$, proving @eq-leak-exact. The two bounds are
Cauchy-Schwarz, first on the full path-factor difference and then only on
its deviation from the derivative at the truth. $\square$

The bound is computable for a specified sensitivity perturbation $h$ and
becomes a conservative sensitivity radius when $\|h\|_w$ is bounded. Under
$\epsilon\leq m(X)\leq1-\epsilon$,
$J_\theta\geq\epsilon(1-\epsilon)$ makes the overlap dependence explicit.
The local remainder is second order when the path derivative is locally
Lipschitz. Nothing forces the first-order term or its bound to be small
relative to the treatment-direction contribution.

# Appendix: simulation and application details {#sec-appendix}

Each design draws a latent utility
$U = \theta_0 D + g_0(X) + \varepsilon$ and converts it to an observed
category. Most designs use logistic $\varepsilon$; the exceptions are
specified below. The designs vary the covariate structure, cutpoints, and
fitted models; covariates remain standard normal throughout. The `R/` and
`R/results/` directories contain the full data-generating code and committed
per-cell results.

## Data-generating processes

All designs instantiate @eq-dgp with one of two covariate structures:

- **S1**, one covariate: $p = 1$, $\pi(X) = 0.8X$, $g_0(X) = X$. Every model
  in play is correctly specified, so only the pooling is under test.
- **S5**, five covariates: $p = 5$, $\pi(X) = X_4 + \cos X_5$,
  $g_0(X) = X_1^2 + \sin X_2 + 0.5 X_3$. The outcome structure is nonlinear
  in $X$, and treatment depends on covariates absent from $g_0$.

| Design | Structure | $F$ | $\kappa$ | $\theta_0$ | $n$ | Replications |
|---|---|---|---|---|---|---|
| Inference ladder (@tbl-ladder) | S1 | logistic | $(0)$ | 1.5 | 5000 | 500 |
| Ordinal parametric (@tbl-ordinal, top) | S1 | logistic | $(-1, 1)$ | 1.0 | 3000 | 500 |
| Binary ML | S5 | logistic | $(0)$ | 0.5, 3 | 1000, 2000 | 200 |
| Ordinal ML (@tbl-ordinal, bottom) | S5 | logistic | $(0, 2)$ | 1.0 | 1000, 2000 | 200 |
| Link misspecification | S1 | Gumbel | $(0)$ or $(0, 2)$ | 1.0 | 5000 | 200 |
| Sensitivity (@tbl-sensitivity) | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 200 |
| Black-box imputation | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 100 |
| PL learner comparison (@tbl-pl-learners) | S5 | logistic | $(0)$ | 1.5, 3 | 2000 | 50 |
| Bayesian imputation (repository only) | S5$'$ | normal | $(0)$ | 1.5, 3 | 2000 | 200 |
| DGP variation, continuous $D$ | S5, $D = 0.7\pi(X) + N(0,1)$ | logistic | $(0)$ or $(0, 2)$ | 1, 2.5 | 2000 | 200 |
| DGP variation, interaction | S5 plus $X_1 X_2$ in $g_0$ | logistic | $(0)$ | 1, 2.5 | 2000 | 200 |
| DGP variation, dimension | S5 with $p = 10$ | logistic | $(0)$ | 1, 2.5 | 2000 | 200 |
| DGP variation, overlap | S5, treatment-assignment index $\pi(X)$ scaled $2.5\times$ | logistic | $(0)$ | 1, 2.5 | 2000 | 200 |

: Data-generating processes. $\kappa$ lists the interior cutpoints, so
$(0)$ is binary and a pair gives $J = 3$. {#tbl-dgp}

S5$'$ is the probit analogue of S5 used for the Bayesian imputation model,
whose binary machinery is probit: the systematic part is divided by
$\mathrm{sd}(\text{logistic}) = \pi/\sqrt3$ before normal errors are added,
which matches the event rates of S5 to within 0.006 and puts $\theta_0$ on
the probit scale at $\theta_0/1.8138$. The $\theta_0$ column reports the
logit-scale values for comparability. That design supports no table here;
it is reported so the claim in the machine-learning section is checkable.

**Models fitted.** The ladder uses $\mathrm{glm}(Y \sim D + X)$ as the
imputation model with $B \in \{10, 25, 50\}$ draws (the table shows
$B = 25$). The ordinal parametric design uses `ordinal::clm`, whose
covariance covers the thresholds, with $B = 25$. The ordinal ML design uses
`clm` on $D$ and natural-spline bases $\mathrm{ns}(X_k, 4)$ with `mgcv::gam`
partialling nuisances and $B = 25$.

**Link misspecification.** This design retains the one-covariate structure
but replaces logistic $\varepsilon$ with an extreme-value error: Gumbel-max
for the binary case, matching `glm` with a cloglog link, and Gumbel-min for
the ordinal case, matching `ordinal::clm` with a cloglog link. The analyst
fits either logit (misspecified) or cloglog (correct). Holding the simple
covariate structure fixed isolates error in the link.

**Sensitivity ladder (@tbl-sensitivity).** The ladder starts from the oracle
index and oracle partialling nuisances, then adds a known perturbation of
size $\delta$ to one input: the outcome nuisance only (DML-1), both
nuisances (DML-2), the index in the covariate direction (SUDO-X), or the
index in the treatment direction (SUDO-D). Bias is traced against a common
latent-scale root-mean-square error (RMSE) so the four are comparable. The
fitted SUDO-D slope reproduces $\bar c_1$ (0.669 at
$\theta_0 = 1.5$, 0.861 at $\theta_0 = 3$) to within 0.001, which is the
independent check on @eq-ce.

**Partially-linear imputation models.** The original black-box result uses
$\hat V = \hat\alpha D + \hat f(X)$ with $\hat f$ a single-hidden-layer
neural net (`nnet`, `size = 4`, `decay = 0.3`, five backfitting iterations,
selected on target-level Monte Carlo bias) and 100 full-pipeline
bootstrap resamples.
The common adaptive-learner comparison uses 50 resamples for each arm. Its
GAM has an unpenalized linear treatment term and smooths of the five
covariates. The MARS fit builds degree-two bases from $X$ only (`nk = 21`,
penalty 3), then jointly estimates those basis coefficients and a mandatory
linear treatment coefficient by logistic regression. This construction is needed because
`earth::linpreds` alone still permits the named predictor in interaction
terms. The theorem-reference series is evaluated separately. It uses
additive normalized Hermite polynomials through degree three, a dictionary
fixed before observing $Y$, and a separate mandatory linear treatment term.

**Wine application.** Full model `ordinal::clm` on $D$ and either the raw
covariates (linear) or $\mathrm{ns}(X_k, 3)$ (spline); partialling nuisances
$E[S \mid X]$ and $E[D \mid X]$ by `mgcv::gam` (Gaussian for the continuous
$D$); five-fold cross-fitting; $B = 50$ proper draws; Rubin pooling. Volatile
acidity is standardized.

## Nuisances, cross-fitting, and draws

Partialling nuisances $\hat\ell(X) = E[S \mid X]$ and $\hat m(X) = E[D \mid X]$
are `mgcv::gam` throughout, fit with five-fold cross-fitting. Proper draws
sample the full-model parameters from $N(\hat\beta, \widehat{\mathrm{Cov}}(\hat\beta))$
(thresholds included for `clm`) before each completion. Diagnostic draws use
a random number generator (RNG) stream independent of the estimation draws.

One algorithmic detail differs by outcome type. Because the surrogate $S$
changes with every draw, its nuisance $\hat\ell$ can also be refit each time.
Binary experiments found this refitting negligible, so the binary estimator
fits $\hat\ell$ once on a plug-in completion and reuses it. The ordinal
nuisance-isolation ladder gives a different result: holding $\hat\ell$ fixed
nearly doubles the within-draw component and accounts for the stage-5
conservatism. The corrected ordinal estimator therefore cross-fits
$\hat\ell$ separately for every completion. The full validation with
$B=25$ and 200 replications attains 0.960 coverage. Refitting costs roughly
$B$ times as much for the outcome nuisance, so the less expensive binary
behavior is retained where it has been validated.

## Monte Carlo precision

Coverage is a binomial proportion, so its Monte Carlo standard error at a
true 0.95 is about $\sqrt{0.95 \cdot 0.05 / R}$: 0.010 at $R = 500$
replications, 0.015 at $R = 200$, and 0.022 at $R = 100$. Per-cell bias and
coverage Monte Carlo standard errors are in the committed result files.

# Supplement: numerical verification of the theory {#sec-numverify}

Every closed-form expression in @sec-theory is checked through an independent
numerical route. The corresponding scripts are in the project repository and
assert their own pass conditions.

## Pass-through derivatives (Proposition A1)

`R/theory_ordinal_passthrough.R` computes $\psi$ by quadrature and compares
the analytic index and threshold derivatives of @eq-gblocks against central
differences, for $J = 2, 3, 4$ under logit, probit, and both Gumbel
conventions. Maximum discrepancy across all configurations is $3\times10^{-10}$
for the index block and $4\times10^{-10}$ for the threshold block; the
identity $c = 1 - \sum_k \partial_{\kappa_k}\psi$ holds to $10^{-15}$. A
Monte Carlo check confirms `complete_surrogate()` samples the map it should
($|z| < 4$ at $4\times10^5$ draws). The analytic map therefore matches the
implemented estimator.

At $e = 0$ under a logit, the threshold-placement effect discussed in
@sec-theory is:

| $\kappa$ | $(0)$ binary | $(-0.5, 0.8)$ | $(-1, 0.2, 1.3)$ | $(\mp 0.5)$ | $(\mp 1.5)$ | $(\mp 3)$ | $(\mp 6)$ |
|---|---|---|---|---|---|---|---|
| $c(0)$ | 0.307 | 0.157 | 0.090 | 0.175 | 0.223 | 0.636 | 0.966 |

The final four entries all use $J = 3$. Widening the three-category
partition moves $c$ from well below the binary value to nearly 1.

## Deterministic-series expansion (Theorems S1 and S2)

`R/theory_pl_series.R` checks the proof-specific quantities without using
the fitted SUDO estimate as its own benchmark. It computes the population
logistic-series projection and its influence function by product
Gauss-Hermite quadrature, then compares them with independently fitted
five-fold estimates. The implemented dictionary uses normalized additive
Hermite polynomials through degree three and a separate mandatory linear
treatment term.

The population projection shifts the SUDO target by $-0.00082$ at
$\theta_0=1.5$ and $-0.00167$ at $\theta_0=3$. These small values are
measured approximation errors for this design, not a substitute for (SL4).
At $n=4000$, the empirical standard deviation divided by the
influence-function prediction is 0.989 and 0.985 at the two signals. The
root-mean-square expansion remainder divided by the predicted standard
deviation falls from 0.439 to 0.198 at the moderate signal and from 0.623 to
0.289 at the strong signal as $n$ increases from 1000 to 4000. The
strong-signal remainder is visibly slower to disappear, consistent with the
near-separation concern isolated in (SL3).

For 30 datasets at $n=2000$ and 50 Efron resamples per dataset, the bootstrap
standard-deviation ratio to the weighted influence expansion is
1.015 with fixed folds and 1.014 with redrawn folds at $\theta_0=1.5$; the
corresponding ratios are 1.076 and 1.077 at $\theta_0=3$. All 3000 paired
bootstrap resamples are complete. These checks validate the algebra and
finite-sample behavior of the theorem reference. Theorems S1 and S2 provide
the proof; the simulation does not.

## Machine-learning expansion (Implication M1)

`R/theory_ml_expansion.R` compares the exact conditional generated-outcome
shift $\psi(\hat v;v_0)-v_0$ with its linear approximation
$c(v_0)(\hat v-v_0)$ for GAM, MARS, and neural backfitting. It uses the S5
design, $\theta_0\in\{1.5,3\}$, $n\in\{1000,2000\}$, and 50 replications per
cell. The analytic derivative agrees with central differences to
$7.2\times10^{-11}$. Across all 12 learner cells, the mean absolute
pointwise Taylor remainder divided by mean squared index error ranges from
0.016 to 0.027. At $n=2000$, the absolute residual-projected remainder ranges
from 0.0015 to 0.0107, against a first-order scale of 0.063 to 0.138. The
generated-outcome part of (M1) is therefore second order in the fitted-index
error in every implemented design. This check does not establish the
learner-specific asymptotic representation in (M2).

## Machine-learning bootstrap (Implication M2)

`R/theory_ml_bootstrap.R` compares fixed and redrawn folds on the same 30
generated datasets for each learner and signal, with 15 outer resamples and
15 completions per resample. The fixed/redrawn mean-SE ratio ranges from
0.97 to 1.08, and every paired bias difference is smaller than its
prespecified Monte Carlo tolerance. At this resolution, redrawing folds is
not empirically necessary in this design.

The uncertainty hierarchy depends on signal strength but is consistent
across learners. At $\theta_0=1.5$, the within-fold SE is 0.90 to 0.93 of the
full-pipeline SE; at $\theta_0=3$ it is only 0.75 to 0.77. Holding the fitted
index fixed reduces the SE further, to 0.75 to 0.92 of the within-fold
value. The within-fold method therefore does not reproduce the improper
method. It captures some learner uncertainty but misses the data, nuisance,
and fold layers propagated by the full pipeline. Normal coverage is 0.93 to
0.97 across the six cells. Percentile coverage is 0.87 to 0.97 and
conventionally centered studentized coverage is 0.80 to 0.93, which supports
the fixed-$B$ centering qualification rather than either alternative as a
default. With 30 replications, coverage Monte Carlo standard errors are
about 0.03 to 0.07. These checks support the implications of (MB4) and the
variance ordering, but do not establish learner-specific condition (MB1).

## The variance decomposition (Conjectures A1 and A2)

`R/theory_variance_terms.R` evaluates $J_\theta$, $\sigma_e^2$, $\sigma_u^2$,
$G$, $\Sigma_\beta$ and $C$ in closed form on the congenial testbed by
product-grid quadrature, then compares @eq-vb against 4000 replications at
$n = 20{,}000$ for $B \in \{1, 2, 5, 20\}$ and $\theta_0 \in \{1, 3\}$. All
eight cells agree within 1.2 Monte Carlo standard errors. Comparing across
$B$ is what identifies the individual terms; a single pooled variance ratio
cannot, because its Monte Carlo error is the same size as the 1 to 2 percent
effect being measured, which is why the earlier ratio-based reading of these
simulations was inconclusive rather than confirmatory.

The fixed-$B$ spread of @eq-bstar is confirmed directly: at $\theta_0 = 1$,
$B = 20$, $n = 20{,}000$, the observed standard deviation of $nT$ across
replications is 2.13 against a mean of 9.73, while @eq-bstar predicts
$(1 + 1/B)V_{\mathrm{draw}}\sqrt{2/(B-1)} = 2.08$. The residual spread in the
reported variance is therefore attributable to the between-draw sample
variance rather than estimation error, and it does not shrink with $n$.

`R/theory_fixed_b_inference.R` evaluates @eq-fixed-b-reference at
$B\in\{5,10,25,50,100\}$ and then checks it with 1,000 independent samples
at $n=5{,}000$. Prediction and empirical coverage differ by at most 0.012
across all 40 comparisons in @tbl-fixed-b. Normal inference is inadequate
at small $B$ and strong signal; Barnard-Rubin is close to the exact mixture,
and the predicted calibrated Satterthwaite coverage is within 0.006 of 0.95
in every cell.

```{r}
#| label: tbl-fixed-b
#| echo: false
#| tbl-cap: "Fixed-imputation 95% coverage, theory prediction / finite-sample estimate; $n=5000$, 1000 replications."
fb <- read_result("theory_fixed_b_inference.csv")
pair_cov <- function(pred, emp) sprintf("%.3f / %.3f", pred, emp)
table_data <- data.frame(
  `$\\theta_0$` = nn(fb$theta0, 0),
  `$B$` = fb$B,
  Normal = pair_cov(fb$coverage_normal_pred, fb$coverage_normal_emp),
  `Barnard-Rubin` = pair_cov(fb$coverage_br_pred, fb$coverage_br_emp),
  Satterthwaite = pair_cov(
    fb$coverage_satterthwaite_pred,
    fb$coverage_satterthwaite_emp
  ),
  Mixture = pair_cov(fb$coverage_mixture_pred, fb$coverage_mixture_emp),
  check.names = FALSE
)
knitr::kable(table_data, align = c("r", "r", "r", "r", "r", "r"))
```

The practical default remains Barnard-Rubin with at least 25 imputations. It
requires only the usual pooled quantities and is close to the exact mixture,
whereas the calibrated alternatives require consistent estimates of
$V_B$, $W_0$, and $Q_0$. Increasing $B$ reduces the random chi-square
component but does not remove Conjecture A2's systematic Rubin discrepancy.
For a flexible imputation learner, the full-pipeline bootstrap normal
interval remains the corresponding default; the percentile and conventional
studentized comparisons in @tbl-ml-bootstrap do not improve it.

## Covariate-direction leakage (Proposition A4)

`R/theory_covariate_leakage.R` evaluates the exact path identity and both
bounds with two million population draws from the stage-7 interaction
design. It omits $\lambda X_1X_2$ for
$\lambda\in\{\mp0.5,\mp1,\mp1.5\}$ at $\theta_0\in\{1,2.5\}$. The global
and local bounds cover the exact covariate leak in all 12 cells. The local
bound is 2.82 to 7.79 times the absolute leak, while the absolute Taylor
remainder divided by $\lambda^2$ is 0.0027 to 0.0091. It is therefore
conservative but has the correct second-order scale in this symmetric
interaction design.

The calculation also corrects the earlier interpretation of stage 7.
Omitting the interaction shifts the pseudo-true logistic treatment
coefficient even though $D$ is constrained to enter linearly. At
$\lambda=1$, the population decomposition is:

| $\theta_0$ | Covariate leak | Induced treatment component | Total prediction | Stage-7 bias |
|---|---:|---:|---:|---:|
| 1.0 | -0.015 | -0.021 | -0.036 | -0.050 |
| 2.5 | -0.013 | -0.068 | -0.081 | -0.147 |

The population calculation is deliberately simpler than the fitted
additive GAM and understates the observed magnitude, especially at strong
signal. Its differences from stage 7, 0.014 and 0.066, are inside the
prespecified Monte Carlo tolerances of 0.030 and 0.075. More importantly,
the decomposition establishes that the interaction experiment is not a pure
covariate-leakage arm: the induced treatment-direction channel is at least
as large.

Both closed forms behind these quantities rest on one identity,
$p(1-p)(\mu_+ - \mu_-) = H(p)$ for the logistic law, which yields both
$c(e) = 1 - H(F(e))$ and $C = \Sigma_\beta(A - G)$ with $A = E[zR]$. It is
verified pointwise by quadrature to $8\times10^{-16}$.

The ordinal counterpart is checked by
`R/theory_ordinal_variance_terms.R` on a correctly specified
three-category cumulative-logit design with continuous treatment and exactly
linear partialling nuisances. The calculation includes both threshold
derivatives in $G$ and matches the fixed-$B$ sampling variance and expected
Rubin variance within 2.3 Monte Carlo standard errors in every cell at
$n=8{,}000$, 500 replications, $B\in\{1,2,5,20\}$, and
$\theta_0\in\{1,2.5\}$. At $B=20$, the predicted $T/V$ is 1.028 and 0.980,
respectively. The ordinal threshold machinery therefore changes the same
small Rubin discrepancy identified in the binary testbed but does not explain
the much larger stage-5 gap.

`R/theory_ordinal_nuisance_ladder.R` then changes one partialling component at
a time on the exact stage-5 DGP, using common samples, folds, imputation fits,
parameter draws, and surrogates across arms. At $n=2{,}000$, $B=10$, and 100
replications, the current fixed-$\hat\ell$ arm has empirical variance 0.01380
and mean $T=0.02077$, while refitting $\hat\ell$ on every draw has empirical
variance 0.01375 and mean $T=0.01295$. Thus the point estimator is unchanged
but $T/V$ falls from 1.505 to 0.942. The difference is almost entirely in
$\bar W$, which falls from 0.01666 to 0.00876; the between-draw component is
unchanged. This isolates reuse of the plug-in outcome nuisance as the source
of the stage-5 variance inflation. `R/stage5r_ordinal_refit.R` performs the
full correction check at $n=2{,}000$, $B=25$, and 200 replications. Bias is
0.016 (Monte Carlo SE 0.007), coverage is 0.960 (Monte Carlo SE 0.014), and
$T/V=1.214$ (Monte Carlo SE 0.122), satisfying the stage's bias, coverage,
and variance-calibration criteria. The corrected per-draw refit is therefore
the ordinal default.

# References

::: {#refs}
:::
